### 2020-04-08 paper.tex
initial commit

\documentclass{article}
\usepackage[letterpaper, margin=1in]{geometry}

\title{The yaq Project: a Protocol for Scientific Instrumentation}
\author{Kyle F. Sunden,
  John C. Wright,
  Blaise J. Thompson}

\begin{document}

\maketitle

\section{Introduction}

Many existing projects.
Lack of reusability and portability.

\section{Protocol}

TCP

JSON-RPC

configuration

recommend implementing as daemon (service)

\section{Daemons}

Daemon roundup.

Discussion of traits and purpose of compositional approach.

\section{Clients}

A few ``short stories'' of particular instruments utilizing yaq.

\section{Conclusions}

\end{document}

### 2022-04-01 paper.tex

@@
 
+\section{Abstract}
+
 \section{Introduction}
 
-Many existing projects.
-Lack of reusability and portability.
+Instrumentation development is a key part of the scientific enterprise.
+Scientists have always relied on their creativity and wit to assemble new instruments capable of carrying out new experimental procedures.
+As the scientific industry has advanced, scientists are increasingly able to directly purchase pieces of their instruments.
+A modern instrument might contain a motor from company A, a sensor from company B, a specialty light source from company C, and a totally custom robotic sample stage.
+The brand new scientific instrument, doing a brand new experiment, is cobbled together from a few dozen pieces of hardware, some brand new and some reused.
+The instrumental novelty arises from the creative way that all of those pieces are assembled and controlled in the context of the experiment.
+
+Software integration can be a frustrating piece of the modern instrument development process.
+Each individual component has its own drivers and interface.
+A lack of standards means that each scientist must develop her own instrument control application.
+Weeks can be spent just integrating one new component into an existing project.
+Code reusability is typically poor, and technical debt grows quickly in academic and educational contexts.
+Sientists may struggle to rapidly innovate on their experimental design when each hardware addition requires major software development.
+
+Briefly, we
+Minimum viable capabilities for desired protocol.
+Focus on making client development easy.
+\begin{itemize}
+  \item Networked machines (subitem--multilingual)
+  \item Unified interface (traits)
+  \item Self describing
+  \item Built on exisiting protocols
+  \item Reusable
+  \item Synchronous clients
+  \item Portable, easy to install
+  \item Interface agnostic. Not just serial. Certainly not just SCPI.
+\end{itemize}
+Well defined, concrete interface.
+Static interface.
+
+There are several large exisiting open source hardware interface projects [cite EPICS, TANGOS, pymodaq, other].
+These projects are impressive at facility scale, but in our experience these do not scale well to single-investigator lab environments.
+We have aimed to keep yaq as simple as possible, with the goal of staying easy to manage at very small scales without dedicated staff.
+Certainly yaq lacks features compared to these exisiting projects, as we will highlight.
+
+There are exisiting open source hardware interface projects with a focus on simplicity [cite INSTRBUILDER etc].
+To our knowledge none of these meet the minimal capabilities as outlined above.
+
+Overview of this publication.
+yaq is a protocol.
+yaq is an implementation that runs as a stand-alone daemon.
+yaq has a variety of clients.
+yaq is not opinonated about orchestration layer.
 
@@
 
-JSON-RPC
+AVRO-RPC
 
 configuration
-
-recommend implementing as daemon (service)
 
@@
 
+Fakeable.
+
+\subsubsection{yaqc}
+
+Lightweight scriptable.
+
+\subsubsection{yaqc-qtpy}
+
+Capitalizes heavily on traits
+
+\subsubsection{Stahl}
+
+Self contained ecosystem on a raspberry pi.
+
+Dedicated client application.
+
+Exceedingly simple.
+
+(reference publication)
+
+\subsubsection{Landis}
+
+....
+
+\subsubsection{Goldsmith}
+
+Mixing with other hand built or commercially derived software.
+
+\subsubsection{Bluesky}
+
+Bridging to larger open source control layers.
+
 \section{Conclusions}

### 2022-04-04 paper.tex

@@
 
-There are several large exisiting open source hardware interface projects [cite EPICS, TANGOS, pymodaq, other].
+There are several large exisiting open source hardware interface projects, such as EPICS \cite{DalesioLR1991a}, TANGO \cite{tango-controls.org}, and pymodaq \cite{WeberSebastien2021a}.
 These projects are impressive at facility scale, but in our experience these do not scale well to single-investigator lab environments.
@@
 
+Bluesky is an extensive open source control layer \cite{AllanDanielB2019a}.
 Bridging to larger open source control layers.
@@
 
+\clearpage
+\bibliographystyle{unsrt}
+\bibliography{references}
+
 \end{document}

### 2022-04-29 paper.tex
skeleton

@@
 
+\subsection{TCP}
+
+\subsection{Avro RPC}
+
 AVRO-RPC
+
+\subsection{Configuration and State}
 
@@
 
-\section{Clients}
+\subsection{Traits}
 
-A few ``short stories'' of particular instruments utilizing yaq.
+Graphic of traits?
+
+Examples of daemons for core traits.
 
@@
 
-\subsubsection{yaqc}
+\section{Clients}
 
-Lightweight scriptable.
+yaq's RPC design invites the creation of multiple client libraries specialized for specific purposes.
+Again, daemons can communicate simultaniously with multiple clients.
+For a yaq instrument it is normal to have multiple clients running simultaniously, each focused on a particular purpose.
+This modularity speeds development and increases overall flexibility of the system.
+We have created a few client libraries which we will discuss here.
 
-\subsubsection{yaqc-qtpy}
+yaqc [CITE] is a very basic Python client for yaq.
+yaqc is totally generic and aims to support any conceivable yaq RPC.
+This library is easy to use directly in Python scripts or within the REPL, but it's also very useful as a dependency when building more specialized Python clients.
 
-Capitalizes heavily on traits
+yaqc-qtpy [CITE] is a specialty Qt-based [CITE] GUI client.
+yaqc-qtpy provides a special QClient object which uses Qt threading primatives to provide performant and safe access to yaq from a Qt environment.
+In addition, yaqc-qtpy provides an entry point which automatically builds a GUI from arbitrary protocols using traits.
+This is an invaluable tool which provides a ``free'' graphical user interface to any daemon.
 
-\subsubsection{Stahl}
+Bluesky is an extensive open source control layer developed collaboratively across several user facilities \cite{AllanDanielB2019a}.
 
-Self contained ecosystem on a raspberry pi.
+Bridging to larger open source control layers.
+PyMoDaq \cite{WeberSebastien2021a}
 
-Dedicated client application.
+\section{Case Studies}
 
-Exceedingly simple.
+Introductory paragraph.
 
-(reference publication)
+The Landis Group at UW-Madison created WiQK, a quench kinetics reactor.
+Reactions can be orchestrated with a simple script using Python's built-in time library and yaqc.
+This has worked for a few years as developers continue to modify their reactor.
 
-\subsubsection{Landis}
+The Stahl Group at UW-Madison created a custom reactor which monitors gasses being produced or consumed in the reaction head-space.  \cite{SalazarChaseA2021a}
+From an automation control perspective, this reactor incorporates a collection of sensors and a single heating process value under computer control
+This reactor was made as a self-contained unit with a Raspberry Pi [CITE] as the computer.
+Users interact with a custom Qt-based GUI to use the reactor.
+yaq daemons run locally along-side the lightweight Qt application making the entire system simple to run.
 
-....
-
-\subsubsection{Goldsmith}
-
-Mixing with other hand built or commercially derived software.
-
-\subsubsection{Bluesky}
-
-Bluesky is an extensive open source control layer \cite{AllanDanielB2019a}.
-Bridging to larger open source control layers.
+The Wright Group needs to orchestrate a large variety of hardware in multidimensional scans for their complex spectroscopy experiments.
+We have built out a toolchain incorporating yaq into Bluesky to take advantage of the sophisticated orchestration that Bluesky supports.
+We are excited about Bluesky moving forward.
 

### 2022-04-29 paper.tex

@@
 
-TCP
+yaq is built on top of Apache Avro RPC \cite{AvroSpecification}.
+Avro RPC is a Remote Procedure Call framework which allows interprocess communication between a client application and a daemon \(the server\) which has direct hardware access.
+Avro provides an agreed upon standard for serialization of data and method calls from a remote \(client\) process.
+The protocols defined by Avro are extensible, allowing daemons to define their own data types and declare additional messages that can be processed by the daemon.
+Each daemon provides an Avro Protocol file \(avpr\) which provides a complete list of the types and allowed messages.
 
-\subsection{TCP}
-
-\subsection{Avro RPC}
-
-AVRO-RPC
+While Avro RPC itself is agnostic to transport layer, yaq specifies TCP sockets as the transport layer.
+TCP socket libraries are ubiquitous, allowing for clients (or daemons) to be implemented in a variety of languages if required.
+The use of TCP also works across standard networks, allowing for simple multi-machine instruments.
+yaq uses ``stateful'' Avro RPC to allow multiple requests without requiring additional handshaking.
 
@@
 
-configuration
+In addition to the standard, yaq specifices schema for configuration options and for state that is persistent across restarts of a daemon.
+
+\subsection{Properties and Traits}
 

### 2022-05-13 paper.tex
lots of content

@@
 
+\section{Daemons}
+
+In yaq, each component of an instrument is supported by a tiny lightweight process that runs in the background of your computer: a daemon.
+Since each component has its own process, each daemon can be developed separately.
+Using yaq, a typical instrument might have several daemons each supporting a particular component.
+Daemon A might support a motor, daemon B a sensor, and daemon C a light source.
+To do an experiment, a control program (a "client") must send commands to each of these daemons.
+Each of the daemons is a separate application running in its own process, and importantly the client is also a separate application.
+
+The separation between each daemon and client makes the yaq framework less fragile than monolithic applications.
+The yaq framework is language agnostic.
+For example, a daemon written in Python might be controlled by a client implemented using LabVIEW.
+In yaq, each component of an instrument can be developed and distributed separately.
+For example, two very different instruments might happen to use the same temperature sensor.
+Because the temperature sensor daemon is its own independent program, both instruments can benefit from the same daemon.
+As yaq grows, the "ecosystem" of existing daemons means that future instruments become easier and easier to develop.
+
+There are currently 62 daemons in the yaq project supporting at least 40 kinds of hardware.
+Because yaq is protocol based, anyone can design and publish new daemons extending our hardware support.
+A living list of all daemons and supported hardware can be found on the yaq website.
+
 \section{Protocol}
@@
 
-\subsection{Configuration and State}
-
 In addition to the standard, yaq specifices schema for configuration options and for state that is persistent across restarts of a daemon.
 
-\subsection{Properties and Traits}
+Properties and traits in general.
 
-\section{Daemons}
+\section{Traits}
 
-Daemon roundup.
+Each component of a scientific instrument may have many parameters.
+Similar components typically have similar parameters, but there may be incidental differences between each implementation.
+For example, consider two different monochromators.
+The driver for monochromator A may expose the function "set wavelength", while the driver for monochromator B exposes "set angle".
+Any experimentalist wishing to switch from monochromator A to B must first go through their entire codebase to fix all of the small differences in the driver implementation.
+yaq is designed using an (optional) compositional system that helps enforce consistency between different kinds of hardware.
 
-Discussion of traits and purpose of compositional approach.
-
-\subsection{Traits}
-
-Graphic of traits?
-
-Examples of daemons for core traits.
-
-Fakeable.
+yaq defines "traits": collections of behaviors which are logically grouped together.
+Each trait specifies a set of configuration options, state entries, and methods that a daemon can implement.
+Daemons, then, can be composed of multiple traits.
+Clients can trust that daemons which implement a given trait will behave in similar ways.
+For example, any daemon that implements the "is-position" trait must expose the methods "get\_position" and "set\_position".
+It becomes easy to write generic clients that work with any daemon that has certain traits.
+Of course, daemons are always allowed to implement additional configuration, state, and methods in addition to those which are implied by their traits.
 
@@
 
-Introductory paragraph.
+yaq's simplicity and extensible nature has allowed us to use it in a broad variety of instrumentation designs.
+We highlight a few such designs here.
 
-The Landis Group at UW-Madison created WiQK, a quench kinetics reactor.
-Reactions can be orchestrated with a simple script using Python's built-in time library and yaqc.
-This has worked for a few years as developers continue to modify their reactor.
+The Landis Group at UW-Madison is currently working on a new type of flow reactor: the Wisconsin Quench Kinetics Reactor (WiQK).
+This reactor incorporates several computer-controlled valves and syringe pumps as well as various sensors.
+The WiQK is rapidly evolving as researchers continue to test and refine their design.
+Only a few researchers are actively using the reactor during this prototyping stage.
+The Landis Group has written very basic Python scripts utilizing yaqc to orchestrate hardware for their reactor.
+These lightweight scripts can be extensively refactored as the hardware and orchestration strategy changes dramatically during WiQK development.
+This approach ensures that the Landis Group is not limited by their ability to orchestrate hardware as they refine their reactor design.
 
 The Stahl Group at UW-Madison created a custom reactor which monitors gasses being produced or consumed in the reaction head-space.  \cite{SalazarChaseA2021a}
-From an automation control perspective, this reactor incorporates a collection of sensors and a single heating process value under computer control
-This reactor was made as a self-contained unit with a Raspberry Pi [CITE] as the computer.
-Users interact with a custom Qt-based GUI to use the reactor.
-yaq daemons run locally along-side the lightweight Qt application making the entire system simple to run.
+This reactor incorporates a collection of sensitive pressure transducers and a single heating process value under computer control.
+yaq daemons are used to interface with each sensor and the heater controller.
+We were able to create a lightweight Qt-based GUI which offers an easy user experience to researchers.
+By separating hardware interface code from GUI code, each piece of the system became easier to produce and maintain.
+Taken together, these components have been a solid software experience for many users over several years.
 
-The Wright Group needs to orchestrate a large variety of hardware in multidimensional scans for their complex spectroscopy experiments.
-We have built out a toolchain incorporating yaq into Bluesky to take advantage of the sophisticated orchestration that Bluesky supports.
-We are excited about Bluesky moving forward.
+The Wright Group at UW-Madison needs to orchestrate a large variety of hardware in multidimensional scans for their complex spectroscopy experiments \cite{MukamelShaul2000a, WrightJohnCurtis2011a}.
+This need for exquisite hardware control has resulted in several prior attempts at ``home-built'' orchestration software \cite{CarlsonRogerJohn1988a, MeyerKentAlbert2004b, KainSchuyler2017a, ThompsonBlaiseJonathan2018a}.
+Now, using yaq, the Wright Group has been able to move to Bluesky rather than inventing their own sophisticated control software ``from scratch''.
+Moving forward, the Wright Group hopes to spend less energy developing control software and more energy developing creative spectroscopy experiments.
+
+yaq is also being used in several smaller ways throughout UW-Madison Chemistry.
+Excitingly, all of these research groups are able easily benefit from each-other's daemon developments.
+This level of collaboration is new to us in the orchestration software space.
 

### 2022-07-21 paper.tex
Rework intro and protocol sections

@@
 Instrumentation development is a key part of the scientific enterprise.
-Scientists have always relied on their creativity and wit to assemble new instruments capable of carrying out new experimental procedures.
-As the scientific industry has advanced, scientists are increasingly able to directly purchase pieces of their instruments.
-A modern instrument might contain a motor from company A, a sensor from company B, a specialty light source from company C, and a totally custom robotic sample stage.
-The brand new scientific instrument, doing a brand new experiment, is cobbled together from a few dozen pieces of hardware, some brand new and some reused.
-The instrumental novelty arises from the creative way that all of those pieces are assembled and controlled in the context of the experiment.
-
-Software integration can be a frustrating piece of the modern instrument development process.
-Each individual component has its own drivers and interface.
-A lack of standards means that each scientist must develop her own instrument control application.
+As the scientific industry has advanced, novel instruments are typically constructed of many individual components, purchased and home-built.
+%scientists are increasingly able to directly purchase pieces of their instruments.
+Orchestration software must communicate with each hardware component in the course of a scientific experiment.
+This can involve utilization of many protocols: NI DAQmx [CITE], SCPI [CITE], ModBus [CITE], PICam [CITE], Thorlabs APT [CITE], among many others.
+The challenge of integrating all of these is a frustrating piece of the modern instrument development process.
 Weeks can be spent just integrating one new component into an existing project.
 Code reusability is typically poor, and technical debt grows quickly in academic and educational contexts.
-Sientists may struggle to rapidly innovate on their experimental design when each hardware addition requires major software development.
+Scientists may struggle to rapidly innovate on their experimental design when each hardware addition requires major software development.
 
-Briefly, we
-Minimum viable capabilities for desired protocol.
-Focus on making client development easy.
-\begin{itemize}
-  \item Networked machines (subitem--multilingual)
-  \item Unified interface (traits)
-  \item Self describing
-  \item Built on exisiting protocols
-  \item Reusable
-  \item Synchronous clients
-  \item Portable, easy to install
-  \item Interface agnostic. Not just serial. Certainly not just SCPI.
-\end{itemize}
-Well defined, concrete interface.
-Static interface.
+Some large user-facilities have addressed protocol complexity via the creation of new unified protcols, EPICS \cite{DalesioLR1991a}, TANGO \cite{tango-controls.org}
+The unified protocols define a standard network interface for any hardware component.
+Orchestration software can target this unified protocol for reading and writing hardware state.
+Small background services are written to translate the myriad component protocols into the standard EPICS IOCs and TANGO Devices.
+These programs are peformant, but require expert management to set up and provide descriptions via a separate server program.
+These projects are impressive at facility scale, but in our experience these do not scale well to single-investigator lab environments.
 
-There are several large exisiting open source hardware interface projects, such as EPICS \cite{DalesioLR1991a}, TANGO \cite{tango-controls.org}, and pymodaq \cite{WeberSebastien2021a}.
-These projects are impressive at facility scale, but in our experience these do not scale well to single-investigator lab environments.
-We have aimed to keep yaq as simple as possible, with the goal of staying easy to manage at very small scales without dedicated staff.
-Certainly yaq lacks features compared to these exisiting projects, as we will highlight.
+As smaller research labs have grown in experimental complexity, many hardware abstraction layers have been produced by individual labs.
+Most small custom research instrumentation continues to rely on monolithic home-built orchestration software which has hardcorded protocol support for each particular connected device.
+These monolithic applications tend to be very inflexible and difficult to develop.
+In the last few years, several open source projects by-and-for small-scale experimentalists have grown in popularity [cite many].
+While this growth is encouraging, many of these are limited by their focus on particular types of hardware or particular experimental domains.
+To our knowledge, none of these projects take a strict protocol-based approach.
 
-There are exisiting open source hardware interface projects with a focus on simplicity [cite INSTRBUILDER etc].
-To our knowledge none of these meet the minimal capabilities as outlined above.
+We have created a new protocol [XKCD] for scientific instrumentation, yaq.
+This protocol attempts to borrow the most important ideas from established projects while retaining simplicity appropriate for small research labs.
+We have built this protocol to be self-describing, portable, and reusable wherever possible.
+Our primary goal has been to make an interface which simplifies orchestration software development as much as possible.
+Here we discuss the design of the yaq ecosystem.
+Several case studies provide a view in to the flexible ways that yaq can be applied to perform very different scientific experiments.
 
-Overview of this publication.
-yaq is a protocol.
-yaq is an implementation that runs as a stand-alone daemon.
-yaq has a variety of clients.
-yaq is not opinonated about orchestration layer.
+\section{Protocol}
+
+The yaq protcol is client-server network based.
+To communicate with a given piece of hardware, a client connects to a particular TCP port either locally or to a remote machine over the network.
+Once connected, the client will recieve an exhaustive description of what capabilities that particular hardware interface supports.
+Clients can connect to many TCP ports and each TCP port can simultaniopusly communicate with many clients.
+While Avro RPC itself is agnostic to transport layer, yaq specifies TCP sockets as the transport layer.
+TCP socket libraries are ubiquitous, allowing for clients (or daemons) to be implemented in a variety of languages if required.
+The use of TCP also works across standard networks, allowing for simple multi-machine instruments.
+
+Apache Avro RPC \cite{AvroSpecification} is used to structure the yaq protcol.
+Avro RPC is a Remote Procedure Call framework which allows interprocess communication between a client application and a daemon \(the server\) which has direct hardware access.
+Avro provides an agreed upon standard for efficient serialization of data and method calls from a remote \(client\) process.
+The protocols defined by Avro are extensible, allowing daemons to define their own data types and declare additional messages that can be processed by the daemon.
+Each daemon provides an Avro Protocol file \(avpr\) which provides a complete list of the types and allowed messages.
+yaq uses ``stateful'' Avro RPC to allow multiple requests without requiring additional handshaking.
+
+In practice, each hardware interface is exposed to all clients as a set of well-described function calls.
+
+\section{Traits}
+
+Each component of a scientific instrument may have many parameters.
+Similar components typically have similar parameters, but there may be incidental differences between each implementation.
+For example, consider two different monochromators.
+The driver for monochromator A may expose the function "set wavelength", while the driver for monochromator B exposes "set angle".
+Any experimentalist wishing to switch from monochromator A to B must first go through their entire codebase to fix all of the small differences in the driver implementation.
+yaq is designed using an (optional) compositional system that helps enforce consistency between different kinds of hardware.
+
+yaq defines "traits": collections of behaviors which are logically grouped together.
+Each trait specifies a set of configuration options, state entries, and methods that a daemon can implement.
+Daemons, then, can be composed of multiple traits.
+Clients can trust that daemons which implement a given trait will behave in similar ways.
+For example, any daemon that implements the "is-position" trait must expose the methods "get\_position" and "set\_position".
+It becomes easy to write generic clients that work with any daemon that has certain traits.
+Of course, daemons are always allowed to implement additional configuration, state, and methods in addition to those which are implied by their traits.
 
@@
 A living list of all daemons and supported hardware can be found on the yaq website.
-
-\section{Protocol}
-
-yaq is built on top of Apache Avro RPC \cite{AvroSpecification}.
-Avro RPC is a Remote Procedure Call framework which allows interprocess communication between a client application and a daemon \(the server\) which has direct hardware access.
-Avro provides an agreed upon standard for serialization of data and method calls from a remote \(client\) process.
-The protocols defined by Avro are extensible, allowing daemons to define their own data types and declare additional messages that can be processed by the daemon.
-Each daemon provides an Avro Protocol file \(avpr\) which provides a complete list of the types and allowed messages.
-
-While Avro RPC itself is agnostic to transport layer, yaq specifies TCP sockets as the transport layer.
-TCP socket libraries are ubiquitous, allowing for clients (or daemons) to be implemented in a variety of languages if required.
-The use of TCP also works across standard networks, allowing for simple multi-machine instruments.
-yaq uses ``stateful'' Avro RPC to allow multiple requests without requiring additional handshaking.
-
-In addition to the standard, yaq specifices schema for configuration options and for state that is persistent across restarts of a daemon.
-
-Properties and traits in general.
-
-\section{Traits}
-
-Each component of a scientific instrument may have many parameters.
-Similar components typically have similar parameters, but there may be incidental differences between each implementation.
-For example, consider two different monochromators.
-The driver for monochromator A may expose the function "set wavelength", while the driver for monochromator B exposes "set angle".
-Any experimentalist wishing to switch from monochromator A to B must first go through their entire codebase to fix all of the small differences in the driver implementation.
-yaq is designed using an (optional) compositional system that helps enforce consistency between different kinds of hardware.
-
-yaq defines "traits": collections of behaviors which are logically grouped together.
-Each trait specifies a set of configuration options, state entries, and methods that a daemon can implement.
-Daemons, then, can be composed of multiple traits.
-Clients can trust that daemons which implement a given trait will behave in similar ways.
-For example, any daemon that implements the "is-position" trait must expose the methods "get\_position" and "set\_position".
-It becomes easy to write generic clients that work with any daemon that has certain traits.
-Of course, daemons are always allowed to implement additional configuration, state, and methods in addition to those which are implied by their traits.
 

### 2022-08-02 paper.tex

@@
-\documentclass{article}
-\usepackage[letterpaper, margin=1in]{geometry}
+%  1)  latex  paper
+%  2)  bibtex paper
+%  3)  latex  paper
+%  4)  latex  paper
+%
+\documentclass[aip, amsmath, amssymb, reprint,]{revtex4-1}
 
-\title{The yaq Project: a Protocol for Scientific Instrumentation}
-\author{Kyle F. Sunden,
-  John C. Wright,
-  Blaise J. Thompson}
+\usepackage{graphicx}
+\usepackage{dcolumn}
+\usepackage{bm}
+\usepackage[utf8]{inputenc}
+\usepackage[T1]{fontenc}
+\usepackage{mathptmx}
+\usepackage{etoolbox}
 
+\makeatletter
+\def\@email#1#2{%
+ \endgroup
+ \patchcmd{\titleblock@produce}
+  {\frontmatter@RRAPformat}
+  {\frontmatter@RRAPformat{\produce@RRAP{*#1\href{mailto:#2}{#2}}}\frontmatter@RRAPformat}
+  {}{}
+}%
+\makeatother
 \begin{document}
 
+\preprint{AIP/123-QED}
+
+\title{The yaq Project: \\ Another Protocol for Scientific Instrumentation}
+\author{Kyle F. Sunden}
+\author{John C. Wright}
+\author{Blaise J. Thompson}
+\affiliation{
+  University of Wisconsin--Madison
+}
+\email{blaise.thompson@wisc.edu}
+
+\date{\today}
+
+\begin{abstract}
+An article usually includes an abstract, a concise summary of the work
+covered at length in the main body of the article. It is used for
+secondary publications and for information retrieval purposes.
+\end{abstract}
+
 \maketitle
-
-\section{Abstract}
 
@@
 Instrumentation development is a key part of the scientific enterprise.
-As the scientific industry has advanced, novel instruments are typically constructed of many individual components, purchased and home-built.
-%scientists are increasingly able to directly purchase pieces of their instruments.
+Novel instruments are typically constructed of many individual components, purchased and home-built.
 Orchestration software must communicate with each hardware component in the course of a scientific experiment.
@@
 
-As smaller research labs have grown in experimental complexity, many hardware abstraction layers have been produced by individual labs.
-Most small custom research instrumentation continues to rely on monolithic home-built orchestration software which has hardcorded protocol support for each particular connected device.
+As smaller research labs have grown in experimental complexity, many individual labs have created domain-specific orchestration software.
+Most small custom research instrumentation continues to rely on monolithic software which has hardcorded protocol support for each particular connected device.
 These monolithic applications tend to be very inflexible and difficult to develop.
@@
 
-The yaq protcol is client-server network based.
+\subsection{Architecture}
+
+This section has to introduce the ideas of clients, daemons, and network traffic.
+Should get the multilingual idea here.
+
+Like EPICS and TANGO, the yaq protcol is client-server network based.
 To communicate with a given piece of hardware, a client connects to a particular TCP port either locally or to a remote machine over the network.
-Once connected, the client will recieve an exhaustive description of what capabilities that particular hardware interface supports.
-Clients can connect to many TCP ports and each TCP port can simultaniopusly communicate with many clients.
+Clients can connect to many TCP ports and each TCP port can simultaneously communicate with many clients.
+
+Since each component has its own process, each daemon can be developed separately.
+Daemon A might support a motor, daemon B a sensor, and daemon C a light source.
+To do an experiment, a control program (a "client") must send commands to each of these daemons.
+Each of the daemons is a separate application running in its own process, and importantly the client is also a separate application.
+
+\subsection{Avro}
+
+\begin{figure}
+  \includegraphics[width=\columnwidth]{./figures/methods.png}
+  \caption{  \label{fig:methods} A figure caption. The figure captions are automatically numbered.}
+\end{figure}
+
+Once connected, the client will receive an exhaustive description of what capabilities that particular hardware interface supports.
+Among this description is a machine-readable summary of what the signature of each remote procedure is.
+Figure \ref{fig:methods} summarizes what this might look like.
+Here we see three remote procedures, each with a described set of arguments ``request'' and a described response.
+As the figure shows, these signatures are typed.
+This is only a small piece of the information a client receives.
+
 While Avro RPC itself is agnostic to transport layer, yaq specifies TCP sockets as the transport layer.
@@
 
-In practice, each hardware interface is exposed to all clients as a set of well-described function calls.
+\subsection{Traits}
 
-\section{Traits}
+\begin{figure}
+  \includegraphics[width=\columnwidth]{./figures/traits.png}
+  \caption{  \label{fig:traits} A figure caption. The figure captions are automatically numbered.}
+\end{figure}
 
@@
 yaq is designed using an (optional) compositional system that helps enforce consistency between different kinds of hardware.
+We call this system ``traits''.
 
-yaq defines "traits": collections of behaviors which are logically grouped together.
-Each trait specifies a set of configuration options, state entries, and methods that a daemon can implement.
-Daemons, then, can be composed of multiple traits.
-Clients can trust that daemons which implement a given trait will behave in similar ways.
-For example, any daemon that implements the "is-position" trait must expose the methods "get\_position" and "set\_position".
-It becomes easy to write generic clients that work with any daemon that has certain traits.
-Of course, daemons are always allowed to implement additional configuration, state, and methods in addition to those which are implied by their traits.
+The most basic feature of each trait is a collection of well-defined method signatures, as illustrated by Figure \ref{fig:traits}.
+Here, ``has-position'' is a trait and ``set\_position'' and ``get\_position'' are two method signatures defined by that trait.
+For brevity we only show method names here, but in reality each method signature is fully described as discussed in the previous section.
+Any protocol that wants to implement the ``has-position'' trait must then support those two method signatures.
+Importantly, protocols can also implement arbitrary additional methods that are not defined by any trait.
 
-\section{Daemons}
+The trait system was introduced in pursuit of our primary goal of easing the client development process.
+Clients can trust that protocols that implement a given trait will behave in similar ways.
+In Section \ref{sec:clients} we will show how this system makes it easy to write generic clients for a given experiment.
+Each trait is carefully defined through prose and examples by an enhancement proposal system that helps the community to gain consensus and serves as living documentation.
+These enhancement proposals can be found on the yaq website.
+
+\subsection{Daemons}
+
+\begin{figure}
+  \label{fig:daemons}
+  \includegraphics[width=\columnwidth]{./figures/daemons.png}
+  \caption{ A figure caption. The figure captions are automatically numbered.}
+\end{figure}
 
 In yaq, each component of an instrument is supported by a tiny lightweight process that runs in the background of your computer: a daemon.
-Since each component has its own process, each daemon can be developed separately.
-Using yaq, a typical instrument might have several daemons each supporting a particular component.
-Daemon A might support a motor, daemon B a sensor, and daemon C a light source.
-To do an experiment, a control program (a "client") must send commands to each of these daemons.
-Each of the daemons is a separate application running in its own process, and importantly the client is also a separate application.
+Using yaq, a typical instrument has several daemons each supporting a particular component.
+As illustrated in Figure \ref{fig:daemons}, each of these daemons has a single protocol which is typically composed of multiple traits as well as unique methods.
 
-The separation between each daemon and client makes the yaq framework less fragile than monolithic applications.
-The yaq framework is language agnostic.
-For example, a daemon written in Python might be controlled by a client implemented using LabVIEW.
+Paragraph to describe that methods return quickly and existence of busy.
+
 In yaq, each component of an instrument can be developed and distributed separately.
@@
 
-\section{Clients}
+\subsection{Clients} \label{sec:clients}
+
+\begin{figure}
+  \label{fig:acquisition}
+  \includegraphics[width=\columnwidth]{./figures/acquisition.png}
+  \caption{ A figure caption. The figure captions are automatically numbered.}
+\end{figure}
 
@@
 This modularity speeds development and increases overall flexibility of the system.
-We have created a few client libraries which we will discuss here.
+
+Figure \ref{fig:acquisition} illustrates using pseudocode what a very basic experimental acquisition might look like using yaq.
+Here we have access to the protocols described in Figure \ref{fig:daemons}.
+In the outermost for loop the variable ``pos'' is set to a series of different positions.
+Each loop iteration involves setting a motor, launching a sensor measurement, then recording all independent and dependent variables.
+Because all of these methods are trait-defined, this simple acquisition loop will run as-is for an arbitrary set of motors and sensors.
+While this example is simplified, in practice we have found writing simple scripts to drive experiments to be a very productive approach.
 
@@
 
-\section{Conclusions}
+\section{Conclusion}
 
-\clearpage
-\bibliographystyle{unsrt}
+TODO
+
+\begin{acknowledgments}
+  TODO
+\end{acknowledgments}
+
+\section*{Data Availability Statement}
+
+Data sharing is not applicable to this article as no new data were created or analyzed in this study.
+
+\section*{References}
+
+\nocite{*}
 \bibliography{references}

### 2022-08-16 paper.tex
fill out protocol section

@@
 
-Some large user-facilities have addressed protocol complexity via the creation of new unified protcols, EPICS \cite{DalesioLR1991a}, TANGO \cite{tango-controls.org}
+Some large user-facilities have addressed protocol complexity via the adoption unified protcols, such as EPICS \cite{DalesioLR1991a} or TANGO \cite{AGotz1999TANGOA}.
 The unified protocols define a standard network interface for any hardware component.
@@
 
-This section has to introduce the ideas of clients, daemons, and network traffic.
-Should get the multilingual idea here.
-
 Like EPICS and TANGO, the yaq protcol is client-server network based.
@@
 Each of the daemons is a separate application running in its own process, and importantly the client is also a separate application.
+Since the traffic over the network is standardized, yaq daemons and clients can be written in any programming language.
+The reference implementation is provided in Python [CITE].
 
@@
   \includegraphics[width=\columnwidth]{./figures/methods.png}
-  \caption{  \label{fig:methods} A figure caption. The figure captions are automatically numbered.}
+  \caption{  \label{fig:methods} A description of standard yaq methods, providing their parameters and return value}
 \end{figure}
@@
 Among this description is a machine-readable summary of what the signature of each remote procedure is.
-Figure \ref{fig:methods} summarizes what this might look like.
+Figure \ref{fig:methods} summarizes the information provided about methods exposed by yaq daemons.
 Here we see three remote procedures, each with a described set of arguments ``request'' and a described response.
@@
 This is only a small piece of the information a client receives.
+
+Apache Avro RPC \cite{AvroSpecification} is used to structure the yaq protcol.
+Avro RPC is a Remote Procedure Call framework which allows interprocess communication between a client application and a daemon (the server) which has direct hardware access.
+Avro provides an agreed upon standard for efficient serialization of data and method calls from a remote (client) process.
+The protocols defined by Avro are extensible, allowing daemons to define their own data types and declare additional methods that can be processed by the daemon.
+Upon connecting to a client, each daemon provides an Avro Protocol file (avpr) which provides a complete list of the types and allowed methods in a process known as a "handshake".
+yaq uses ``stateful'' Avro RPC to allow multiple requests without requiring additional handshaking.
 
@@
 
-Apache Avro RPC \cite{AvroSpecification} is used to structure the yaq protcol.
-Avro RPC is a Remote Procedure Call framework which allows interprocess communication between a client application and a daemon \(the server\) which has direct hardware access.
-Avro provides an agreed upon standard for efficient serialization of data and method calls from a remote \(client\) process.
-The protocols defined by Avro are extensible, allowing daemons to define their own data types and declare additional messages that can be processed by the daemon.
-Each daemon provides an Avro Protocol file \(avpr\) which provides a complete list of the types and allowed messages.
-yaq uses ``stateful'' Avro RPC to allow multiple requests without requiring additional handshaking.
-
 \subsection{Traits}
@@
   \includegraphics[width=\columnwidth]{./figures/traits.png}
-  \caption{  \label{fig:traits} A figure caption. The figure captions are automatically numbered.}
+  \caption{  \label{fig:traits} Traits provide collections of RPC methods that are implemented for multiple kinds of hardware.}
 \end{figure}
@@
 For example, consider two different monochromators.
-The driver for monochromator A may expose the function "set wavelength", while the driver for monochromator B exposes "set angle".
+The driver for monochromator A may expose the function "set wavelength", while the driver for monochromator B exposes "set color".
 Any experimentalist wishing to switch from monochromator A to B must first go through their entire codebase to fix all of the small differences in the driver implementation.
@@
 Here, ``has-position'' is a trait and ``set\_position'' and ``get\_position'' are two method signatures defined by that trait.
-For brevity we only show method names here, but in reality each method signature is fully described as discussed in the previous section.
-Any protocol that wants to implement the ``has-position'' trait must then support those two method signatures.
-Importantly, protocols can also implement arbitrary additional methods that are not defined by any trait.
+For brevity we only show method names here, but in reality each method signature is fully described as in Figure \ref{fig:methods}.
+Any daemon that wants to implement the ``has-position'' trait must then support those two method signatures.
+Importantly, specific daemons can also implement arbitrary additional methods that are not defined by any trait.
 
@@
   \includegraphics[width=\columnwidth]{./figures/daemons.png}
-  \caption{ A figure caption. The figure captions are automatically numbered.}
+  \caption{A collection of yaq daemons, highlighting the traits which each daemon implements.}
 \end{figure}
@@
 
-Paragraph to describe that methods return quickly and existence of busy.
+The yaq ecosystem is optimized for ease of writing client applications.
+As such, it is expected that each method call over the yaq interface will return rapidly, ensuring that client applications are not blocked for extended periods of time.
+This principle applies even when the method starts an action that might take several seconds to complete, such as homing a motor or initiating a measurement for a sensor.
+In these instances, the initial method simply starts the action and returns, with a separate method provided to retrieve results if they exist.
+In order to know how long to wait, the ``is-daemon`` trait provides a method called ``is\_busy'' which should return "true" while the long running action is not complete, and "false" once it is finished.
 
@@
   \includegraphics[width=\columnwidth]{./figures/acquisition.png}
-  \caption{ A figure caption. The figure captions are automatically numbered.}
+	\caption{Pseudocode of a simple acquisition with one motor and two sensors as described in Figure \ref{fig:daemons}.}
 \end{figure}
@@
 yaq's RPC design invites the creation of multiple client libraries specialized for specific purposes.
-Again, daemons can communicate simultaniously with multiple clients.
-For a yaq instrument it is normal to have multiple clients running simultaniously, each focused on a particular purpose.
+For a yaq instrument it is normal to have multiple clients running simultaneously, each focused on a particular purpose.
 This modularity speeds development and increases overall flexibility of the system.
@@
 yaqc [CITE] is a very basic Python client for yaq.
-yaqc is totally generic and aims to support any conceivable yaq RPC.
-This library is easy to use directly in Python scripts or within the REPL, but it's also very useful as a dependency when building more specialized Python clients.
+yaqc is totally generic and aims to support any conceivable yaq daemon.
+This library is easy to use directly in Python scripts or interactively, but it's also very useful as a dependency when building more specialized Python clients.
 
 yaqc-qtpy [CITE] is a specialty Qt-based [CITE] GUI client.
-yaqc-qtpy provides a special QClient object which uses Qt threading primatives to provide performant and safe access to yaq from a Qt environment.
-In addition, yaqc-qtpy provides an entry point which automatically builds a GUI from arbitrary protocols using traits.
+yaqc-qtpy provides a ``QClient'' object which uses Qt threading primatives to provide performant and safe access to yaq from a Qt environment.
+In addition, yaqc-qtpy provides a graphical application which automatically populates from arbitrary protocols using traits.
 This is an invaluable tool which provides a ``free'' graphical user interface to any daemon.
 
+Naturally, many scientists would assume that yaq is in competition with other hardware interface layers.
+We, however, do not see yaq as an "either/or" choice, and feel that the self describing nature of yaq coupled with the trait system for shared behavior make for opportunities to bridge between existing projects.
 Bluesky is an extensive open source control layer developed collaboratively across several user facilities \cite{AllanDanielB2019a}.
+yaq provides \texttt{yaqc-bluesky} [CITE] which translates yaq RPC method calls into the equivalent formatting used by Bluesky.
+This allows extraordinary access to the broad array of orchestration and data acquisition capabilities of Bluesky.
+Users of Bluesky can mix-and-match hardware interface layers, and are not required to use only yaq-based hardware.
 
-Bridging to larger open source control layers.
-PyMoDaq \cite{WeberSebastien2021a}
+Additionally, the authors of yaq and the authors of PyMoDaq \cite{WeberSebastien2021a} are in ongoing conversations about how we can best support each other's community and hardware interface options.
 

### 2022-09-08 paper.tex
Rework paper for new strategy

@@
 \usepackage{etoolbox}
+
+\newcommand\yaq{\texttt{yaq}}
 
@@
 
-\section{Protocol}
+\section{Accessibility to Automate}
 
-\subsection{Architecture}
+Complex instrumentation is often driven with custom orchestration software written in LabVIEW or Python.
+This software is powerful, but in our experience it often fails to capture all of the unique hardware.
+Scientists find themselves using multiple pieces of software, some developed by manufacturers and some developed in-house, to control their instrument in a piecemeal way.
+Often, there is no single location where a scientist can address all of their hardware in a fully automated fashion.
+This lack of accessibility to automate limits experimental throughput and creativity, increasing the complexity of the user experience.
 
-Like EPICS and TANGO, the yaq protcol is client-server network based.
+We describe three major barriers to implementing full automation of custom instrumentation.
+First barrier: Multiple incompatible protocols are used to communicate with each component of the system.
+A fully automated system must be able to use all of these protocols, a daunting task for scientists who do not specialize in software development\cite{}.
+Second barrier: Certain specialty hardware require particular hardware and software requirements.
+A specialty camera will only work with an ISA card and drivers for Windows XP.
+A data acquisition manufacturer provides an API that only works in Python 3.7.
+A graduate student wishes to drive some stepper motor using a Raspberry Pi.
+These examples are representative of the challenging barriers that most automation systems have no avenue to address.
+Third barrier: Some hardware interfaces are blocking.
+A LabVIEW interface freezes communication with motors while waiting for a camera to collect data.
+Custom orchestration software needs to be closed before the manufacturers configuration software can be used.
+A graduate student finds themself needing to master advanced concepts in concurrency in order to orchestrate many motors performantly.
+
+\begin{figure}
+\caption{\label{fig:network} Networking diagram}
+\end{figure}
+
+\yaq{} defines a new network-based protocol which addresses these three barriers.
+Figure \ref{fig:network} ...
+
+Like EPICS and TANGO, the \yaq{} protocol is client-server network based.
 To communicate with a given piece of hardware, a client connects to a particular TCP port either locally or to a remote machine over the network.
@@
 
-Since each component has its own process, each daemon can be developed separately.
 Daemon A might support a motor, daemon B a sensor, and daemon C a light source.
@@
 
-\subsection{Avro}
+\section{Experimental Flexibility}
 
-\begin{figure}
-  \includegraphics[width=\columnwidth]{./figures/methods.png}
-  \caption{  \label{fig:methods} A description of standard yaq methods, providing their parameters and return value}
-\end{figure}
+Existing experimental orchestration software is often highly inflexible.
+A laser lab needs to spend weeks on software development when introducing a single new motor into their experimental procedure.
+A researcher needs to shut down an entire instrument when debugging the details of a particular serial interface.
+Researchers are disappointed to realize that they are forced to start from scratch when developing software for a similar instrument built with trivially different hardware.
+An experimentalist will spend many hours in lab to do trivial tasks such as repeated acquisitions because their software does not support repetition and it is too challenging to add the functionality.
 
-Once connected, the client will receive an exhaustive description of what capabilities that particular hardware interface supports.
-Among this description is a machine-readable summary of what the signature of each remote procedure is.
-Figure \ref{fig:methods} summarizes the information provided about methods exposed by yaq daemons.
-Here we see three remote procedures, each with a described set of arguments ``request'' and a described response.
-As the figure shows, these signatures are typed.
-This is only a small piece of the information a client receives.
+In our view, software inflexibility is a natural consequence of the typical software development practices used by custom instrument builders.
+Instrumental software is often built as one monolithic program that does everything from providing a graphical interface, through hardware interfacing, and writing data files.
+Such software is typically impossible to debug without access to real hardware, often requiring all of the hardware to be available to simply start the program.
+As such, instrumentation software development time is in conflict with valuable data acquisition time.
+The hardware interfaces that these programs implement are typically implemented quickly and without regard to standardization with similar hardware.
+The orchestration routines are intimately tied to the particular hardware configuration of one instrument.
 
-Apache Avro RPC \cite{AvroSpecification} is used to structure the yaq protcol.
+Unique experiments will always need custom orchestration and user experience.
+We believe that novel instrumentation development naturally and necessarily includes the creation of targeted user interfaces.
+\yaq{} is architected to encourage better software development practices when creating such interfaces.
+In a \yaq{} context, orchestration code and graphical control interfaces are implemented as clients.
+These clients are automatically simpler because they only need to implement the \yaq{} protocol and do not need to include the vast array of hardware-specific communication protocols.
+Beyond this, the \yaq{} protocol is standardized to provide a common interface for similar hardware.
+A client written to perform a two dimensional fluorescence experiment using two Acton monochromators will also work with Horiba monochromators without any modification.
+
+Apache Avro RPC \cite{AvroSpecification} is used to structure the \yaq{} protocol.
 Avro RPC is a Remote Procedure Call framework which allows interprocess communication between a client application and a daemon (the server) which has direct hardware access.
 Avro provides an agreed upon standard for efficient serialization of data and method calls from a remote (client) process.
-The protocols defined by Avro are extensible, allowing daemons to define their own data types and declare additional methods that can be processed by the daemon.
-Upon connecting to a client, each daemon provides an Avro Protocol file (avpr) which provides a complete list of the types and allowed methods in a process known as a "handshake".
-yaq uses ``stateful'' Avro RPC to allow multiple requests without requiring additional handshaking.
+Practically, the \yaq{} protocol looks like a collection of methods or functions, which Avro calls ``messages''.
+Each message has defined input parameters and output return types.
+A sensor might implement a message called ``\texttt{get\_measured}'' which takes no parameters and returns a dictionary mapping channel names to numeric or array data.
+A motor would implement a pair of messages for setting and reading back the motor position: ``\texttt{set\_position(float position)} $\rightarrow$ \texttt{null} '' and ``\texttt{get\_position()} $\rightarrow$ \texttt{float}''.
+These self describing messages make up the lowest level functionality of the \yaq{} protocol.
 
-While Avro RPC itself is agnostic to transport layer, yaq specifies TCP sockets as the transport layer.
-TCP socket libraries are ubiquitous, allowing for clients (or daemons) to be implemented in a variety of languages if required.
-The use of TCP also works across standard networks, allowing for simple multi-machine instruments.
+yaq's RPC design invites the creation of multiple client libraries specialized for specific purposes.
+For a yaq instrument it is normal to have multiple clients running simultaneously, each focused on a particular purpose.
+This modularity speeds development and increases overall flexibility of the system.
 
-\subsection{Traits}
-
-\begin{figure}
-  \includegraphics[width=\columnwidth]{./figures/traits.png}
-  \caption{  \label{fig:traits} Traits provide collections of RPC methods that are implemented for multiple kinds of hardware.}
-\end{figure}
-
-Each component of a scientific instrument may have many parameters.
-Similar components typically have similar parameters, but there may be incidental differences between each implementation.
-For example, consider two different monochromators.
-The driver for monochromator A may expose the function "set wavelength", while the driver for monochromator B exposes "set color".
-Any experimentalist wishing to switch from monochromator A to B must first go through their entire codebase to fix all of the small differences in the driver implementation.
-yaq is designed using an (optional) compositional system that helps enforce consistency between different kinds of hardware.
-We call this system ``traits''.
-
-The most basic feature of each trait is a collection of well-defined method signatures, as illustrated by Figure \ref{fig:traits}.
-Here, ``has-position'' is a trait and ``set\_position'' and ``get\_position'' are two method signatures defined by that trait.
-For brevity we only show method names here, but in reality each method signature is fully described as in Figure \ref{fig:methods}.
-Any daemon that wants to implement the ``has-position'' trait must then support those two method signatures.
-Importantly, specific daemons can also implement arbitrary additional methods that are not defined by any trait.
-
+\yaq{} introduces a concept called ``Traits'' which are collections of related messages that are shared among multiple protocols.
+Motors implement the ``has-position'' trait, which defines ``\texttt{set\_position}'', ``\texttt{get\_position}'', and ``\texttt{get\_units}''.
+Sensors would implement the ``is-sensor'' trait which defines ``\texttt{get\_measured}'', ``\texttt{get\_channel\_names}'', and ``\texttt{get\_channel\_units}''. 
+Protocols which implement a trait must support all of the messages from the trait.
 The trait system was introduced in pursuit of our primary goal of easing the client development process.
 Clients can trust that protocols that implement a given trait will behave in similar ways.
-In Section \ref{sec:clients} we will show how this system makes it easy to write generic clients for a given experiment.
-Each trait is carefully defined through prose and examples by an enhancement proposal system that helps the community to gain consensus and serves as living documentation.
-These enhancement proposals can be found on the yaq website.
-
-\subsection{Daemons}
-
-\begin{figure}
-  \label{fig:daemons}
-  \includegraphics[width=\columnwidth]{./figures/daemons.png}
-  \caption{A collection of yaq daemons, highlighting the traits which each daemon implements.}
-\end{figure}
-
-In yaq, each component of an instrument is supported by a tiny lightweight process that runs in the background of your computer: a daemon.
-Using yaq, a typical instrument has several daemons each supporting a particular component.
-As illustrated in Figure \ref{fig:daemons}, each of these daemons has a single protocol which is typically composed of multiple traits as well as unique methods.
+Importantly, specific protocols can also implement arbitrary additional messages that are not defined by any trait.
 
@@
 
-In yaq, each component of an instrument can be developed and distributed separately.
-For example, two very different instruments might happen to use the same temperature sensor.
-Because the temperature sensor daemon is its own independent program, both instruments can benefit from the same daemon.
-As yaq grows, the "ecosystem" of existing daemons means that future instruments become easier and easier to develop.
-
-There are currently 62 daemons in the yaq project supporting at least 40 kinds of hardware.
-Because yaq is protocol based, anyone can design and publish new daemons extending our hardware support.
-A living list of all daemons and supported hardware can be found on the yaq website.
-
-\subsection{Clients} \label{sec:clients}
-
-\begin{figure}
-  \label{fig:acquisition}
-  \includegraphics[width=\columnwidth]{./figures/acquisition.png}
-	\caption{Pseudocode of a simple acquisition with one motor and two sensors as described in Figure \ref{fig:daemons}.}
-\end{figure}
-
-yaq's RPC design invites the creation of multiple client libraries specialized for specific purposes.
-For a yaq instrument it is normal to have multiple clients running simultaneously, each focused on a particular purpose.
-This modularity speeds development and increases overall flexibility of the system.
-
-Figure \ref{fig:acquisition} illustrates using pseudocode what a very basic experimental acquisition might look like using yaq.
-Here we have access to the protocols described in Figure \ref{fig:daemons}.
-In the outermost for loop the variable ``pos'' is set to a series of different positions.
-Each loop iteration involves setting a motor, launching a sensor measurement, then recording all independent and dependent variables.
-Because all of these methods are trait-defined, this simple acquisition loop will run as-is for an arbitrary set of motors and sensors.
-While this example is simplified, in practice we have found writing simple scripts to drive experiments to be a very productive approach.
+Several general purpose \yaq{} clients exist.
+\texttt{yaqc}\cite{} is a lightweight Python client which is excellent for using in scripts or any other Python program.
+\texttt{yaqc-qtpy}\cite{} is a graphical application which builds interactive controls based on traits for any conceivable \yaq{} protocol.
+\texttt{yaqc-bluesky}\cite{} provides a bridge to the Bluesky ecosystem\cite{}.
+Bluesky provides a powerful orchestration layer for conducting and recording data for a wide variety of experimental procedures.
+Similar translation layers could be built for a variety of orchestration layers such as PyMoDAQ\cite{}, Instrumental\cite{}, or TRSpectrometer\cite{}.
 
@@
 
-Naturally, many scientists would assume that yaq is in competition with other hardware interface layers.
-We, however, do not see yaq as an "either/or" choice, and feel that the self describing nature of yaq coupled with the trait system for shared behavior make for opportunities to bridge between existing projects.
-Bluesky is an extensive open source control layer developed collaboratively across several user facilities \cite{AllanDanielB2019a}.
-yaq provides \texttt{yaqc-bluesky} [CITE] which translates yaq RPC method calls into the equivalent formatting used by Bluesky.
-This allows extraordinary access to the broad array of orchestration and data acquisition capabilities of Bluesky.
-Users of Bluesky can mix-and-match hardware interface layers, and are not required to use only yaq-based hardware.
+Figure \ref{fig:foundation}...
+Cover scripts, fakes, escape hatches.
+non-excusivity
 
-Additionally, the authors of yaq and the authors of PyMoDaq \cite{WeberSebastien2021a} are in ongoing conversations about how we can best support each other's community and hardware interface options.
+\begin{figure}
+  \caption{  \label{fig:foundation} A caption}
+\end{figure}
+
+\section{Hardware Flexibility}
+
+Research groups tend to be understandably apprehensive about incorporating new hardware into their instrument.
+We previously discussed this issue from the perspective of orchestration and user interface, we now approach the same issue from the perspective of needing to master the new hardware interface itself.
+It is natural to focus more on excellent client experiences because client development is something that every \yaq{} user needs to become familiar with.
+A \yaq{} daemon needs only be implemented once for all instruments, while each specific experiment requires specific client implementation.
+We have worked to make daemon development simple where possible.
+
+The \yaq{} architecture automatically simplifies hardware interface development in several ways.
+First, because daemons are separate and portable programs, the development effort can be spread across the community of \yaq{} users.
+Second, \yaq{} daemon development can be performed separated from the particulars of any individual client.
+Often, this means that initial hardware enablement work can be done on a researchers personal machine before the new hardware peripheral is installed in the instrument.
+Third, when developing trait-compliant protocols, it becomes easy to design and fully test your hardware interface.
+In this instance, you even can use \texttt{yaqc-qtpy} to provide a graphical program to interact with your hardware immediately.
+Finally, as discussed in Section II, \yaq{} gives you options to design using remote hardware or unusual interfaces when necessary.
+
+We have created several tools to aide in daemon development.
+First, a Python library, \texttt{yaqd-core}\cite{}, which implements shared functionality.
+Simple interfaces, such as Brooks MFC\cite{}, can be implemented in as little as 50 lines of Python code.
+Second, \texttt{yaq-traits}\cite{} is a command line application which allows the description of messages provided by a \yaq{} protocol to be written in a human-readable fashion and translated into a more fully described machine readable format.
+The format it generates is an important part of how \yaq{} protocols are self describing.
+This shields developers from the details of Apache Avro, which can be somewhat esoteric.
+
+\section{Technical Debt}
+
+Years after the original researchers leave, large monolithic acquisition programs become unknowable, undocumented, and unmaintained.
+A graduate student discovers a long standing hard-coded conversion factor that is incorrect years after implementation.
+Scientists resort to sourcing a replacement for an old, broken oscilloscope due to their reliance of their software on that particular interface; newer, cheaper options are readily available.
+A graduate student is forced to meticulously reverse engineer the LabView codebase that they inherited in order to understand the details of their experiment.
+Technical debt grows especially fast in academic environments where graduate students are constantly being replaced.
+
+We have found that the \yaq{} approach favors many small single-purpose applications above large monolithic ones.
+For daemons, the purpose of each application is obvious and unambiguous. 
+There is a strict, well defined interface which explicitly limits the kinds of interactions that are provided to the hardware, thus limiting opportunity for unintended consequences.
+The lack of hardware interface code makes \yaq{} clients much simpler and easier to describe and maintain.
+Tools like \texttt{yaqd-fakes} allow clients to be tested and improved outside of their instrument, including the possibility of fully automated testing\cite{}.
+Simple, script-based clients written using the expressiveness of Python can be read and understood in hours rather than weeks.
+
+In \yaq{}, each component of an instrument can be developed and distributed separately.
+For example, two very different instruments might happen to use the same temperature sensor.
+Because the temperature sensor daemon is its own independent program, both instruments can benefit from the same daemon.
+As \yaq{} grows, the "ecosystem" of existing daemons means that future instruments become easier and easier to develop.
+Growing this ecosystem is a collaborative effort where many \yaq{} users create portable daemons that they need and share them for the community to use and improve.
+
+There are currently 62 daemons in the yaq project supporting at least 40 kinds of hardware.
+Because yaq is protocol based, anyone can design and publish new daemons extending our hardware support.
+A living list of all daemons and supported hardware can be found on the yaq website.
+
+Software documentation is famously difficult and thankless work.
+\yaq{} attempts do automate daemon documentation as much as possible.
+Our website, \href{https://yaq.fyi}, automatically builds generated protocol reference pages for all known protocols.
+These pages are automatically updated when new versions are published.
 

### 2022-09-15 paper.tex
lots of edits

@@
 
-\section{Accessibility to Automate}
-
 Complex instrumentation is often driven with custom orchestration software written in LabVIEW or Python.
@@
 
-We describe three major barriers to implementing full automation of custom instrumentation.
+\section{Hardware Interface Challenges}
+
+Hardware-software interface design presents several technical challenges.
+In this section we describe the architecture of the \yaq{} framework in light of three major barriers that we have encountered in scientific instrumentation development.
 First barrier: Multiple incompatible protocols are used to communicate with each component of the system.
 A fully automated system must be able to use all of these protocols, a daunting task for scientists who do not specialize in software development\cite{}.
-Second barrier: Certain specialty hardware require particular hardware and software requirements.
+Second barrier: Certain specialty hardware have inconvenient interface requirements.
 A specialty camera will only work with an ISA card and drivers for Windows XP.
@@
 A graduate student wishes to drive some stepper motor using a Raspberry Pi.
-These examples are representative of the challenging barriers that most automation systems have no avenue to address.
 Third barrier: Some hardware interfaces are blocking.
-A LabVIEW interface freezes communication with motors while waiting for a camera to collect data.
+A graphical user interface while waiting for a camera to collect data.
 Custom orchestration software needs to be closed before the manufacturers configuration software can be used.
@@
 \begin{figure}
+	\includegraphics[width=\columnwidth]{figures/network}
 \caption{\label{fig:network} Networking diagram}
@@
 
-\yaq{} defines a new network-based protocol which addresses these three barriers.
-Figure \ref{fig:network} ...
+Figure \ref{fig:network} diagrams the \yaq{} architecture.
+Here we show three different computers connected via an ethernet network.
+The top and bottom computer are connected to monitors for interactive use while the middle computer is only accessible via the network.
+This might represent a complex scientific instrument involving several operator terminals as well as embedded computers.
+At the top, a single computer is connected to four hardware peripherals through RS232, TTL, USB, and PCI as indicated by the colored lines.
+That same computer is running four separate programs, one for each peripheral.
+These small, targeted, programs are managed by the operating system and run in the background.
+It is conventional to call such programs ``daemons''\cite{}.
+The middle computer is connected two two additional peripherals, and runs daemons for each.
+Besides communicating with the hardware peripheral, each daemon can communicate with other programs, ``clients'',  through the network.
+The four client programs shown in Figure \ref{fig:network} can each communicate with all six hardware peripherals shown.
+As an example, a client running on the bottom computer could communicate with the RS232 peripheral shown in green via the following path: client $\leftrightarrow$ network switch $\leftrightarrow$ top computer $\leftrightarrow$ daemon $\leftrightarrow$ hardware peripheral.
+This powerful architecture can scale down to single computers or up to many computers, including fully remote operator interfaces.
+As we will show, usage of standards and the creation of tooling makes this architecture accessible to instrument builders.
 
-Like EPICS and TANGO, the \yaq{} protocol is client-server network based.
-To communicate with a given piece of hardware, a client connects to a particular TCP port either locally or to a remote machine over the network.
-Clients can connect to many TCP ports and each TCP port can simultaneously communicate with many clients.
+In \yaq{}, communication between daemons and clients is performed over TCP/IP using a protocol called Apache Avro RPC \cite{AvroSpecification}.
+Avro provides an agreed upon standard for efficient serialization of data and method calls from a remote (client) process.
+Practically, the \yaq{} interface looks like a collection of methods or functions, which Avro calls ``messages''.
+Each message has defined input parameters and output return types.
+A sensor might implement a message called ``\texttt{get\_measured}'' which takes no parameters and returns a dictionary mapping channel names to numeric or array data.
+A motor would implement a pair of messages for setting and reading back the motor position: ``\texttt{set\_position(float position)} $\rightarrow$ \texttt{null} '' and ``\texttt{get\_position()} $\rightarrow$ \texttt{float}''.
+These self describing messages make up the lowest level functionality of the \yaq{} interface.
+Each individual communication between client and daemon involves one message being requested by the client and the response returned from the daemon.
+Each daemon supports a collection of messages for its unique functionality, called the ``protocol''.
 
-Daemon A might support a motor, daemon B a sensor, and daemon C a light source.
-To do an experiment, a control program (a "client") must send commands to each of these daemons.
-Each of the daemons is a separate application running in its own process, and importantly the client is also a separate application.
-Since the traffic over the network is standardized, yaq daemons and clients can be written in any programming language.
-The reference implementation is provided in Python [CITE].
+\yaq{} introduces a concept called ``Traits'' which are collections of related messages that are shared among multiple protocols.
+Motors implement the ``has-position'' trait, which defines ``\texttt{set\_position}'', ``\texttt{get\_position}'', and ``\texttt{get\_units}''.
+Sensors would implement the ``is-sensor'' trait which defines ``\texttt{get\_measured}'', ``\texttt{get\_channel\_names}'', and ``\texttt{get\_channel\_units}''. 
+Protocols which implement a trait must support all of the messages from the trait.
+Importantly, specific protocols can also implement arbitrary additional messages that are not defined by any trait.
+
+The first hardware interface barrier: Multiple incompatible protocols are used to communicate with each component of the system.
+\yaq{} provides a unified TCP/IP interface to all hardware peripherals based on the well-described Avro RPC protocol.
+In particular, the trait system was introduced in pursuit of our primary goal of easing the client development process.
+Clients can trust that protocols that implement a given trait will behave in similar ways.
+
+The second hardware interface barrier: Certain specialty hardware have inconvenient interface requirements.
+Since \yaq{} enables multiple machines, any hardware requirements can be addressed by putting a machine for that specific hardware on the network.
+For example, the Windows XP computer which has the interface card and drivers can be placed into a private network with only \yaq{} interface communication.
+Because each \yaq{} daemon is running in its own process, the software environment can be tailored to its needs.
+A client running up to date Python 3.10 communicates seamlessly with a daemon running Python 3.7.
+
+The third hardware interface barrier: Some hardware interfaces are blocking.
+Each \yaq{} daemon runs as a separate process (possibly not even on the same machine) and so many of the hard concurrency problems with communicating with hardware are mitigated.
+It is expected that each message call over the \yaq{} interface will return rapidly, ensuring that client applications are not blocked for extended periods of time.
+This principle applies even when the message starts an action that might take several seconds to complete, such as homing a motor or initiating a measurement for a sensor.
+In these instances, the initial message simply starts the action and returns, with a separate message provided to retrieve results when they exist.
+In order to know how long to wait, the ``is-daemon`` trait provides a message called ``is\_busy'' which should return "true" while the long running action is not complete, and "false" once it is finished.
+Additionally, multiple clients can communicate with the same daemon simultaneously.
+A complex instrument may involve multiple operators watching sensor data in real time, while one program is orchestrating the hardware and recording the data.
 
@@
 Existing experimental orchestration software is often highly inflexible.
-A laser lab needs to spend weeks on software development when introducing a single new motor into their experimental procedure.
-A researcher needs to shut down an entire instrument when debugging the details of a particular serial interface.
+An experimentalist will spend many hours in lab repeating acquisitions because it is too challenging to add repetition functionality to their software.
+A laser lab needs to spend weeks on software development when introducing a single new step into their experimental procedure.
 Researchers are disappointed to realize that they are forced to start from scratch when developing software for a similar instrument built with trivially different hardware.
-An experimentalist will spend many hours in lab to do trivial tasks such as repeated acquisitions because their software does not support repetition and it is too challenging to add the functionality.
 
@@
 Unique experiments will always need custom orchestration and user experience.
-We believe that novel instrumentation development naturally and necessarily includes the creation of targeted user interfaces.
-\yaq{} is architected to encourage better software development practices when creating such interfaces.
+We believe that novel instrumentation development naturally and necessarily includes the creation of targeted software.
+\yaq{} is architected to encourage better software development practices when creating such programs.
 In a \yaq{} context, orchestration code and graphical control interfaces are implemented as clients.
-These clients are automatically simpler because they only need to implement the \yaq{} protocol and do not need to include the vast array of hardware-specific communication protocols.
-Beyond this, the \yaq{} protocol is standardized to provide a common interface for similar hardware.
+These clients are automatically simpler because they only need to implement the \yaq{} interface and do not need to include the vast array of hardware-specific communication protocols.
+Beyond this, clients can use traits to interact with similar hardware identically.
 A client written to perform a two dimensional fluorescence experiment using two Acton monochromators will also work with Horiba monochromators without any modification.
 
-Apache Avro RPC \cite{AvroSpecification} is used to structure the \yaq{} protocol.
-Avro RPC is a Remote Procedure Call framework which allows interprocess communication between a client application and a daemon (the server) which has direct hardware access.
-Avro provides an agreed upon standard for efficient serialization of data and method calls from a remote (client) process.
-Practically, the \yaq{} protocol looks like a collection of methods or functions, which Avro calls ``messages''.
-Each message has defined input parameters and output return types.
-A sensor might implement a message called ``\texttt{get\_measured}'' which takes no parameters and returns a dictionary mapping channel names to numeric or array data.
-A motor would implement a pair of messages for setting and reading back the motor position: ``\texttt{set\_position(float position)} $\rightarrow$ \texttt{null} '' and ``\texttt{get\_position()} $\rightarrow$ \texttt{float}''.
-These self describing messages make up the lowest level functionality of the \yaq{} protocol.
+A \yaq{} client can scale in complexity from a small, lightweight script all the way to a complicated graphical program, as shown in Figure \ref{fig:foundation}.
+At the bottom, the daemons lay a solid foundation to build the variety of clients on top of.
+Client sophistication can be introduced naturally, as novel experiments are tested and refined.
 
-yaq's RPC design invites the creation of multiple client libraries specialized for specific purposes.
-For a yaq instrument it is normal to have multiple clients running simultaneously, each focused on a particular purpose.
-This modularity speeds development and increases overall flexibility of the system.
-
-\yaq{} introduces a concept called ``Traits'' which are collections of related messages that are shared among multiple protocols.
-Motors implement the ``has-position'' trait, which defines ``\texttt{set\_position}'', ``\texttt{get\_position}'', and ``\texttt{get\_units}''.
-Sensors would implement the ``is-sensor'' trait which defines ``\texttt{get\_measured}'', ``\texttt{get\_channel\_names}'', and ``\texttt{get\_channel\_units}''. 
-Protocols which implement a trait must support all of the messages from the trait.
-The trait system was introduced in pursuit of our primary goal of easing the client development process.
-Clients can trust that protocols that implement a given trait will behave in similar ways.
-Importantly, specific protocols can also implement arbitrary additional messages that are not defined by any trait.
-
-The yaq ecosystem is optimized for ease of writing client applications.
-As such, it is expected that each method call over the yaq interface will return rapidly, ensuring that client applications are not blocked for extended periods of time.
-This principle applies even when the method starts an action that might take several seconds to complete, such as homing a motor or initiating a measurement for a sensor.
-In these instances, the initial method simply starts the action and returns, with a separate method provided to retrieve results if they exist.
-In order to know how long to wait, the ``is-daemon`` trait provides a method called ``is\_busy'' which should return "true" while the long running action is not complete, and "false" once it is finished.
+\begin{figure}
+\includegraphics[width=\columnwidth]{figures/clients}
+  \caption{  \label{fig:foundation} A caption}
+\end{figure}
 
@@
 \texttt{yaqc}\cite{} is a lightweight Python client which is excellent for using in scripts or any other Python program.
+\texttt{yaqc} is totally generic and aims to support any conceivable \yaq{} daemon.
 \texttt{yaqc-qtpy}\cite{} is a graphical application which builds interactive controls based on traits for any conceivable \yaq{} protocol.
+This is an invaluable tool which provides a ``free'' graphical user interface to any daemon.
 \texttt{yaqc-bluesky}\cite{} provides a bridge to the Bluesky ecosystem\cite{}.
@@
 
-yaqc [CITE] is a very basic Python client for yaq.
-yaqc is totally generic and aims to support any conceivable yaq daemon.
-This library is easy to use directly in Python scripts or interactively, but it's also very useful as a dependency when building more specialized Python clients.
+The Landis Group at UW-Madison is currently working on a new type of flow reactor: the Wisconsin Quench Kinetics Reactor (WiQK).
+This reactor incorporates several computer-controlled valves and syringe pumps as well as various sensors.
+The set of hardware peripherals is rapidly changing as researchers continue to test and refine their design.
+Only a few researchers are actively using the reactor during this prototyping stage.
+These researchers are experimentalists who have limited background in software development.
+The Landis Group has written very basic Python scripts to orchestrate hardware for their reactor.
+These lightweight scripts can be extensively refactored by the experimentalists as the hardware and orchestration strategy changes dramatically during WiQK development.
+This approach ensures that the Landis Group is not slowed down by complex, inflexible orchestration software.
+Once the reactor is complete, more sophisticated graphical clients will be created to accommodate end users who were not involved as the reactor was built.
 
-yaqc-qtpy [CITE] is a specialty Qt-based [CITE] GUI client.
-yaqc-qtpy provides a ``QClient'' object which uses Qt threading primatives to provide performant and safe access to yaq from a Qt environment.
-In addition, yaqc-qtpy provides a graphical application which automatically populates from arbitrary protocols using traits.
-This is an invaluable tool which provides a ``free'' graphical user interface to any daemon.
+The Wright Group at UW-Madison needs to orchestrate a large variety of hardware in multidimensional scans for their complex spectroscopy experiments \cite{MukamelShaul2000a, WrightJohnCurtis2011a}.
+This need for exquisite hardware control has resulted in several prior attempts at ``home-built'' orchestration software \cite{CarlsonRogerJohn1988a, MeyerKentAlbert2004b, KainSchuyler2017a, ThompsonBlaiseJonathan2018a}.
+Now, using \yaq{}, the Wright Group has been able to move to Bluesky rather than inventing their own sophisticated control software ``from scratch''.
+The Wright Group uses simulated hardware to enable client development away from the active laboratory computers.
+The same client is used on four laser systems, each with different complements of hardware.
+Moving forward, the Wright Group hopes to spend less energy developing control software and more energy developing creative spectroscopy experiments.
 
-Figure \ref{fig:foundation}...
-Cover scripts, fakes, escape hatches.
-non-excusivity
+\section{Incorporating New Hardware}
 
-\begin{figure}
-  \caption{  \label{fig:foundation} A caption}
-\end{figure}
-
-\section{Hardware Flexibility}
-
-Research groups tend to be understandably apprehensive about incorporating new hardware into their instrument.
-We previously discussed this issue from the perspective of orchestration and user interface, we now approach the same issue from the perspective of needing to master the new hardware interface itself.
-It is natural to focus more on excellent client experiences because client development is something that every \yaq{} user needs to become familiar with.
-A \yaq{} daemon needs only be implemented once for all instruments, while each specific experiment requires specific client implementation.
-We have worked to make daemon development simple where possible.
-
+In a \yaq{} context, new hardware can be incorporated into an instrument through the creation of a new daemon.
 The \yaq{} architecture automatically simplifies hardware interface development in several ways.
@@
 This shields developers from the details of Apache Avro, which can be somewhat esoteric.
+
+The Stahl Group at UW-Madison created a custom reactor which monitors gasses being produced or consumed in the reaction head-space.  \cite{SalazarChaseA2021a}
+This reactor incorporates a collection of sensitive pressure transducers and a single heating process value under computer control.
+yaq daemons are used to interface with each sensor and the heater controller.
+Recently, experimentalists have been attempting reactions involving smaller, slower, pressure changes.
+A fundamental flaw in the initial analog to digital converter board was revealed by these attempts.
+As a result, a new digitizer has been purchased.
+This new converter will be incorporated into the existing reactor without modifying the existing graphical user interface and data recording program, minimizing downtime.
 
@@
 
-There are currently 62 daemons in the yaq project supporting at least 40 kinds of hardware.
+There are currently XX daemons in the yaq project supporting at least YY kinds of hardware.
 Because yaq is protocol based, anyone can design and publish new daemons extending our hardware support.
@@
 \yaq{} attempts do automate daemon documentation as much as possible.
-Our website, \href{https://yaq.fyi}, automatically builds generated protocol reference pages for all known protocols.
+Our website, https://yaq.fyi, automatically builds generated protocol reference pages for all known protocols.
 These pages are automatically updated when new versions are published.
-
-\section{Case Studies}
-
-yaq's simplicity and extensible nature has allowed us to use it in a broad variety of instrumentation designs.
-We highlight a few such designs here.
-
-The Landis Group at UW-Madison is currently working on a new type of flow reactor: the Wisconsin Quench Kinetics Reactor (WiQK).
-This reactor incorporates several computer-controlled valves and syringe pumps as well as various sensors.
-The WiQK is rapidly evolving as researchers continue to test and refine their design.
-Only a few researchers are actively using the reactor during this prototyping stage.
-The Landis Group has written very basic Python scripts utilizing yaqc to orchestrate hardware for their reactor.
-These lightweight scripts can be extensively refactored as the hardware and orchestration strategy changes dramatically during WiQK development.
-This approach ensures that the Landis Group is not limited by their ability to orchestrate hardware as they refine their reactor design.
-
-The Stahl Group at UW-Madison created a custom reactor which monitors gasses being produced or consumed in the reaction head-space.  \cite{SalazarChaseA2021a}
-This reactor incorporates a collection of sensitive pressure transducers and a single heating process value under computer control.
-yaq daemons are used to interface with each sensor and the heater controller.
-We were able to create a lightweight Qt-based GUI which offers an easy user experience to researchers.
-By separating hardware interface code from GUI code, each piece of the system became easier to produce and maintain.
-Taken together, these components have been a solid software experience for many users over several years.
-
-The Wright Group at UW-Madison needs to orchestrate a large variety of hardware in multidimensional scans for their complex spectroscopy experiments \cite{MukamelShaul2000a, WrightJohnCurtis2011a}.
-This need for exquisite hardware control has resulted in several prior attempts at ``home-built'' orchestration software \cite{CarlsonRogerJohn1988a, MeyerKentAlbert2004b, KainSchuyler2017a, ThompsonBlaiseJonathan2018a}.
-Now, using yaq, the Wright Group has been able to move to Bluesky rather than inventing their own sophisticated control software ``from scratch''.
-Moving forward, the Wright Group hopes to spend less energy developing control software and more energy developing creative spectroscopy experiments.
 

### 2022-09-20 paper.tex
Add authors
conclusion

@@
 
-\title{The yaq Project: \\ Another Protocol for Scientific Instrumentation}
+\title{The yaq Project: \\ A Protocol for Scientific Instrumentation}
 \author{Kyle F. Sunden}
+\author{Daniel D. Kohler}
+\author{Peter C. Parilla}
+\author{Kent A. Meyer}
 \author{John C. Wright}
@@
 Our primary goal has been to make an interface which simplifies orchestration software development as much as possible.
-Here we discuss the design of the yaq ecosystem.
-Several case studies provide a view in to the flexible ways that yaq can be applied to perform very different scientific experiments.
 
-Complex instrumentation is often driven with custom orchestration software written in LabVIEW or Python.
-This software is powerful, but in our experience it often fails to capture all of the unique hardware.
-Scientists find themselves using multiple pieces of software, some developed by manufacturers and some developed in-house, to control their instrument in a piecemeal way.
-Often, there is no single location where a scientist can address all of their hardware in a fully automated fashion.
-This lack of accessibility to automate limits experimental throughput and creativity, increasing the complexity of the user experience.
+Here we discuss the design of the yaq protocol in the context of four challenges facing instrument designers.
+First, we discuss how particularly challenging hardware interfaces can become seemingly insurmountable barriers to software control.
+Next, we discuss how inflexible orchestration software can limit experimental flexibility.
+Next, we focus on challenges that arise when enhancing or modifying existing instruments with new hardware.
+Finally, we discuss the heavy software maitenence burden that many instrument designers face.
+In each case we will highlight how the yaq project is designed to aliveate alleviate that challenge.
+Several case studies provide a view into the flexible ways that yaq can be applied to perform very different scientific experiments.
 
@@
 Third barrier: Some hardware interfaces are blocking.
-A graphical user interface while waiting for a camera to collect data.
+A graphical user interface stalls while waiting for a camera to collect data.
 Custom orchestration software needs to be closed before the manufacturers configuration software can be used.
@@
 This powerful architecture can scale down to single computers or up to many computers, including fully remote operator interfaces.
-As we will show, usage of standards and the creation of tooling makes this architecture accessible to instrument builders.
+Superficially, this network-based client-server design is identical to EPICS and TANGO [CITE].
+As we will show, usage of standards and the creation of tooling makes this architecture accessible to instrument builders outside of large facilities.
 
@@
 
-TODO
+The \yaq{} project defines a new general-purpose protocol for hardware control in the context of scientific instrumentation.
+This protocol has some of the powerful features of facility-scale protocols while remaining simple enough for feasable implementation and maintenance in small research labs.
+We have shown how this approach alleviates common problems through discussion and case studies.
+Designing around self-describing protocols is a productive approach that has great promise in scientific software deveopment.
 

### 2022-10-02 paper.tex
Update citations

@@
 Orchestration software must communicate with each hardware component in the course of a scientific experiment.
-This can involve utilization of many protocols: NI DAQmx [CITE], SCPI [CITE], ModBus [CITE], PICam [CITE], Thorlabs APT [CITE], among many others.
+This can involve utilization of many protocols: NI DAQmx\cite{nidaqmx}, SCPI\cite{scpi}, ModBus\cite{modbus}, PICam\cite{picam}, Thorlabs APT\cite{thorlabs_apt}, among many others.
 The challenge of integrating all of these is a frustrating piece of the modern instrument development process.
@@
 These monolithic applications tend to be very inflexible and difficult to develop.
-In the last few years, several open source projects by-and-for small-scale experimentalists have grown in popularity [cite many].
+In the last few years, several open source projects by-and-for small-scale experimentalists have grown in popularity [CITE many].
 While this growth is encouraging, many of these are limited by their focus on particular types of hardware or particular experimental domains.
@@
 
-We have created a new protocol [XKCD] for scientific instrumentation, yaq.
+We have created a new protocol for scientific instrumentation, yaq.
 This protocol attempts to borrow the most important ideas from established projects while retaining simplicity appropriate for small research labs.
@@
 First barrier: Multiple incompatible protocols are used to communicate with each component of the system.
-A fully automated system must be able to use all of these protocols, a daunting task for scientists who do not specialize in software development\cite{}.
+A fully automated system must be able to use all of these protocols, a daunting task for scientists who do not specialize in software development.
 Second barrier: Certain specialty hardware have inconvenient interface requirements.
@@
 These small, targeted, programs are managed by the operating system and run in the background.
-It is conventional to call such programs ``daemons''\cite{}.
+It is conventional to call such programs ``daemons''\cite{Raymond_1996}.
 The middle computer is connected two two additional peripherals, and runs daemons for each.
@@
 This powerful architecture can scale down to single computers or up to many computers, including fully remote operator interfaces.
-Superficially, this network-based client-server design is identical to EPICS and TANGO [CITE].
+Superficially, this network-based client-server design is identical to EPICS and TANGO \cite{DalesioLR1991a}\cite{AGotz1999TANGOA}.
 As we will show, usage of standards and the creation of tooling makes this architecture accessible to instrument builders outside of large facilities.
@@
 Several general purpose \yaq{} clients exist.
-\texttt{yaqc}\cite{} is a lightweight Python client which is excellent for using in scripts or any other Python program.
+\texttt{yaqc}\cite{yaqc} is a lightweight Python client which is excellent for using in scripts or any other Python program.
 \texttt{yaqc} is totally generic and aims to support any conceivable \yaq{} daemon.
-\texttt{yaqc-qtpy}\cite{} is a graphical application which builds interactive controls based on traits for any conceivable \yaq{} protocol.
+\texttt{yaqc-qtpy}\cite{yaqc-qtpy} is a graphical application which builds interactive controls based on traits for any conceivable \yaq{} protocol.
 This is an invaluable tool which provides a ``free'' graphical user interface to any daemon.
-\texttt{yaqc-bluesky}\cite{} provides a bridge to the Bluesky ecosystem\cite{}.
+\texttt{yaqc-bluesky}\cite{yaqc-bluesky} provides a bridge to the Bluesky ecosystem\cite{AllanDanielB2019a}.
 Bluesky provides a powerful orchestration layer for conducting and recording data for a wide variety of experimental procedures.
-Similar translation layers could be built for a variety of orchestration layers such as PyMoDAQ\cite{}, Instrumental\cite{}, or TRSpectrometer\cite{}.
+Similar translation layers could be built for a variety of orchestration layers such as PyMoDAQ\cite{WeberSebastien2021a}, Instrumental\cite{Bogdanowicz2022}, or TRSpectrometer\cite{trspectrometer}.
 
@@
 We have created several tools to aide in daemon development.
-First, a Python library, \texttt{yaqd-core}\cite{}, which implements shared functionality.
-Simple interfaces, such as Brooks MFC\cite{}, can be implemented in as little as 50 lines of Python code.
-Second, \texttt{yaq-traits}\cite{} is a command line application which allows the description of messages provided by a \yaq{} protocol to be written in a human-readable fashion and translated into a more fully described machine readable format.
+First, a Python library, \texttt{yaqd-core}\cite{yaqd-core-python}, which implements shared functionality.
+Simple interfaces, such as Brooks MFC\cite{yaqd-brooks-mfc}, can be implemented in as little as 100 lines of Python code.
+Second, \texttt{yaq-traits}\cite{yaq-traits} is a command line application which allows the description of messages provided by a \yaq{} protocol to be written in a human-readable fashion and translated into a more fully described machine readable format.
 The format it generates is an important part of how \yaq{} protocols are self describing.
@@
 The lack of hardware interface code makes \yaq{} clients much simpler and easier to describe and maintain.
-Tools like \texttt{yaqd-fakes} allow clients to be tested and improved outside of their instrument, including the possibility of fully automated testing\cite{}.
+Tools like \texttt{yaqd-fakes} allow clients to be tested and improved outside of their instrument, including the possibility of fully automated testing\cite{yaqd-fakes}.
 Simple, script-based clients written using the expressiveness of Python can be read and understood in hours rather than weeks.

### 2022-10-02 paper.tex
cite many

@@
 These monolithic applications tend to be very inflexible and difficult to develop.
-In the last few years, several open source projects by-and-for small-scale experimentalists have grown in popularity [CITE many].
+In the last few years, several open source projects by-and-for small-scale experimentalists have grown in popularity\cite{WeberSebastien2021a,Bogdanowicz2022,trspectrometer,Koerner_2019,Campagnola_2014,pymeasure}.
 While this growth is encouraging, many of these are limited by their focus on particular types of hardware or particular experimental domains.

### 2022-10-03 paper.tex
Use alternate fig2, touch up references
Add Hardware-Control and pylablib

--- 2022-10-02 paper.tex
@@
 These monolithic applications tend to be very inflexible and difficult to develop.
-In the last few years, several open source projects by-and-for small-scale experimentalists have grown in popularity [CITE many].
+In the last few years, several open source projects by-and-for small-scale experimentalists have grown in popularity\cite{WeberSebastien2021a,Bogdanowicz2022,trspectrometer,Koerner_2019,Campagnola_2014,pymeasure,Giesbrecht_2022,pylablib}.
 While this growth is encouraging, many of these are limited by their focus on particular types of hardware or particular experimental domains.
@@
 \begin{figure}
-\includegraphics[width=\columnwidth]{figures/clients}
-  \caption{  \label{fig:foundation} A caption}
+\includegraphics[width=\columnwidth]{figures/client_spectrum},
+	\caption{  \label{fig:foundation} \yaq{} scales from simple scripts to integrations with featureful frameworks.}
 \end{figure}

### 2022-10-13 paper.tex
peter edits, protocol/standard/interface consistent usage

@@
 
-\title{The yaq Project: \\ A Protocol for Scientific Instrumentation}
+\title{The yaq Project: \\ A Standard for Scientific Instrumentation}
 \author{Kyle F. Sunden}
 \author{Daniel D. Kohler}
-\author{Peter C. Parilla}
 \author{Kent A. Meyer}
+\author{Peter L. Cruz Parilla}
 \author{John C. Wright}
@@
 Orchestration software must communicate with each hardware component in the course of a scientific experiment.
-This can involve utilization of many protocols: NI DAQmx\cite{nidaqmx}, SCPI\cite{scpi}, ModBus\cite{modbus}, PICam\cite{picam}, Thorlabs APT\cite{thorlabs_apt}, among many others.
+This can involve utilization of many interfaces: NI DAQmx\cite{nidaqmx}, SCPI\cite{scpi}, ModBus\cite{modbus}, PICam\cite{picam}, Thorlabs APT\cite{thorlabs_apt}, among many others.
 The challenge of integrating all of these is a frustrating piece of the modern instrument development process.
@@
 
-Some large user-facilities have addressed protocol complexity via the adoption unified protcols, such as EPICS \cite{DalesioLR1991a} or TANGO \cite{AGotz1999TANGOA}.
-The unified protocols define a standard network interface for any hardware component.
-Orchestration software can target this unified protocol for reading and writing hardware state.
-Small background services are written to translate the myriad component protocols into the standard EPICS IOCs and TANGO Devices.
-These programs are peformant, but require expert management to set up and provide descriptions via a separate server program.
+Some large user-facilities have addressed interface complexity via the adoption unified standards, such as EPICS \cite{DalesioLR1991a} or TANGO \cite{AGotz1999TANGOA}.
+The unified standards define a network interface for any hardware component.
+Orchestration software can target this unified standards for reading and writing hardware state.
+Small background services are written to translate the myriad component interfaces into the standard EPICS IOCs and TANGO Devices.
+These programs are performant, but require expert management to set up and provide descriptions via a separate server program.
 These projects are impressive at facility scale, but in our experience these do not scale well to single-investigator lab environments.
@@
 As smaller research labs have grown in experimental complexity, many individual labs have created domain-specific orchestration software.
-Most small custom research instrumentation continues to rely on monolithic software which has hardcorded protocol support for each particular connected device.
-These monolithic applications tend to be very inflexible and difficult to develop.
 In the last few years, several open source projects by-and-for small-scale experimentalists have grown in popularity\cite{WeberSebastien2021a,Bogdanowicz2022,trspectrometer,Koerner_2019,Campagnola_2014,pymeasure,Giesbrecht_2022,pylablib}.
 While this growth is encouraging, many of these are limited by their focus on particular types of hardware or particular experimental domains.
-To our knowledge, none of these projects take a strict protocol-based approach.
+Most small custom research instrumentation continues to rely on monolithic software which has hard-coded interface support for each particular connected device.
+These monolithic applications tend to be inflexible and difficult to develop.
+%To our knowledge, there are no existing open source projects that attempt to define a unified standard for hardware control of small scale experiments.
 
-We have created a new protocol for scientific instrumentation, yaq.
-This protocol attempts to borrow the most important ideas from established projects while retaining simplicity appropriate for small research labs.
-We have built this protocol to be self-describing, portable, and reusable wherever possible.
+We have created a new network-based standard for scientific instrumentation, \yaq{}.
+This standard borrows the most important ideas from established projects used by large user facilities while retaining simplicity appropriate for small research labs.
+We have built this standard to be self-describing, portable, and reusable wherever possible.
 Our primary goal has been to make an interface which simplifies orchestration software development as much as possible.
 
-Here we discuss the design of the yaq protocol in the context of four challenges facing instrument designers.
+Here, we discuss the design of the \yaq{} standard in the context of four challenges facing instrument designers.
 First, we discuss how particularly challenging hardware interfaces can become seemingly insurmountable barriers to software control.
-Next, we discuss how inflexible orchestration software can limit experimental flexibility.
+Next, we discuss how inflexible orchestration software can limit experimental creativity.
 Next, we focus on challenges that arise when enhancing or modifying existing instruments with new hardware.
-Finally, we discuss the heavy software maitenence burden that many instrument designers face.
-In each case we will highlight how the yaq project is designed to aliveate alleviate that challenge.
-Several case studies provide a view into the flexible ways that yaq can be applied to perform very different scientific experiments.
+Finally, we discuss the heavy software maintenance burden that many instrument designers face.
+In each case, we will highlight how the \yaq{} project is designed to alleviate that challenge.
+Several case studies provide a view into the flexible ways that \yaq{} can be applied to perform different scientific experiments.
+This document provides an overview of the concepts and motivations behind \yaq{}, but does not serve as a formal specification\cite{yaq_fyi}.
 
@@
 In this section we describe the architecture of the \yaq{} framework in light of three major barriers that we have encountered in scientific instrumentation development.
-First barrier: Multiple incompatible protocols are used to communicate with each component of the system.
-A fully automated system must be able to use all of these protocols, a daunting task for scientists who do not specialize in software development.
+First barrier: Multiple interfaces are used to communicate with each component of the system.
+A fully automated system must be able to use all of these interfaces, a daunting task for scientists who do not specialize in software development.
 Second barrier: Certain specialty hardware have inconvenient interface requirements.
-A specialty camera will only work with an ISA card and drivers for Windows XP.
-A data acquisition manufacturer provides an API that only works in Python 3.7.
-A graduate student wishes to drive some stepper motor using a Raspberry Pi.
+A camera will only work with an obsolete interface card and drivers for Windows XP.
+A data acquisition manufacturer provides an Application Programmer Interface (API) that only works in Python 3.7.
+A graduate student wishes to drive several stepper motors using a Raspberry Pi.
 Third barrier: Some hardware interfaces are blocking.
@@
 Figure \ref{fig:network} diagrams the \yaq{} architecture.
-Here we show three different computers connected via an ethernet network.
+Here, we show three different computers connected via an ethernet network.
 The top and bottom computer are connected to monitors for interactive use while the middle computer is only accessible via the network.
@@
 As an example, a client running on the bottom computer could communicate with the RS232 peripheral shown in green via the following path: client $\leftrightarrow$ network switch $\leftrightarrow$ top computer $\leftrightarrow$ daemon $\leftrightarrow$ hardware peripheral.
-This powerful architecture can scale down to single computers or up to many computers, including fully remote operator interfaces.
-Superficially, this network-based client-server design is identical to EPICS and TANGO \cite{DalesioLR1991a}\cite{AGotz1999TANGOA}.
+This powerful architecture can scale down to a single computer or up to many networked computers, including fully remote operator interfaces.
+This client-server architecture offers similar network capabilities to EPICS and TANGO.
 As we will show, usage of standards and the creation of tooling makes this architecture accessible to instrument builders outside of large facilities.
 
-In \yaq{}, communication between daemons and clients is performed over TCP/IP using a protocol called Apache Avro RPC \cite{AvroSpecification}.
+In \yaq{}, communication between daemons and clients is performed over TCP/IP using Apache Avro RPC \cite{AvroSpecification}.
 Avro provides an agreed upon standard for efficient serialization of data and method calls from a remote (client) process.
@@
 
-The first hardware interface barrier: Multiple incompatible protocols are used to communicate with each component of the system.
+The first hardware interface barrier: Multiple incompatible interfaces are used to communicate with each component of the system.
 \yaq{} provides a unified TCP/IP interface to all hardware peripherals based on the well-described Avro RPC protocol.
@@
 Since \yaq{} enables multiple machines, any hardware requirements can be addressed by putting a machine for that specific hardware on the network.
-For example, the Windows XP computer which has the interface card and drivers can be placed into a private network with only \yaq{} interface communication.
+For example, the Windows XP computer which has the obsolete interface card and drivers can be placed into a private network with only \yaq{} interface communication.
 Because each \yaq{} daemon is running in its own process, the software environment can be tailored to its needs.
@@
 The third hardware interface barrier: Some hardware interfaces are blocking.
-Each \yaq{} daemon runs as a separate process (possibly not even on the same machine) and so many of the hard concurrency problems with communicating with hardware are mitigated.
+Monolithic orchestration software often necessitates separate threads for each component hardware interface, a fragile pattern in which small mistakes become difficult to find and critical errors.
+In \yaq{}, each daemon is only responsible for a single hardware interface.
+%Each \yaq{} daemon runs as a separate process (possibly not even on the same machine) and thus many of the hard concurrency problems with communicating with hardware are mitigated.
 It is expected that each message call over the \yaq{} interface will return rapidly, ensuring that client applications are not blocked for extended periods of time.
@@
 In a \yaq{} context, orchestration code and graphical control interfaces are implemented as clients.
-These clients are automatically simpler because they only need to implement the \yaq{} interface and do not need to include the vast array of hardware-specific communication protocols.
+These clients are automatically simpler because they only need to implement the \yaq{} standard and do not need to include the vast array of hardware-specific communication interfaces.
 Beyond this, clients can use traits to interact with similar hardware identically.
@@
 These researchers are experimentalists who have limited background in software development.
-The Landis Group has written very basic Python scripts to orchestrate hardware for their reactor.
+The Landis Group has written basic Python scripts to orchestrate hardware for their reactor.
 These lightweight scripts can be extensively refactored by the experimentalists as the hardware and orchestration strategy changes dramatically during WiQK development.
@@
 A graduate student discovers a long standing hard-coded conversion factor that is incorrect years after implementation.
-Scientists resort to sourcing a replacement for an old, broken oscilloscope due to their reliance of their software on that particular interface; newer, cheaper options are readily available.
+Scientists resort to sourcing a replacement for an old, broken oscilloscope due to their reliance of their software on that particular interface; newer, cheaper oscilloscopes are readily available.
 A graduate student is forced to meticulously reverse engineer the LabView codebase that they inherited in order to understand the details of their experiment.
@@
 In \yaq{}, each component of an instrument can be developed and distributed separately.
-For example, two very different instruments might happen to use the same temperature sensor.
+For example, two different instruments might happen to use the same temperature sensor.
 Because the temperature sensor daemon is its own independent program, both instruments can benefit from the same daemon.
@@
 There are currently XX daemons in the yaq project supporting at least YY kinds of hardware.
-Because yaq is protocol based, anyone can design and publish new daemons extending our hardware support.
+Because \yaq{} is standards based, anyone can design and publish new daemons extending our hardware support.
 A living list of all daemons and supported hardware can be found on the yaq website.
@@
 \yaq{} attempts do automate daemon documentation as much as possible.
-Our website, https://yaq.fyi, automatically builds generated protocol reference pages for all known protocols.
+Our website, https://yaq.fyi, automatically builds generated reference pages for all known protocols.
 These pages are automatically updated when new versions are published.
@@
 
-The \yaq{} project defines a new general-purpose protocol for hardware control in the context of scientific instrumentation.
-This protocol has some of the powerful features of facility-scale protocols while remaining simple enough for feasable implementation and maintenance in small research labs.
+The \yaq{} project defines a new general-purpose standard for hardware control in the context of scientific instrumentation.
+This standard has some of the powerful features of facility-scale standards while remaining simple enough for feasible implementation and maintenance in small research labs.
 We have shown how this approach alleviates common problems through discussion and case studies.
-Designing around self-describing protocols is a productive approach that has great promise in scientific software deveopment.
+Designing around self-describing protocols is a productive approach that has great promise in scientific software development.
 

### 2022-10-18 paper.tex
rework sections 3-5

@@
 We believe that novel instrumentation development naturally and necessarily includes the creation of targeted software.
+Developing experimental software is an iterative process tied to the scientific goals of the instrument.
+Often experimentalists must apply their specialty scientific knowledge to develop this software.\cite{SegalJudith2005a}
 \yaq{} is architected to encourage better software development practices when creating such programs.
@@
 
-A \yaq{} client can scale in complexity from a small, lightweight script all the way to a complicated graphical program, as shown in Figure \ref{fig:foundation}.
-At the bottom, the daemons lay a solid foundation to build the variety of clients on top of.
-Client sophistication can be introduced naturally, as novel experiments are tested and refined.
-
 \begin{figure}
@@
 
-Several general purpose \yaq{} clients exist.
-\texttt{yaqc}\cite{yaqc} is a lightweight Python client which is excellent for using in scripts or any other Python program.
-\texttt{yaqc} is totally generic and aims to support any conceivable \yaq{} daemon.
+A \yaq{} client can scale in complexity from a small, lightweight script all the way to a complicated graphical program, as shown in Figure \ref{fig:foundation}.
+At the bottom, we represent the most lightweight interface to \yaq{}, Python scripting.
+\texttt{yaqc}\cite{yaqc} is a Python client which is excellent for using in scripts or any other Python program.
+The code shown creates three client objects which could be used directly through an interactive Python prompt or in a reusable script.
+Each client object provides Python methods for each Avro message specified by the associated protocol.
+
+In the middle, the same three hardware peripherals are represented in an interactive, graphical form.
 \texttt{yaqc-qtpy}\cite{yaqc-qtpy} is a graphical application which builds interactive controls based on traits for any conceivable \yaq{} protocol.
-This is an invaluable tool which provides a ``free'' graphical user interface to any daemon.
-\texttt{yaqc-bluesky}\cite{yaqc-bluesky} provides a bridge to the Bluesky ecosystem\cite{AllanDanielB2019a}.
+The self describing \yaq{} interface built using protocols and traits is used to provide interface elements for the most commonly used messages.
+This is an invaluable tool which provides a ``free'' graphical user interface (GUI) to any daemon.
+
+At the top, we show several features associated with the integration between \yaq{} and the Bluesky project\cite{AllanDanielB2019a}.
 Bluesky provides a powerful orchestration layer for conducting and recording data for a wide variety of experimental procedures.
+Bluesky is a framework which allows users to implement their own experimental procedures.
+Shown in Figure \ref{fig:foundation}, we see one such GUI for parameterizing a bluesky procedure and a representation of the resultant data.
+\texttt{yaqc-bluesky}\cite{yaqc-bluesky} provides a bridge to the Bluesky ecosystem.
 Similar translation layers could be built for a variety of orchestration layers such as PyMoDAQ\cite{WeberSebastien2021a}, Instrumental\cite{Bogdanowicz2022}, or TRSpectrometer\cite{trspectrometer}.
+
+All three types of clients represented in Figure \ref{fig:foundation} have been shown addressing the same three hardware peripherals.
+All types of clients can be used simultaneously to interact with the same instrument in different modes.
+Client sophistication can be introduced naturally, as novel experiments are tested and refined.
+Different clients can specialize for different requirements of a custom instrument
+For example, \texttt{yaqc-qtpy} can provide a quick interface for setting and viewing hardware positions while Bluesky can focus on experimental data acquisition.
 
@@
 
-The Wright Group at UW-Madison needs to orchestrate a large variety of hardware in multidimensional scans for their complex spectroscopy experiments \cite{MukamelShaul2000a, WrightJohnCurtis2011a}.
+The Wright Group at UW-Madison needs to orchestrate a large variety of hardware in multidimensional scans for their complex spectroscopy experiments \cite{Mukamel/Shaul2000a, WrightJohnCurtis2011a}.
 This need for exquisite hardware control has resulted in several prior attempts at ``home-built'' orchestration software \cite{CarlsonRogerJohn1988a, MeyerKentAlbert2004b, KainSchuyler2017a, ThompsonBlaiseJonathan2018a}.
@@
 
+There are currently XX daemons in the yaq project supporting at least YY kinds of hardware.
+Because \yaq{} is standards based, anyone can design and publish new daemons extending our hardware support.
+A living list of all daemons and supported hardware can be found on the yaq website.
+
 The Stahl Group at UW-Madison created a custom reactor which monitors gasses being produced or consumed in the reaction head-space.  \cite{SalazarChaseA2021a}
@@
 A graduate student is forced to meticulously reverse engineer the LabView codebase that they inherited in order to understand the details of their experiment.
-Technical debt grows especially fast in academic environments where graduate students are constantly being replaced.
+Software developers refer to the extra effort required to modify or fix large unmaintained codebases as ``technical debt''.\cite{}
+Technical debt grows especially fast in academic environments where graduate students are involved in projects for a limited time.
 
-We have found that the \yaq{} approach favors many small single-purpose applications above large monolithic ones.
+The \yaq{} approach favors many small single-purpose applications above large monolithic ones.
 For daemons, the purpose of each application is obvious and unambiguous. 
@@
 The lack of hardware interface code makes \yaq{} clients much simpler and easier to describe and maintain.
-Tools like \texttt{yaqd-fakes} allow clients to be tested and improved outside of their instrument, including the possibility of fully automated testing\cite{yaqd-fakes}.
+Tools like \texttt{yaqd-fakes}\cite{yaqd-fakes} allow clients to be tested and improved outside of their instrument, including the possibility of fully automated testing.
 Simple, script-based clients written using the expressiveness of Python can be read and understood in hours rather than weeks.
+Integrations with communities like Bluesky offer powerful features which are actively maintained across many institutions.
 
@@
 
-There are currently XX daemons in the yaq project supporting at least YY kinds of hardware.
-Because \yaq{} is standards based, anyone can design and publish new daemons extending our hardware support.
-A living list of all daemons and supported hardware can be found on the yaq website.
-
 Software documentation is famously difficult and thankless work.
@@
 
-yaq is also being used in several smaller ways throughout UW-Madison Chemistry.
+\yaq{} is open source software.
+Anyone can view, install, edit, and suggest changes to our growing collection of daemons and clients.
+ZZ individuals have contributed to the development of \yaq{}.
+
+\yaq{} is also being used in several smaller ways throughout UW-Madison Chemistry.
 Excitingly, all of these research groups are able easily benefit from each-other's daemon developments.

### 2022-10-20 paper.tex
histogram and abstract

@@
 \begin{abstract}
-An article usually includes an abstract, a concise summary of the work
-covered at length in the main body of the article. It is used for
-secondary publications and for information retrieval purposes.
+
+Modern instrumentation development often involves incorporation of many dissimilar hardware peripherals into a single unified instrument.
+Increasing hardware availability has brought greater instrument complexity to small research groups.
+The flexibility promised to instrument makers by large hardware availability stretches the capability of traditional monolithic orchestration software.
+In many cases, a lack of software flexibility leads creative researchers to feel frustrated and unable to perform the experiments thy envision.
+Highly modular hardware components enable extreme experimental flexibility.
+Herein, we describe \yaq{}, a software project defining a new standard way of communicating with diverse hardware peripherals.
+\yaq{} encourages a highly modular approach to experimental software development, which is well suited to address the experimental flexibility needs of such instruments.
+\yaq{} is architected to overcome hardware communication barriers that are insurmountable with typical experimental software.
+A large number of hardware peripherals are already supported with tooling available to create more.
+The \yaq{} standard enables collaboration between multiple research groups, increasing code quality while lowering development effort.
+
 \end{abstract}
@@
 As such, instrumentation software development time is in conflict with valuable data acquisition time.
-The hardware interfaces that these programs implement are typically implemented quickly and without regard to standardization with similar hardware.
+The hardware interfaces that these programs implement are typically made quickly and without regard to standardization with similar hardware.
 The orchestration routines are intimately tied to the particular hardware configuration of one instrument.
@@
 \texttt{yaqc-qtpy}\cite{yaqc-qtpy} is a graphical application which builds interactive controls based on traits for any conceivable \yaq{} protocol.
-The self describing \yaq{} interface built using protocols and traits is used to provide interface elements for the most commonly used messages.
+The self describing \yaq{} interface is used to provide graphical elements for the most commonly used messages.
 This is an invaluable tool which provides a ``free'' graphical user interface (GUI) to any daemon.
@@
 Bluesky is a framework which allows users to implement their own experimental procedures.
-Shown in Figure \ref{fig:foundation}, we see one such GUI for parameterizing a bluesky procedure and a representation of the resultant data.
+Shown in Figure \ref{fig:foundation}, we see one such GUI for parameterizing a Bluesky procedure and a representation of the resultant data.
 \texttt{yaqc-bluesky}\cite{yaqc-bluesky} provides a bridge to the Bluesky ecosystem.
@@
 
-The Wright Group at UW-Madison needs to orchestrate a large variety of hardware in multidimensional scans for their complex spectroscopy experiments \cite{Mukamel/Shaul2000a, WrightJohnCurtis2011a}.
+The Wright Group at UW-Madison needs to orchestrate a large variety of hardware in multidimensional scans for their complex spectroscopy experiments \cite{MukamelShaul2000a, WrightJohnCurtis2011a}.
 This need for exquisite hardware control has resulted in several prior attempts at ``home-built'' orchestration software \cite{CarlsonRogerJohn1988a, MeyerKentAlbert2004b, KainSchuyler2017a, ThompsonBlaiseJonathan2018a}.
@@
 
-In a \yaq{} context, new hardware can be incorporated into an instrument through the creation of a new daemon.
+In a \yaq{} context, new hardware can be incorporated into an instrument through the addition of a new daemon.
 The \yaq{} architecture automatically simplifies hardware interface development in several ways.
 First, because daemons are separate and portable programs, the development effort can be spread across the community of \yaq{} users.
+Often, researchers can download an existing daemon rather than writing a new one.
 Second, \yaq{} daemon development can be performed separated from the particulars of any individual client.
@@
 First, a Python library, \texttt{yaqd-core}\cite{yaqd-core-python}, which implements shared functionality.
-Simple interfaces, such as Brooks MFC\cite{yaqd-brooks-mfc}, can be implemented in as little as 100 lines of Python code.
 Second, \texttt{yaq-traits}\cite{yaq-traits} is a command line application which allows the description of messages provided by a \yaq{} protocol to be written in a human-readable fashion and translated into a more fully described machine readable format.
@@
 
-There are currently XX daemons in the yaq project supporting at least YY kinds of hardware.
+\begin{figure}
+\includegraphics[width=\columnwidth]{figures/lines_histogram},
+	\caption{  \label{fig:histogram} Histogram of the number of lines for each implemented daemon.
+	Some daemons are implemented in ways that share code, resulting in apparent line counts less than 10.
+	For example, the Thorlabs APT\cite{thorlabs_apt} motor implementation supports at least eight different daemons with each specifying only a handful of constants.
+	These are extreme examples which are not representative of most hardware interfaces, so we omit them here.
+	}
+\end{figure}
+
+Figure \ref{fig:histogram} shows the distribution of unique lines of Python code written to implement each of the daemons in our current ecosystem.
+While lines of code is an imperfect metric, we use it here to try and represent the amount of work required to create a daemon for a new hardware peripheral.
+Most interfaces, such as Brooks MFC\cite{yaqd-brooks-mfc}, have been implemented in as less than 100 lines of Python code.
+Even the most complicated daemons are implemented in about 300 lines.
+Implementing \yaq{} daemons using Python is enabled through our own tooling mentioned above and the large and growing ecosystem of hardware interface tooling libraries that Python now provides\cite{}.
+In our experience, the process of creating a new daemon involves about a day of work after communication with the hardware using Python is well understood.
+There are currently 70 daemons in the yaq project supporting at least 50 kinds of hardware, noting that some daemons support the same hardware and others are software only.
 Because \yaq{} is standards based, anyone can design and publish new daemons extending our hardware support.
-A living list of all daemons and supported hardware can be found on the yaq website.
+A living list of all daemons and supported hardware can be found on the \yaq{} website.\cite{}
 
@@
 This reactor incorporates a collection of sensitive pressure transducers and a single heating process value under computer control.
-yaq daemons are used to interface with each sensor and the heater controller.
+\yaq{} daemons are used to interface with each sensor and the heater controller.
 Recently, experimentalists have been attempting reactions involving smaller, slower, pressure changes.
@@
 Scientists resort to sourcing a replacement for an old, broken oscilloscope due to their reliance of their software on that particular interface; newer, cheaper oscilloscopes are readily available.
-A graduate student is forced to meticulously reverse engineer the LabView codebase that they inherited in order to understand the details of their experiment.
+A graduate student is forced to meticulously reverse engineer the LabVIEW codebase that they inherited in order to understand the details of their experiment.
 Software developers refer to the extra effort required to modify or fix large unmaintained codebases as ``technical debt''.\cite{}
@@
 Anyone can view, install, edit, and suggest changes to our growing collection of daemons and clients.
-ZZ individuals have contributed to the development of \yaq{}.
-
-\yaq{} is also being used in several smaller ways throughout UW-Madison Chemistry.
-Excitingly, all of these research groups are able easily benefit from each-other's daemon developments.
-This level of collaboration is new to us in the orchestration software space.
+Thirteen individuals have contributed code to the development of \yaq{}.
 

### 2022-10-21 paper.tex
morris comments

@@
-%  1)  latex  paper
-%  2)  bibtex paper
-%  3)  latex  paper
-%  4)  latex  paper
-%
 \documentclass[aip, amsmath, amssymb, reprint,]{revtex4-1}
@@
 The flexibility promised to instrument makers by large hardware availability stretches the capability of traditional monolithic orchestration software.
-In many cases, a lack of software flexibility leads creative researchers to feel frustrated and unable to perform the experiments thy envision.
+In many cases, a lack of software flexibility leads creative researchers to feel frustrated and unable to perform the experiments they envision.
 Highly modular hardware components enable extreme experimental flexibility.
-Herein, we describe \yaq{}, a software project defining a new standard way of communicating with diverse hardware peripherals.
+Herein we describe \yaq{}, a software project defining a new standard way of communicating with diverse hardware peripherals.
 \yaq{} encourages a highly modular approach to experimental software development, which is well suited to address the experimental flexibility needs of such instruments.
@@
 Weeks can be spent just integrating one new component into an existing project.
-Code reusability is typically poor, and technical debt grows quickly in academic and educational contexts.
+In small academic labs, these software interfaces are typically created by student researchers without software development experience.
+Student researchers rarely focus on software reusability, and a lack of maitenence and documentation can make such software more difficult to use as time goes on.
 Scientists may struggle to rapidly innovate on their experimental design when each hardware addition requires major software development.
 
-Some large user-facilities have addressed interface complexity via the adoption unified standards, such as EPICS \cite{DalesioLR1991a} or TANGO \cite{AGotz1999TANGOA}.
+Some large user-facilities have addressed interface complexity via the adoption of unified standards, such as EPICS \cite{DalesioLR1991a} or TANGO \cite{AGotz1999TANGOA}.
 The unified standards define a network interface for any hardware component.
-Orchestration software can target this unified standards for reading and writing hardware state.
+Orchestration software can target these unified standards for reading and writing hardware state.
 Small background services are written to translate the myriad component interfaces into the standard EPICS IOCs and TANGO Devices.
-These programs are performant, but require expert management to set up and provide descriptions via a separate server program.
-These projects are impressive at facility scale, but in our experience these do not scale well to single-investigator lab environments.
+These programs are performant, open source, and have huge libraries of exisiting hardware interface support, but require expert management to set up and provide descriptions via a separate server program.
+In our experience EPICS and TANGO do not scale well to single-investigator lab environments.
 
@@
 Several case studies provide a view into the flexible ways that \yaq{} can be applied to perform different scientific experiments.
-This document provides an overview of the concepts and motivations behind \yaq{}, but does not serve as a formal specification\cite{yaq_fyi}.
+This paper provides an overview of the concepts and motivations behind \yaq{}.
+Refer to the \yaq{} website for a formal specification of the \yaq{} standard.\cite{yaq_fyi}
 
@@
 
-Hardware-software interface design presents several technical challenges.
 In this section we describe the architecture of the \yaq{} framework in light of three major barriers that we have encountered in scientific instrumentation development.
@@
 As an example, a client running on the bottom computer could communicate with the RS232 peripheral shown in green via the following path: client $\leftrightarrow$ network switch $\leftrightarrow$ top computer $\leftrightarrow$ daemon $\leftrightarrow$ hardware peripheral.
-This powerful architecture can scale down to a single computer or up to many networked computers, including fully remote operator interfaces.
+This powerful architecture can be used on a single computer or used across many networked computers, including fully remote operator interfaces.
 This client-server architecture offers similar network capabilities to EPICS and TANGO.
@@
 \yaq{} provides a unified TCP/IP interface to all hardware peripherals based on the well-described Avro RPC protocol.
-In particular, the trait system was introduced in pursuit of our primary goal of easing the client development process.
+The trait system was introduced in pursuit of our primary goal of easing the client development process.
 Clients can trust that protocols that implement a given trait will behave in similar ways.
+The standardized \yaq{} interface hides all interface complexity away from researchers and client-side scientific code.
 
@@
 The third hardware interface barrier: Some hardware interfaces are blocking.
-Monolithic orchestration software often necessitates separate threads for each component hardware interface, a fragile pattern in which small mistakes become difficult to find and critical errors.
+Monolithic orchestration software often necessitates separate threads for each component hardware interface, a fragile pattern in which small mistakes become both critical and elusive errors.
 In \yaq{}, each daemon is only responsible for a single hardware interface.
@@
 Bluesky provides a powerful orchestration layer for conducting and recording data for a wide variety of experimental procedures.
-Bluesky is a framework which allows users to implement their own experimental procedures.
 Shown in Figure \ref{fig:foundation}, we see one such GUI for parameterizing a Bluesky procedure and a representation of the resultant data.
@@
 The Wright Group uses simulated hardware to enable client development away from the active laboratory computers.
-The same client is used on four laser systems, each with different complements of hardware.
-Moving forward, the Wright Group hopes to spend less energy developing control software and more energy developing creative spectroscopy experiments.
+Remote client development on viritual hardware has allowed client design to proceed without delaying ongling experiments.
+Clients developed by the Wright Group have proved flexible enough to be used on four laser systems, each with different complements of hardware.
+Moving forward, the Wright Group will spend less energy developing control software and more energy developing creative spectroscopy experiments.
 
@@
 In a \yaq{} context, new hardware can be incorporated into an instrument through the addition of a new daemon.
-The \yaq{} architecture automatically simplifies hardware interface development in several ways.
+The \yaq{} architecture simplifies hardware interface development in several ways.
 First, because daemons are separate and portable programs, the development effort can be spread across the community of \yaq{} users.
@@
 Third, when developing trait-compliant protocols, it becomes easy to design and fully test your hardware interface.
-In this instance, you even can use \texttt{yaqc-qtpy} to provide a graphical program to interact with your hardware immediately.
+Traits are unambiguous and well-described, making an obvious target for development.
+Tooling exists to verify full trait compliance, for example you can use \texttt{yaqc-qtpy} to provide a graphical program to interact with your hardware immediately.
 Finally, as discussed in Section II, \yaq{} gives you options to design using remote hardware or unusual interfaces when necessary.
@@
 Figure \ref{fig:histogram} shows the distribution of unique lines of Python code written to implement each of the daemons in our current ecosystem.
-While lines of code is an imperfect metric, we use it here to try and represent the amount of work required to create a daemon for a new hardware peripheral.
+While lines of code is an imperfect metric, we use it here to represent the amount of work required to create a daemon for a new hardware peripheral.
 Most interfaces, such as Brooks MFC\cite{yaqd-brooks-mfc}, have been implemented in as less than 100 lines of Python code.
@@
 As a result, a new digitizer has been purchased.
-This new converter will be incorporated into the existing reactor without modifying the existing graphical user interface and data recording program, minimizing downtime.
+This new digitizer will be incorporated into the existing reactor without modifying the existing graphical user interface and data recording program, minimizing downtime.
 
@@
 A graduate student discovers a long standing hard-coded conversion factor that is incorrect years after implementation.
-Scientists resort to sourcing a replacement for an old, broken oscilloscope due to their reliance of their software on that particular interface; newer, cheaper oscilloscopes are readily available.
+Scientists resort to sourcing a replacement for an old, broken oscilloscope due to their reliance of their software on that particular interface while newer, cheaper oscilloscopes are readily available.
 A graduate student is forced to meticulously reverse engineer the LabVIEW codebase that they inherited in order to understand the details of their experiment.
@@
 Because the temperature sensor daemon is its own independent program, both instruments can benefit from the same daemon.
-As \yaq{} grows, the "ecosystem" of existing daemons means that future instruments become easier and easier to develop.
-Growing this ecosystem is a collaborative effort where many \yaq{} users create portable daemons that they need and share them for the community to use and improve.
+The growing ``ecosystem'' of \yaq{} daemons make future instruments easier and easier to develop.
+Growing this ecosystem is a collaborative effort where many \yaq{} users create portable daemons that they need and share them with the community where they will be used and improved.
 

### 2022-10-23 paper.tex
small manuscript edits and comments

--- 2022-10-20 paper.tex
@@
 %
-\documentclass[aip, amsmath, amssymb, reprint,]{revtex4-1}
+\documentclass[aip, amsmath, amssymb, reprint,]{revtex4-1} % consider using RevTex4.2
 
@@
 Modern instrumentation development often involves incorporation of many dissimilar hardware peripherals into a single unified instrument.
-Increasing hardware availability has brought greater instrument complexity to small research groups.
-The flexibility promised to instrument makers by large hardware availability stretches the capability of traditional monolithic orchestration software.
-In many cases, a lack of software flexibility leads creative researchers to feel frustrated and unable to perform the experiments thy envision.
-Highly modular hardware components enable extreme experimental flexibility.
-Herein, we describe \yaq{}, a software project defining a new standard way of communicating with diverse hardware peripherals.
-\yaq{} encourages a highly modular approach to experimental software development, which is well suited to address the experimental flexibility needs of such instruments.
+Increasing modular, commercial hardware availability has brought greater instrument complexity to small research groups.
+The flexibility promised to instrument makers by large hardware availability stretches the capability of traditional, monolithic orchestration software.
+In many cases, a lack of software flexibility leads creative researchers to feel frustrated and unable to perform the experiments they envision even when highly modular hardware components enable extreme experimental flexibility.
+Herein, we describe \yaq{}, a software project defining a new standardized way of communicating with diverse hardware peripherals.
+\yaq{} encourages a highly modular approach to experimental software development, which is well suited to address the experimental flexibility needs of complex instruments.
 \yaq{} is architected to overcome hardware communication barriers that are insurmountable with typical experimental software.
 A large number of hardware peripherals are already supported with tooling available to create more.
-The \yaq{} standard enables collaboration between multiple research groups, increasing code quality while lowering development effort.
+The \yaq{} standard enables collaboration among multiple research groups, increasing code quality while lowering development effort.
 
@@
 Instrumentation development is a key part of the scientific enterprise.
-Novel instruments are typically constructed of many individual components, purchased and home-built.
+Novel instruments are typically constructed of many individual components that are both purchased and home-built.
 Orchestration software must communicate with each hardware component in the course of a scientific experiment.
 This can involve utilization of many interfaces: NI DAQmx\cite{nidaqmx}, SCPI\cite{scpi}, ModBus\cite{modbus}, PICam\cite{picam}, Thorlabs APT\cite{thorlabs_apt}, among many others.
-The challenge of integrating all of these is a frustrating piece of the modern instrument development process.
+The challenge of integrating all of these interfaces is a frustrating piece of the modern instrument development process.
 Weeks can be spent just integrating one new component into an existing project.
@@
 
-Some large user-facilities have addressed interface complexity via the adoption unified standards, such as EPICS \cite{DalesioLR1991a} or TANGO \cite{AGotz1999TANGOA}.
+Some large user-facilities have addressed interface complexity via the adoption of unified standards, such as EPICS \cite{DalesioLR1991a} or TANGO \cite{AGotz1999TANGOA}.
 The unified standards define a network interface for any hardware component.
@@
 
-We have created a new network-based standard for scientific instrumentation, \yaq{}.
-This standard borrows the most important ideas from established projects used by large user facilities while retaining simplicity appropriate for small research labs.
+We have created a new network-based communication standard for scientific instrumentation, \yaq{}.
+This standard borrows the most important ideas from established projects used by large user facilities while retaining the simplicity appropriate for small research labs.
 We have built this standard to be self-describing, portable, and reusable wherever possible.
@@
 Next, we discuss how inflexible orchestration software can limit experimental creativity.
-Next, we focus on challenges that arise when enhancing or modifying existing instruments with new hardware.
+Then, we focus on challenges that arise when enhancing or modifying existing instruments with new hardware.
 Finally, we discuss the heavy software maintenance burden that many instrument designers face.
@@
 Several case studies provide a view into the flexible ways that \yaq{} can be applied to perform different scientific experiments.
-This document provides an overview of the concepts and motivations behind \yaq{}, but does not serve as a formal specification\cite{yaq_fyi}.
+This document provides an overview of the concepts and motivations behind \yaq{}, but does not serve as a formal specification, which is hosted at \cite{yaq_fyi}. % citation does not compile
 
@@
 It is conventional to call such programs ``daemons''\cite{Raymond_1996}.
-The middle computer is connected two two additional peripherals, and runs daemons for each.
+The middle computer is connected to two additional peripherals, and runs daemons for each.
 Besides communicating with the hardware peripheral, each daemon can communicate with other programs, ``clients'',  through the network.
@@
 The first hardware interface barrier: Multiple incompatible interfaces are used to communicate with each component of the system.
+% consider making this sentence (and the next paragraph starts) subsection headings
 \yaq{} provides a unified TCP/IP interface to all hardware peripherals based on the well-described Avro RPC protocol.
@@
 The third hardware interface barrier: Some hardware interfaces are blocking.
-Monolithic orchestration software often necessitates separate threads for each component hardware interface, a fragile pattern in which small mistakes become difficult to find and critical errors.
+Monolithic orchestration software often necessitates separate threads for each component hardware interface, a fragile pattern in which small mistakes become difficult to find and can lead to critical errors.
 In \yaq{}, each daemon is only responsible for a single hardware interface.
@@
 Developing experimental software is an iterative process tied to the scientific goals of the instrument.
-Often experimentalists must apply their specialty scientific knowledge to develop this software.\cite{SegalJudith2005a}
+Often experimentalists must apply their specialty scientific knowledge to develop this software.\cite{SegalJudith2005a} % citation does not compile
 \yaq{} is architected to encourage better software development practices when creating such programs.
@@
 	These are extreme examples which are not representative of most hardware interfaces, so we omit them here.
+% it's not clear to me why these instances are not included (and scaled down???) It is a _feature_ not a bug that one can implement motor support with tiny amounts of code.
 	}
@@
 While lines of code is an imperfect metric, we use it here to try and represent the amount of work required to create a daemon for a new hardware peripheral.
-Most interfaces, such as Brooks MFC\cite{yaqd-brooks-mfc}, have been implemented in as less than 100 lines of Python code.
+Most interfaces, such as Brooks MFC\cite{yaqd-brooks-mfc}, have been implemented in fewer than 100 lines of Python code.
 Even the most complicated daemons are implemented in about 300 lines.
@@
 Software documentation is famously difficult and thankless work.
-\yaq{} attempts do automate daemon documentation as much as possible.
+\yaq{} attempts to automate daemon documentation as much as possible.
 Our website, https://yaq.fyi, automatically builds generated reference pages for all known protocols.
@@
 \section*{References}
-
+% all yaq,yaqc,and yaqd citations have an unfilled "()" in them... 
 \nocite{*}

### 2022-10-24 paper.tex
Update paper.tex
added MPI paragraph no figure as of yet

--- 2022-10-21 paper.tex
@@
 This can involve utilization of many interfaces: NI DAQmx\cite{nidaqmx}, SCPI\cite{scpi}, ModBus\cite{modbus}, PICam\cite{picam}, Thorlabs APT\cite{thorlabs_apt}, among many others.
-The challenge of integrating all of these is a frustrating piece of the modern instrument development process.
+The process of integrating these is an important but time-consuming step of modern instrument and systems development.
 Weeks can be spent just integrating one new component into an existing project.
@@
 
-We have created a new network-based standard for scientific instrumentation, \yaq{}.
-This standard borrows the most important ideas from established projects used by large user facilities while retaining simplicity appropriate for small research labs.
+We have created a novel network-based standard for scientific instrumentation, called \yaq{}.  This standard borrows the most important ideas from
+ established projects used by large user facilities while retaining simplicity appropriate for small research labs.
 We have built this standard to be self-describing, portable, and reusable wherever possible.
@@
 Clients can trust that protocols that implement a given trait will behave in similar ways.
-The standardized \yaq{} interface hides all interface complexity away from researchers and client-side scientific code.
+The standardized \yaq{} interface presents the same set of interactions for client-side scientific code, simplifying the experience of using hardware.
 
-The second hardware interface barrier: Certain specialty hardware have inconvenient interface requirements.
+The second hardware interface barrier: Reduction of the interfacing code to an easier-to-manage subset of that available.  Here, we implement an approach useful for
+systems integration in the sciences which we call a Minimal Programming Interface (MPI).   The MPI is essentially a reduced set of functions found in the hardware API.   The MPI 
+ is defined as the minimal set of API functions for a given hardware component concluded by the researchers as necessary for the fuller system development. 
+ The value of this MPI is that it can be formed by wrapping and/or converting the API from its native language into a language known to be transparent to researchers, such as the programming
+ language Python.  The \yaq{} traits feature acts as a tool in exposing the MPI from the base APIs often supplied by the hardware manufacturer, as the traits
+ automatically define the primary set of key functions, and any additionally required functions are added by the researcher on top of the primary set.
+
+The third hardware interface barrier: Certain specialty hardware have inconvenient interface requirements.
 Since \yaq{} enables multiple machines, any hardware requirements can be addressed by putting a machine for that specific hardware on the network.
@@
 
-The third hardware interface barrier: Some hardware interfaces are blocking.
+The fourth hardware interface barrier: Some hardware interfaces are blocking.
 Monolithic orchestration software often necessitates separate threads for each component hardware interface, a fragile pattern in which small mistakes become both critical and elusive errors.

### 2022-10-26 paper.tex
fixed sentences at beginning of section to yield proper barrier designations

@@
 A fully automated system must be able to use all of these interfaces, a daunting task for scientists who do not specialize in software development.
-Second barrier: Certain specialty hardware have inconvenient interface requirements.
+Second barrier: Coding interfaces adds to a maintenance overhead seldom manageable by researchers.   Updates to the hardware and operating systems often
+require maintenance updates to interfacing code that is not maintainable as researchers are not usually delegated to be strictly software maintainers.
+Third barrier: Certain specialty hardware have inconvenient interface requirements.
 A camera will only work with an obsolete interface card and drivers for Windows XP.
@@
 A graduate student wishes to drive several stepper motors using a Raspberry Pi.
-Third barrier: Some hardware interfaces are blocking.
+Fourth barrier: Some hardware interfaces are blocking.
 A graphical user interface stalls while waiting for a camera to collect data.
@@
 
-The second hardware interface barrier: Reduction of the interfacing code to an easier-to-manage subset of that available.  Here, we implement an approach useful for
+The second hardware interface barrier:  Coding interfaces adds to a maintenance overhead seldom manageable by researchers.  Here, we implement an approach useful for
 systems integration in the sciences which we call a Minimal Programming Interface (MPI).   The MPI is essentially a reduced set of functions found in the hardware API.   The MPI 

### 2022-10-26 paper.tex

--- 2022-10-24 paper.tex
@@
-\documentclass[aip, amsmath, amssymb, reprint,]{revtex4-1}
+\documentclass[aip, amsmath, amssymb, reprint,]{revtex4-1} % consider using RevTex4.2
 
@@
 Modern instrumentation development often involves incorporation of many dissimilar hardware peripherals into a single unified instrument.
-Increasing hardware availability has brought greater instrument complexity to small research groups.
-The flexibility promised to instrument makers by large hardware availability stretches the capability of traditional monolithic orchestration software.
-In many cases, a lack of software flexibility leads creative researchers to feel frustrated and unable to perform the experiments they envision.
-Highly modular hardware components enable extreme experimental flexibility.
-Herein we describe \yaq{}, a software project defining a new standard way of communicating with diverse hardware peripherals.
-\yaq{} encourages a highly modular approach to experimental software development, which is well suited to address the experimental flexibility needs of such instruments.
+Increasing modular, commercial hardware availability has brought greater instrument complexity to small research groups.
+The flexibility promised to instrument makers by large hardware availability stretches the capability of traditional, monolithic orchestration software.
+In many cases, a lack of software flexibility leads creative researchers to feel frustrated and unable to perform the experiments they envision even when highly modular hardware components enable extreme experimental flexibility.
+Herein, we describe \yaq{}, a software project defining a new standardized way of communicating with diverse hardware peripherals.
+\yaq{} encourages a highly modular approach to experimental software development, which is well suited to address the experimental flexibility needs of complex in
 \yaq{} is architected to overcome hardware communication barriers that are insurmountable with typical experimental software.
 A large number of hardware peripherals are already supported with tooling available to create more.
-The \yaq{} standard enables collaboration between multiple research groups, increasing code quality while lowering development effort.
+The \yaq{} standard enables collaboration among multiple research groups, increasing code quality while lowering development effort.
 
@@
 Instrumentation development is a key part of the scientific enterprise.
-Novel instruments are typically constructed of many individual components, purchased and home-built.
+Novel instruments are typically constructed of many individual components that are both purchased and home-built.
 Orchestration software must communicate with each hardware component in the course of a scientific experiment.
 This can involve utilization of many interfaces: NI DAQmx\cite{nidaqmx}, SCPI\cite{scpi}, ModBus\cite{modbus}, PICam\cite{picam}, Thorlabs APT\cite{thorlabs_apt}, among many others.
-The process of integrating these is an important but time-consuming step of modern instrument and systems development.
+The challenge of integrating all of these interfaces is a frustrating piece of the modern instrument development process.
 Weeks can be spent just integrating one new component into an existing project.
@@
 
-We have created a novel network-based standard for scientific instrumentation, called \yaq{}.  This standard borrows the most important ideas from
- established projects used by large user facilities while retaining simplicity appropriate for small research labs.
+We have created a new network-based communication standard for scientific instrumentation, \yaq{}.
+This standard borrows the most important ideas from established projects used by large user facilities while retaining the simplicity appropriate for small research labs.
 We have built this standard to be self-describing, portable, and reusable wherever possible.
@@
 
-The second hardware interface barrier: Reduction of the interfacing code to an easier-to-manage subset of that available.  Here, we implement an approach useful for
-systems integration in the sciences which we call a Minimal Programming Interface (MPI).   The MPI is essentially a reduced set of functions found in the hardware API.   The MPI 
- is defined as the minimal set of API functions for a given hardware component concluded by the researchers as necessary for the fuller system development. 
- The value of this MPI is that it can be formed by wrapping and/or converting the API from its native language into a language known to be transparent to researchers, such as the programming
- language Python.  The \yaq{} traits feature acts as a tool in exposing the MPI from the base APIs often supplied by the hardware manufacturer, as the traits
- automatically define the primary set of key functions, and any additionally required functions are added by the researcher on top of the primary set.
-
-The third hardware interface barrier: Certain specialty hardware have inconvenient interface requirements.
+The second hardware interface barrier: Certain specialty hardware have inconvenient interface requirements.
 Since \yaq{} enables multiple machines, any hardware requirements can be addressed by putting a machine for that specific hardware on the network.
@@
 
-The fourth hardware interface barrier: Some hardware interfaces are blocking.
+The third hardware interface barrier: Some hardware interfaces are blocking.
 Monolithic orchestration software often necessitates separate threads for each component hardware interface, a fragile pattern in which small mistakes become both critical and elusive errors.
@@
 Figure \ref{fig:histogram} shows the distribution of unique lines of Python code written to implement each of the daemons in our current ecosystem.
+
 While lines of code is an imperfect metric, we use it here to represent the amount of work required to create a daemon for a new hardware peripheral.
-Most interfaces, such as Brooks MFC\cite{yaqd-brooks-mfc}, have been implemented in as less than 100 lines of Python code.
+Most interfaces, such as Brooks MFC\cite{yaqd-brooks-mfc}, have been implemented in fewer than 100 lines of Python code.
 Even the most complicated daemons are implemented in about 300 lines.

### 2022-10-31 paper.tex
tweaked text for errors and added the MPI acronym, which may need to be named a different acronym

--- 2022-10-24 paper.tex
@@
 In small academic labs, these software interfaces are typically created by student researchers without software development experience.
-Student researchers rarely focus on software reusability, and a lack of maitenence and documentation can make such software more difficult to use as time goes on.
+Student researchers rarely focus on software reusability, and a lack of maintenance and documentation can make such software more difficult to use over time.
 Scientists may struggle to rapidly innovate on their experimental design when each hardware addition requires major software development.
 
-Some large user-facilities have addressed interface complexity via the adoption of unified standards, such as EPICS \cite{DalesioLR1991a} or TANGO \cite{AGotz1999TANGOA}.
+Some large user-facilities have addressed interface complexity via the adoption of unified standards, such as EPICS \cite{DalesioLR1991a} and/or TANGO \cite{AGotz1999TANGOA}.
 The unified standards define a network interface for any hardware component.
@@
 
-We have created a novel network-based standard for scientific instrumentation, called \yaq{}.  This standard borrows the most important ideas from
+In order to address these issues, we have created a novel network-based standard for scientific instrumentation, called \yaq{}.  This standard borrows the most important ideas from
  established projects used by large user facilities while retaining simplicity appropriate for small research labs.
@@
 This paper provides an overview of the concepts and motivations behind \yaq{}.
-Refer to the \yaq{} website for a formal specification of the \yaq{} standard.\cite{yaq_fyi}
+The reader may also refer to the \yaq{} website for a formal specification of the \yaq{} standard.\cite{yaq_fyi}
 
@@
 In this section we describe the architecture of the \yaq{} framework in light of three major barriers that we have encountered in scientific instrumentation development.
-First barrier: Multiple interfaces are used to communicate with each component of the system.
+\textit{First barrier: Multiple interfaces are used to communicate with each component of the system.}
 A fully automated system must be able to use all of these interfaces, a daunting task for scientists who do not specialize in software development.
-Second barrier: Certain specialty hardware have inconvenient interface requirements.
+\textit{Second barrier: Coding interfaces adds to a maintenance overhead seldom manageable by researchers.}   Updates to the hardware and operating systems often
+require maintenance updates to interfacing code that is not maintainable as researchers are not usually delegated to be strictly software maintainers.
+\texit{Third barrier: Certain specialty hardware have inconvenient interface requirements.}
 A camera will only work with an obsolete interface card and drivers for Windows XP.
@@
 A graduate student wishes to drive several stepper motors using a Raspberry Pi.
-Third barrier: Some hardware interfaces are blocking.
+\texit{Fourth barrier: Some hardware interfaces are blocking.}
 A graphical user interface stalls while waiting for a camera to collect data.
@@
 \begin{figure}
-	\includegraphics[width=\columnwidth]{figures/network}
-\caption{\label{fig:network} Networking diagram}
+	\includegraphics[width=\columnwidth]{figures/network2}
+\caption{\label{fig:network2} Networking diagram}
 \end{figure}
@@
 As an example, a client running on the bottom computer could communicate with the RS232 peripheral shown in green via the following path: client $\leftrightarrow$ network switch $\leftrightarrow$ top computer $\leftrightarrow$ daemon $\leftrightarrow$ hardware peripheral.
-This powerful architecture can be used on a single computer or used across many networked computers, including fully remote operator interfaces.
-This client-server architecture offers similar network capabilities to EPICS and TANGO.
+This architecture can be used on a single computer or used across many networked computers, including fully remote operator interfaces.
+This client-server arrangement offers similar network capabilities to EPICS and TANGO.
 As we will show, usage of standards and the creation of tooling makes this architecture accessible to instrument builders outside of large facilities.
@@
 Each individual communication between client and daemon involves one message being requested by the client and the response returned from the daemon.
-Each daemon supports a collection of messages for its unique functionality, called the ``protocol''.
+Each daemon supports a collection of messages for its own unique functionality, called the ``protocol''.
 
@@
 
-The first hardware interface barrier: Multiple incompatible interfaces are used to communicate with each component of the system.
+\textit{The first hardware interface barrier: Multiple incompatible interfaces are used to communicate with each component of the system.}
 \yaq{} provides a unified TCP/IP interface to all hardware peripherals based on the well-described Avro RPC protocol.
@@
 
-The second hardware interface barrier: Reduction of the interfacing code to an easier-to-manage subset of that available.  Here, we implement an approach useful for
+\textit{The second hardware interface barrier:  Coding interfaces add to a maintenance overhead seldom manageable by researchers.}  Here, we implement an approach useful for
 systems integration in the sciences which we call a Minimal Programming Interface (MPI).   The MPI is essentially a reduced set of functions found in the hardware API.   The MPI 
@@
 
-The third hardware interface barrier: Certain specialty hardware have inconvenient interface requirements.
+\textit{The third hardware interface barrier: Certain specialty hardware have inconvenient interface requirements.}
 Since \yaq{} enables multiple machines, any hardware requirements can be addressed by putting a machine for that specific hardware on the network.
@@
 
-The fourth hardware interface barrier: Some hardware interfaces are blocking.
+\textit{The fourth hardware interface barrier: Some hardware interfaces are blocking.}
 Monolithic orchestration software often necessitates separate threads for each component hardware interface, a fragile pattern in which small mistakes become both critical and elusive errors.
@@
 In these instances, the initial message simply starts the action and returns, with a separate message provided to retrieve results when they exist.
-In order to know how long to wait, the ``is-daemon`` trait provides a message called ``is\_busy'' which should return "true" while the long running action is not complete, and "false" once it is finished.
+In order to know how long to wait, the ``is-daemon`` trait provides a message called ``is\_busy'' which should return \texttt{True} while the long running action is not complete, and \textt{False} once it is finished.
 Additionally, multiple clients can communicate with the same daemon simultaneously.
@@
 In our view, software inflexibility is a natural consequence of the typical software development practices used by custom instrument builders.
-Instrumental software is often built as one monolithic program that does everything from providing a graphical interface, through hardware interfacing, and writing data files.
+Instrumental software is often built as one monolithic program that does everything from providing a graphical interface, to hardware interfacing, and to writing data files.
 Such software is typically impossible to debug without access to real hardware, often requiring all of the hardware to be available to simply start the program.
@@
 Developing experimental software is an iterative process tied to the scientific goals of the instrument.
-Often experimentalists must apply their specialty scientific knowledge to develop this software.\cite{SegalJudith2005a}
+Often experimentalists must apply their specialized scientific knowledge to develop this software.\cite{SegalJudith2005a}
 \yaq{} is architected to encourage better software development practices when creating such programs.
@@
 The self describing \yaq{} interface is used to provide graphical elements for the most commonly used messages.
-This is an invaluable tool which provides a ``free'' graphical user interface (GUI) to any daemon.
+This is an invaluable tool as it provides a ``free'' graphical user interface (GUI) to any daemon.
 
@@
 Different clients can specialize for different requirements of a custom instrument
-For example, \texttt{yaqc-qtpy} can provide a quick interface for setting and viewing hardware positions while Bluesky can focus on experimental data acquisition.
+For example, \texttt{yaqc-qtpy} can provide a quick interface for setting and viewing hardware positions while Bluesky can focus on the actual data acquisition itself
+and the necessary planning on how the acquisition should be performed.
 
@@
 Clients developed by the Wright Group have proved flexible enough to be used on four laser systems, each with different complements of hardware.
-Moving forward, the Wright Group will spend less energy developing control software and more energy developing creative spectroscopy experiments.
+We expect that the Wright Group will spend less energy developing control software and more energy developing creative spectroscopy experiments.
 
@@
 There are currently 70 daemons in the yaq project supporting at least 50 kinds of hardware, noting that some daemons support the same hardware and others are software only.
-Because \yaq{} is standards based, anyone can design and publish new daemons extending our hardware support.
-A living list of all daemons and supported hardware can be found on the \yaq{} website.\cite{}
+Because \yaq{} is standards-based, anyone can design and publish new daemons extending our hardware support.  The daemons are constructed under MPI methodologies and are more freely
+maintainable and adaptable than the original APIs.  A living list of all daemons and supported hardware can be found on the \yaq{} website.\cite{}
 
@@
 As a result, a new digitizer has been purchased.
-This new digitizer will be incorporated into the existing reactor without modifying the existing graphical user interface and data recording program, minimizing downtime.
+This new digitizer will be incorporated into the existing reactor without modifying the existing graphical user interface and data recording program, and so downtime will be minimized.
 
@@
 
-The \yaq{} approach favors many small single-purpose applications above large monolithic ones.
+The \yaq{} approach attempts to minimize techincal debt by favoring many smaller, more easily adjustable single-purpose applications over larger, monolithic ones.
 For daemons, the purpose of each application is obvious and unambiguous.

### 2022-10-31 paper.tex
fix textit

@@
 require maintenance updates to interfacing code that is not maintainable as researchers are not usually delegated to be strictly software maintainers.
-\texit{Third barrier: Certain specialty hardware have inconvenient interface requirements.}
+\textit{Third barrier: Certain specialty hardware have inconvenient interface requirements.}
 A camera will only work with an obsolete interface card and drivers for Windows XP.
@@
 A graduate student wishes to drive several stepper motors using a Raspberry Pi.
-\texit{Fourth barrier: Some hardware interfaces are blocking.}
+\textit{Fourth barrier: Some hardware interfaces are blocking.}
 A graphical user interface stalls while waiting for a camera to collect data.
@@
 	\includegraphics[width=\columnwidth]{figures/network2}
-\caption{\label{fig:network2} Networking diagram}
+\caption{\label{fig:network} Networking diagram}
 \end{figure}

### 2022-11-01 paper.tex
Variety of smaller fixes to spelling and grammar/word choice (largely) suggested by JCW and DDK

--- 2022-10-26 paper.tex
@@
-\documentclass[aip, amsmath, amssymb, reprint,]{revtex4-1} % consider using RevTex4.2
+\documentclass[aip, amsmath, amssymb, reprint,]{revtex4-1}
 
@@
 Modern instrumentation development often involves incorporation of many dissimilar hardware peripherals into a single unified instrument.
-Increasing modular, commercial hardware availability has brought greater instrument complexity to small research groups.
-The flexibility promised to instrument makers by large hardware availability stretches the capability of traditional, monolithic orchestration software.
-In many cases, a lack of software flexibility leads creative researchers to feel frustrated and unable to perform the experiments they envision even when highly modular hardware components enable extreme experimental flexibility.
-Herein, we describe \yaq{}, a software project defining a new standardized way of communicating with diverse hardware peripherals.
-\yaq{} encourages a highly modular approach to experimental software development, which is well suited to address the experimental flexibility needs of complex in
+Increasing hardware availability has brought greater instrument complexity to small research groups.
+The flexibility promised to instrument makers by large hardware availability stretches the capability of traditional monolithic orchestration software.
+In many cases, a lack of software flexibility leads creative researchers to feel frustrated and unable to perform the experiments they envision.
+Highly modular hardware components enable extreme experimental flexibility.
+Herein we describe \yaq{}, a software project defining a new standard way of communicating with diverse hardware peripherals.
+\yaq{} encourages a highly modular approach to experimental software development, which is well suited to address the experimental flexibility needs of such instruments.
 \yaq{} is architected to overcome hardware communication barriers that are insurmountable with typical experimental software.
 A large number of hardware peripherals are already supported with tooling available to create more.
-The \yaq{} standard enables collaboration among multiple research groups, increasing code quality while lowering development effort.
+The \yaq{} standard enables collaboration between multiple research groups, increasing code quality while lowering development effort.
 
@@
 Instrumentation development is a key part of the scientific enterprise.
-Novel instruments are typically constructed of many individual components that are both purchased and home-built.
+Novel instruments are typically constructed of many individual components, purchased and home-built.
 Orchestration software must communicate with each hardware component in the course of a scientific experiment.
 This can involve utilization of many interfaces: NI DAQmx\cite{nidaqmx}, SCPI\cite{scpi}, ModBus\cite{modbus}, PICam\cite{picam}, Thorlabs APT\cite{thorlabs_apt}, among many others.
-The challenge of integrating all of these interfaces is a frustrating piece of the modern instrument development process.
+The challenge of integrating all of these is a frustrating piece of the modern instrument development process.
 Weeks can be spent just integrating one new component into an existing project.
 In small academic labs, these software interfaces are typically created by student researchers without software development experience.
-Student researchers rarely focus on software reusability, and a lack of maitenence and documentation can make such software more difficult to use as time goes on.
+Student researchers rarely focus on software reusability, and a lack of maintenance and documentation can make such software more difficult to use as time goes on.
 Scientists may struggle to rapidly innovate on their experimental design when each hardware addition requires major software development.
@@
 Small background services are written to translate the myriad component interfaces into the standard EPICS IOCs and TANGO Devices.
-These programs are performant, open source, and have huge libraries of exisiting hardware interface support, but require expert management to set up and provide descriptions via a separate server program.
+These programs are performant, open source, and have huge libraries of existing hardware interface support, but require expert management to set up and provide descriptions via a separate server program.
 In our experience EPICS and TANGO do not scale well to single-investigator lab environments.
@@
 
-We have created a new network-based communication standard for scientific instrumentation, \yaq{}.
-This standard borrows the most important ideas from established projects used by large user facilities while retaining the simplicity appropriate for small research labs.
+We have created a new network-based standard for scientific instrumentation, \yaq{}.
+This standard borrows the most important ideas from established projects used by large user facilities while retaining simplicity appropriate for small research labs.
 We have built this standard to be self-describing, portable, and reusable wherever possible.
@@
 Next, we discuss how inflexible orchestration software can limit experimental creativity.
-Then, we focus on challenges that arise when enhancing or modifying existing instruments with new hardware.
+Next, we focus on challenges that arise when enhancing or modifying existing instruments with new hardware.
 Finally, we discuss the heavy software maintenance burden that many instrument designers face.
@@
 Figure \ref{fig:network} diagrams the \yaq{} architecture.
-Here, we show three different computers connected via an ethernet network.
+Here, we show three different computers connected via an Ethernet network.
 The top and bottom computer are connected to monitors for interactive use while the middle computer is only accessible via the network.
@@
 A sensor might implement a message called ``\texttt{get\_measured}'' which takes no parameters and returns a dictionary mapping channel names to numeric or array data.
-A motor would implement a pair of messages for setting and reading back the motor position: ``\texttt{set\_position(float position)} $\rightarrow$ \texttt{null} '' and ``\texttt{get\_position()} $\rightarrow$ \texttt{float}''.
+A motor would implement a pair of messages for setting and reading back the motor position: ``\texttt{set\_position(float position)} $\rightarrow$ \texttt{null}'' and ``\texttt{get\_position()} $\rightarrow$ \texttt{float}''.
 These self describing messages make up the lowest level functionality of the \yaq{} interface.
@@
 The first hardware interface barrier: Multiple incompatible interfaces are used to communicate with each component of the system.
-% consider making this sentence (and the next paragraph starts) subsection headings
 \yaq{} provides a unified TCP/IP interface to all hardware peripherals based on the well-described Avro RPC protocol.
@@
 Developing experimental software is an iterative process tied to the scientific goals of the instrument.
-Often experimentalists must apply their specialty scientific knowledge to develop this software.\cite{SegalJudith2005a} % citation does not compile
+Often experimentalists must apply their specialty scientific knowledge to develop this software.\cite{SegalJudith2005a}
 \yaq{} is architected to encourage better software development practices when creating such programs.
@@
 \begin{figure}
-\includegraphics[width=\columnwidth]{figures/client_spectrum},
+\includegraphics[width=\columnwidth]{figures/client_spectrum}
 	\caption{  \label{fig:foundation} \yaq{} scales from simple scripts to integrations with featureful frameworks.}
@@
 
-A \yaq{} client can scale in complexity from a small, lightweight script all the way to a complicated graphical program, as shown in Figure \ref{fig:foundation}.
+A \yaq{} client can scale in complexity from a small, lightweight script all the way to a sophisticated graphical program, as shown in Figure \ref{fig:foundation}.
 At the bottom, we represent the most lightweight interface to \yaq{}, Python scripting.
@@
 Client sophistication can be introduced naturally, as novel experiments are tested and refined.
-Different clients can specialize for different requirements of a custom instrument
+Different clients can specialize for different requirements of a custom instrument.
 For example, \texttt{yaqc-qtpy} can provide a quick interface for setting and viewing hardware positions while Bluesky can focus on experimental data acquisition.
@@
 The Wright Group uses simulated hardware to enable client development away from the active laboratory computers.
-Remote client development on viritual hardware has allowed client design to proceed without delaying ongling experiments.
+Remote client development on virtual hardware has allowed client design to proceed without delaying ongoing experiments.
 Clients developed by the Wright Group have proved flexible enough to be used on four laser systems, each with different complements of hardware.
@@
 	These are extreme examples which are not representative of most hardware interfaces, so we omit them here.
-% it's not clear to me why these instances are not included (and scaled down???) It is a _feature_ not a bug that one can implement motor support with tiny amounts of code.
 	}
@@
 Figure \ref{fig:histogram} shows the distribution of unique lines of Python code written to implement each of the daemons in our current ecosystem.
-
 While lines of code is an imperfect metric, we use it here to represent the amount of work required to create a daemon for a new hardware peripheral.
-Most interfaces, such as Brooks MFC\cite{yaqd-brooks-mfc}, have been implemented in fewer than 100 lines of Python code.
+Most interfaces, such as Brooks MFC\cite{yaqd-brooks-mfc}, have been implemented in as less than 100 lines of Python code.
 Even the most complicated daemons are implemented in about 300 lines.
@@
 Software documentation is famously difficult and thankless work.
-\yaq{} attempts to automate daemon documentation as much as possible.
+\yaq{} attempts do automate daemon documentation as much as possible.
 Our website, https://yaq.fyi, automatically builds generated reference pages for all known protocols.
@@
 \section*{References}
-% all yaq,yaqc,and yaqd citations have an unfilled "()" in them... 
+
 \nocite{*}

### 2022-11-01 paper.tex
open source

@@
-\documentclass[aip, amsmath, amssymb, reprint,]{revtex4-1} % consider using RevTex4.2
+\documentclass[aip, amsmath, amssymb, reprint,]{revtex4-2}
 
@@
 Modern instrumentation development often involves incorporation of many dissimilar hardware peripherals into a single unified instrument.
-Increasing modular, commercial hardware availability has brought greater instrument complexity to small research groups.
+Increasing availability of modular hardware has brought greater instrument complexity to small research groups.
 The flexibility promised to instrument makers by large hardware availability stretches the capability of traditional, monolithic orchestration software.
-In many cases, a lack of software flexibility leads creative researchers to feel frustrated and unable to perform the experiments they envision even when highly modular hardware components enable extreme experimental flexibility.
-Herein, we describe \yaq{}, a software project defining a new standardized way of communicating with diverse hardware peripherals.
-\yaq{} encourages a highly modular approach to experimental software development, which is well suited to address the experimental flexibility needs of complex in
+In many cases, a lack of software flexibility leads creative researchers to feel frustrated, unable to perform experiments they envision.
+Herein we describe \yaq{}, a software project defining a new standardized way of communicating with diverse hardware peripherals.
+\yaq{} encourages a highly modular approach to experimental software development, which is well suited to address the experimental flexibility needs of complex instruments.
 \yaq{} is architected to overcome hardware communication barriers that are insurmountable with typical experimental software.
@@
 Since \yaq{} enables multiple machines, any hardware requirements can be addressed by putting a machine for that specific hardware on the network.
-For example, the Windows XP computer which has the obsolete interface card and drivers can be placed into a private network with only \yaq{} interface communication.
+The Raspberry Pi which drives several stepper motors can be placed onto a private network to communicate with the primary instrument computer using \yaq{}.
 Because each \yaq{} daemon is running in its own process, the software environment can be tailored to its needs.
@@
 The Wright Group uses simulated hardware to enable client development away from the active laboratory computers.
-Remote client development on virtual hardware has allowed client design to proceed without delaying ongoing experiments.
+This allows Wright Group researchers to create polished client interfaces without interrupting  ongoing experiments.
 Clients developed by the Wright Group have proved flexible enough to be used on four laser systems, each with different complements of hardware.
@@
 Figure \ref{fig:histogram} shows the distribution of unique lines of Python code written to implement each of the daemons in our current ecosystem.
-
 While lines of code is an imperfect metric, we use it here to represent the amount of work required to create a daemon for a new hardware peripheral.
@@
 Years after the original researchers leave, large monolithic acquisition programs become unknowable, undocumented, and unmaintained.
-A graduate student discovers a long standing hard-coded conversion factor that is incorrect years after implementation.
+A graduate student discovers a hard-coded conversion factor that is incorrect years after implementation.
 Scientists resort to sourcing a replacement for an old, broken oscilloscope due to their reliance of their software on that particular interface while newer, cheaper oscilloscopes are readily available.
@@
 Anyone can view, install, edit, and suggest changes to our growing collection of daemons and clients.
+Furthermore, anyone can create their own totally-custom client or daemon software separately by following the specified \yaq{} standard.
 Thirteen individuals have contributed code to the development of \yaq{}.
+Open-source development can be a powerful approach for research communities looking to share software-development and maintenance burdens. \cite{CohenJeremy2021a}
+It is our hope that a vibrant open-source community will form around \yaq{}.
+While open-source development is not a panacea \cite{NowogrodzkiAnna2019a}, we hope that by maintaining a distributed development strategy with a strict focus on only hardware interfaces the \yaq{} project might prove sustainable.
 

### 2022-11-01 paper.tex
Changes from meeting

@@
 
-\title{The yaq Project: \\ A Standard for Scientific Instrumentation}
+\title{The yaq Project: \\ Standardized Software Enabling Flexible Instrumentation}
 \author{Kyle F. Sunden}
@@
 \author{Kent A. Meyer}
-\author{Peter L. Cruz Parilla}
+\author{Peter L. Cruz Parrilla}
 \author{John C. Wright}
@@
 Increasing availability of modular hardware has brought greater instrument complexity to small research groups.
-The flexibility promised to instrument makers by large hardware availability stretches the capability of traditional, monolithic orchestration software.
+%The flexibility promised to instrument makers by large hardware availability stretches the capability of traditional, monolithic orchestration software.
+This complexity stretches the capability of traditional, monolithic orchestration software.
 In many cases, a lack of software flexibility leads creative researchers to feel frustrated, unable to perform experiments they envision.
@@
 \yaq{} encourages a highly modular approach to experimental software development, which is well suited to address the experimental flexibility needs of complex instruments.
-\yaq{} is architected to overcome hardware communication barriers that are insurmountable with typical experimental software.
-A large number of hardware peripherals are already supported with tooling available to create more.
+\yaq{} is designed to overcome hardware communication barriers that are insurmountable with typical experimental software.
+A large number of hardware peripherals are already supported, with tooling available to expand support.
 The \yaq{} standard enables collaboration among multiple research groups, increasing code quality while lowering development effort.
@@
 \yaq{} introduces a concept called ``Traits'' which are collections of related messages that are shared among multiple protocols.
-Motors implement the ``has-position'' trait, which defines ``\texttt{set\_position}'', ``\texttt{get\_position}'', and ``\texttt{get\_units}''.
-Sensors would implement the ``is-sensor'' trait which defines ``\texttt{get\_measured}'', ``\texttt{get\_channel\_names}'', and ``\texttt{get\_channel\_units}''.
+Motors implement the ``\texttt{has-position}'' trait, which defines ``\texttt{set\_position}'', ``\texttt{get\_position}'', and ``\texttt{get\_units}''.
+Sensors would implement the ``\texttt{is-sensor}'' trait which defines ``\texttt{get\_measured}'', ``\texttt{get\_channel\_names}'', and ``\texttt{get\_channel\_units}''.
 Protocols which implement a trait must support all of the messages from the trait.

### 2022-11-03 paper.tex
hyphenate self-describing

@@
 A motor would implement a pair of messages for setting and reading back the motor position: ``\texttt{set\_position(float position)} $\rightarrow$ \texttt{null}'' and ``\texttt{get\_position()} $\rightarrow$ \texttt{float}''.
-These self describing messages make up the lowest level functionality of the \yaq{} interface.
+These self-describing messages make up the lowest level functionality of the \yaq{} interface.
 Each individual communication between client and daemon involves one message being requested by the client and the response returned from the daemon.
@@
 \texttt{yaqc-qtpy}\cite{yaqc-qtpy} is a graphical application which builds interactive controls based on traits for any conceivable \yaq{} protocol.
-The self describing \yaq{} interface is used to provide graphical elements for the most commonly used messages.
+The self-describing \yaq{} interface is used to provide graphical elements for the most commonly used messages.
 This is an invaluable tool which provides a ``free'' graphical user interface (GUI) to any daemon.
@@
 Second, \texttt{yaq-traits}\cite{yaq-traits} is a command line application which allows the description of messages provided by a \yaq{} protocol to be written in a human-readable fashion and translated into a more fully described machine readable format.
-The format it generates is an important part of how \yaq{} protocols are self describing.
+The format it generates is an important part of how \yaq{} protocols are self-describing.
 This shields developers from the details of Apache Avro, which can be somewhat esoteric.

### 2022-11-06 paper.tex
Fill out citations

--- 2022-11-01 paper.tex
@@
 The Wright Group at UW-Madison needs to orchestrate a large variety of hardware in multidimensional scans for their complex spectroscopy experiments \cite{MukamelShaul2000a, WrightJohnCurtis2011a}.
-This need for exquisite hardware control has resulted in several prior attempts at ``home-built'' orchestration software \cite{CarlsonRogerJohn1988a, MeyerKentAlbert2004b, KainSchuyler2017a, ThompsonBlaiseJonathan2018a}.
+This need for exquisite hardware control has resulted in several prior attempts at ``home-built'' orchestration software \cite{CarlsonRogerJohn1988a, MeyerKentAlbert2004b, KainSchuyler2017a, ThompsonBlaiseJonathan2018a, SundenKyle2022}.
 Now, using \yaq{}, the Wright Group has been able to move to Bluesky rather than inventing their own sophisticated control software ``from scratch''.
@@
 Even the most complicated daemons are implemented in about 300 lines.
-Implementing \yaq{} daemons using Python is enabled through our own tooling mentioned above and the large and growing ecosystem of hardware interface tooling libraries that Python now provides\cite{}.
+Implementing \yaq{} daemons using Python is enabled through our own tooling mentioned above and the large and growing ecosystem of hardware interface tooling libraries that Python now provides\cite{pyserial, pyusb, pyvisa, pymodbus}.
 In our experience, the process of creating a new daemon involves about a day of work after communication with the hardware using Python is well understood.
@@
 Because \yaq{} is standards based, anyone can design and publish new daemons extending our hardware support.
-A living list of all daemons and supported hardware can be found on the \yaq{} website.\cite{}
+A living list of all daemons and supported hardware can be found on the \yaq{} website.\cite{yaq_fyi_daemons}
 
@@
 A graduate student is forced to meticulously reverse engineer the LabVIEW codebase that they inherited in order to understand the details of their experiment.
-Software developers refer to the extra effort required to modify or fix large unmaintained codebases as ``technical debt''.\cite{}
+Software developers refer to the extra effort required to modify or fix large unmaintained codebases as ``technical debt''.\cite{Allman_2012}
 Technical debt grows especially fast in academic environments where graduate students are involved in projects for a limited time.

### 2022-11-11 paper.tex

@@
 
+\section*{Supplementary Material}
+
+Full set of line-count data and script used to produce Figure \ref{fig:histogram}.
+
 \begin{acknowledgments}
-  TODO
+
+The authors would like to thank all \yaq{} users and contributors.
+We would also like to acknowledge the developers of the broader open source software community upon-which this project rests.
+
+This work was supported by the National Science Foundation under grant CHE-1709060.
+
 \end{acknowledgments}
@@
 
-Data sharing is not applicable to this article as no new data were created or analyzed in this study.
+The data that supports the findings of this study are available within the supplementary material.
 

### 2022-11-11 paper.tex
captions

@@
     \includegraphics[width=\columnwidth]{figures/network}
-\caption{\label{fig:network} Networking diagram}
+    \caption{\label{fig:network} Simplified network diagram for a hypothetical instrument.
+Three computers are connected via orange Ethernet cables and a network switch (middle right).
+Some of the computers are also connected to hardware peripherals via one of six interfaces (colored lines).
+For each interface, the computer runs a \yaq{} daemon (small colored squares).
+The top and bottom computers are also running \yaq{} client applications.}
 \end{figure}
@@
   \includegraphics[width=\columnwidth]{figures/client_spectrum}
-	\caption{  \label{fig:foundation} \yaq{} scales from simple scripts to integrations with featureful frameworks.}
+  \caption{\label{fig:foundation} Representations of three very different \yaq{} client experiences.
+Top: light-weight scripting.
+Middle: graphical user interface.
+Bottom: integration with Bluesky for data collection.
+In all three cases the same daemons are addressed.}
 \end{figure}

### 2022-11-11 paper.tex
references

@@
 These monolithic applications tend to be inflexible and difficult to develop.
-%To our knowledge, there are no existing open source projects that attempt to define a unified standard for hardware control of small scale experiments.
 
@@
 This paper provides an overview of the concepts and motivations behind \yaq{}.
-Refer to the \yaq{} website for a formal specification of the \yaq{} standard.\cite{yaq_fyi}
+Refer to the \yaq{} website for a formal specification of the \yaq{} standard.\footnote{The \yaq{} protocol is fully specified by the set of accepted \yaq{} Enhancement Proposals at https://yeps.yaq.fyi/}
 
@@
 At the bottom, we represent the most lightweight interface to \yaq{}, Python scripting.
-\texttt{yaqc}\cite{yaqc} is a Python client which is excellent for using in scripts or any other Python program.
+\texttt{yaqc}\footnote{\texttt{yaqc} available on PyPI at https://pypi.org/project/yaqc/} is a Python client which is excellent for using in scripts or any other Python program.
 The code shown creates three client objects which could be used directly through an interactive Python prompt or in a reusable script.
@@
 In the middle, the same three hardware peripherals are represented in an interactive, graphical form.
-\texttt{yaqc-qtpy}\cite{yaqc-qtpy} is a graphical application which builds interactive controls based on traits for any conceivable \yaq{} protocol.
+\texttt{yaqc-qtpy}\footnote{\texttt{yaqc-qtpy} available on PyPI at https://pypi.org/project/yaqc-qtpy/} is a graphical application which builds interactive controls based on traits for any conceivable \yaq{} protocol.
 The self-describing \yaq{} interface is used to provide graphical elements for the most commonly used messages.
@@
 Shown in Figure \ref{fig:foundation}, we see one such GUI for parameterizing a Bluesky procedure and a representation of the resultant data.
-\texttt{yaqc-bluesky}\cite{yaqc-bluesky} provides a bridge to the Bluesky ecosystem.
+\texttt{yaqc-bluesky}\footnote{\texttt{yaqc-bluesky} available on PyPI at https://pypi.org/project/yaqc-bluesky/} provides a bridge to the Bluesky ecosystem.
 Similar translation layers could be built for a variety of orchestration layers such as PyMoDAQ\cite{WeberSebastien2021a}, Instrumental\cite{Bogdanowicz2022}, or TRSpectrometer\cite{trspectrometer}.
@@
 We have created several tools to aide in daemon development.
-First, a Python library, \texttt{yaqd-core}\cite{yaqd-core-python}, which implements shared functionality.
-Second, \texttt{yaq-traits}\cite{yaq-traits} is a command line application which allows the description of messages provided by a \yaq{} protocol to be written in a human-readable fashion and translated into a more fully described machine readable format.
+First, a Python library, \texttt{yaqd-core}\footnote{\texttt{yaqd-core-python} available on PyPI at https://pypi.org/project/yaqd-core}, which implements shared functionality.
+Second, \texttt{yaq-traits}\footnote{\texttt{yaq-traits} available on PyPI at https://pypi.org/project/yaq-traits} is a command line application which allows the description of messages provided by a \yaq{} protocol to be written in a human-readable fashion and translated into a more fully described machine readable format.
 The format it generates is an important part of how \yaq{} protocols are self-describing.
@@
 While lines of code is an imperfect metric, we use it here to represent the amount of work required to create a daemon for a new hardware peripheral.
-Most interfaces, such as Brooks MFC\cite{yaqd-brooks-mfc}, have been implemented in fewer than 100 lines of Python code.
+Most interfaces, such as Brooks MFC\footnote{\texttt{yaqd-brooks-mfc-gf} source code available on GitHub at \texttt{https://github.com/yaq-project/yaqd-brooks/}}, have been implemented in fewer than 100 lines of Python code.
 Even the most complicated daemons are implemented in about 300 lines.
@@
 Because \yaq{} is standards based, anyone can design and publish new daemons extending our hardware support.
-A living list of all daemons and supported hardware can be found on the \yaq{} website.\cite{yaq_fyi_daemons}
+A living list of all daemons and supported hardware can be found on the \yaq{} website.\footnote{See https://yaq.fyi/daemons/ and https://yaq.fyi/hardware/ for living lists of all \yaq{} daemons and supported hardware, respectively.}
 
@@
 The lack of hardware interface code makes \yaq{} clients much simpler and easier to describe and maintain.
-Tools like \texttt{yaqd-fakes}\cite{yaqd-fakes} allow clients to be tested and improved outside of their instrument, including the possibility of fully automated testing.
+Tools like \texttt{yaqd-fakes}\footnote{\texttt{yaqd-fakes} available on PyPI at https://pypi.org/project/yaqd-fakes/} allow clients to be tested and improved outside of their instrument, including the possibility of fully automated testing.
 Simple, script-based clients written using the expressiveness of Python can be read and understood in hours rather than weeks.

### 2022-11-11 paper.tex
multilingual

@@
 Increasing availability of modular hardware has brought greater instrument complexity to small research groups.
-%The flexibility promised to instrument makers by large hardware availability stretches the capability of traditional, monolithic orchestration software.
 This complexity stretches the capability of traditional, monolithic orchestration software.
@@
 Herein we describe \yaq{}, a software project defining a new standardized way of communicating with diverse hardware peripherals.
-\yaq{} encourages a highly modular approach to experimental software development, which is well suited to address the experimental flexibility needs of complex instruments.
+\yaq{} encourages a highly modular approach to experimental software development which is well suited to address the experimental flexibility needs of complex instruments.
 \yaq{} is designed to overcome hardware communication barriers that are insurmountable with typical experimental software.
@@
 
-Here, we discuss the design of the \yaq{} standard in the context of four challenges facing instrument designers.
+Here, we discuss the design of the \yaq{} standard in the context of challenges facing instrument designers.
 First, we discuss how particularly challenging hardware interfaces can become seemingly insurmountable barriers to software control.
@@
 A graduate student discovers a hard-coded conversion factor that is incorrect years after implementation.
-Scientists resort to sourcing a replacement for an old, broken oscilloscope due to their reliance of their software on that particular interface while newer, cheaper oscilloscopes are readily available.
+Scientists resort to sourcing an exact replacement for an old, broken oscilloscope due to their software's reliance on that particular interface---newer, cheaper oscilloscopes are readily available.
 A graduate student is forced to meticulously reverse engineer the LabVIEW codebase that they inherited in order to understand the details of their experiment.
@@
 Growing this ecosystem is a collaborative effort where many \yaq{} users create portable daemons that they need and share them with the community where they will be used and improved.
+\yaq{} components can be incorporated seamlessly into existing software projects, including languages other than Python.
+For example, an existing LabVIEW project may add additional hardware support via \yaq{}.
+In this way, a more modular instrument can be built gradually without needing to overhaul all of the software simultaneously.
 
@@
 These pages are automatically updated when new versions are published.
+Our website also contains written and video tutorials on \yaq{} usage and development.
 

### 2022-11-11 paper.tex
polish

@@
 A motor would implement a pair of messages for setting and reading back the motor position: ``\texttt{set\_position(float position)} $\rightarrow$ \texttt{null}'' and ``\texttt{get\_position()} $\rightarrow$ \texttt{float}''.
-These self-describing messages make up the lowest level functionality of the \yaq{} interface.
+Messages make up the lowest level functionality of the \yaq{} interface.
 Each individual communication between client and daemon involves one message being requested by the client and the response returned from the daemon.
 Each daemon supports a collection of messages for its unique functionality, called the ``protocol''.
+When a client first connects to a daemon over TCP/IP, the daemon provides a complete description of its own protocol.
+We therefore refer to the daemons as ``self-describing''.
 
@@
 
-A \yaq{} client can scale in complexity from a small, lightweight script all the way to a sophisticated graphical program, as shown in Figure \ref{fig:foundation}.
-At the bottom, we represent the most lightweight interface to \yaq{}, Python scripting.
+Figure \ref{fig:foundation} shows that \yaq{} supports a diverse array of client types.
+At the top, we represent the most lightweight interface to \yaq{}, Python scripting.
 \texttt{yaqc}\footnote{\texttt{yaqc} available on PyPI at https://pypi.org/project/yaqc/} is a Python client which is excellent for using in scripts or any other Python program.
@@
 
-At the top, we show several features associated with the integration between \yaq{} and the Bluesky project\cite{AllanDanielB2019a}.
+At the bottom, we show several features associated with the integration between \yaq{} and the Bluesky project\cite{AllanDanielB2019a}.
 Bluesky provides a powerful orchestration layer for conducting and recording data for a wide variety of experimental procedures.

### 2022-11-11 paper.tex
minor changes

@@
 Clients developed by the Wright Group have proved flexible enough to be used on four laser systems, each with different complements of hardware.
-Moving forward, the Wright Group will spend less energy developing control software and more energy developing creative spectroscopy experiments.
+The Wright Group expects to spend less energy developing control software and more energy developing creative spectroscopy experiments in future work.
 
@@
 Tooling exists to verify full trait compliance, for example you can use \texttt{yaqc-qtpy} to provide a graphical program to interact with your hardware immediately.
-Finally, as discussed in Section II, \yaq{} gives you options to design using remote hardware or unusual interfaces when necessary.
+Finally, as discussed in Section II, \yaq{} provides options to design using remote hardware or unusual interfaces when necessary.
 
@@
 \yaq{} components can be incorporated seamlessly into existing software projects, including languages other than Python.
-For example, an existing LabVIEW project may add additional hardware support via \yaq{}.
+For example, an existing LabVIEW project may provide additional hardware support via \yaq{}.
 In this way, a more modular instrument can be built gradually without needing to overhaul all of the software simultaneously.

### 2022-11-11 paper.tex
minor changes (#29)
* minor changes

* Update paper.tex

Co-authored-by: Blaise Thompson <blaise@untzag.com>
hardware recount

@@
 Clients developed by the Wright Group have proved flexible enough to be used on four laser systems, each with different complements of hardware.
-The Wright Group expects to spend less energy developing control software and more energy developing creative spectroscopy experiments in future work.
+Moving forward, the Wright Group will spend less energy developing control software and more energy developing creative spectroscopy experiments.
 
@@
 In our experience, the process of creating a new daemon involves about a day of work after communication with the hardware using Python is well understood.
-There are currently 70 daemons in the yaq project supporting at least 50 kinds of hardware, noting that some daemons support the same hardware and others are software only.
+There are currently 72 daemons in the yaq project supporting at least 66 types of hardware, noting that some daemons support the same hardware and others are software only.
 Because \yaq{} is standards based, anyone can design and publish new daemons extending our hardware support.

### 2022-11-17 paper.tex
JCW comments

@@
 Orchestration software must communicate with each hardware component in the course of a scientific experiment.
-This can involve utilization of many interfaces: NI DAQmx\cite{nidaqmx}, SCPI\cite{scpi}, ModBus\cite{modbus}, PICam\cite{picam}, Thorlabs APT\cite{thorlabs_apt}, among many others.
+Orchestration can involve utilization of many interfaces: NI DAQmx\cite{nidaqmx}, SCPI\cite{scpi}, ModBus\cite{modbus}, PICam\cite{picam}, Thorlabs APT\cite{thorlabs_apt}, among many others.
 The challenge of integrating all of these interfaces is a frustrating piece of the modern instrument development process.
@@
 Here, we show three different computers connected via an Ethernet network.
-The top and bottom computer are connected to monitors for interactive use while the middle computer is only accessible via the network.
-This might represent a complex scientific instrument involving several operator terminals as well as embedded computers.
+The top and bottom computer are connected to monitors for interactive use while the middle computer is only accessible via the networks.
+This diagram might represent a complex scientific instrument involving several operator terminals as well as embedded computers.
 At the top, a single computer is connected to four hardware peripherals through RS232, TTL, USB, and PCI as indicated by the colored lines.
@@
 The self-describing \yaq{} interface is used to provide graphical elements for the most commonly used messages.
-This is an invaluable tool which provides a ``free'' graphical user interface (GUI) to any daemon.
+\texttt{yaqc-qtpy} is an invaluable tool which provides a ``free'' graphical user interface (GUI) to any daemon.
 
@@
 The Wright Group uses simulated hardware to enable client development away from the active laboratory computers.
-This allows Wright Group researchers to create polished client interfaces without interrupting  ongoing experiments.
+Hardware simulation allows Wright Group researchers to create polished client interfaces without interrupting  ongoing experiments.
 Clients developed by the Wright Group have proved flexible enough to be used on four laser systems, each with different complements of hardware.
@@
 Often, researchers can download an existing daemon rather than writing a new one.
-Second, \yaq{} daemon development can be performed separated from the particulars of any individual client.
-Often, this means that initial hardware enablement work can be done on a researchers personal machine before the new hardware peripheral is installed in the instrument.
+Second, \yaq{} daemon development can be performed separately from the particulars of any individual client.
+Often, this separation allows initial hardware enablement work to be done on a researchers personal machine before the new hardware peripheral is installed in the instrument.
 Third, when developing trait-compliant protocols, it becomes easy to design and fully test your hardware interface.
@@
 Implementing \yaq{} daemons using Python is enabled through our own tooling mentioned above and the large and growing ecosystem of hardware interface tooling libraries that Python now provides\cite{pyserial, pyusb, pyvisa, pymodbus}.
-In our experience, the process of creating a new daemon involves about a day of work after communication with the hardware using Python is well understood.
+In our experience, the process of creating a new daemon involves about a day of work after mastering communication with the hardware.
 There are currently 72 daemons in the yaq project supporting at least 66 types of hardware, noting that some daemons support the same hardware and others are software only.
@@
 It is our hope that a vibrant open-source community will form around \yaq{}.
-While open-source development is not a panacea \cite{NowogrodzkiAnna2019a}, we hope that by maintaining a distributed development strategy with a strict focus on only hardware interfaces the \yaq{} project might prove sustainable.
+While open-source development is not a panacea \cite{NowogrodzkiAnna2019a}, we hope that by maintaining a distributed development strategy with a strict focus on only hardware interfaces, the \yaq{} project might prove sustainable.
 

### 2022-11-17 paper.tex
DDK edits

@@
 
-\yaq{} introduces a concept called ``Traits'' which are collections of related messages that are shared among multiple protocols.
+\yaq{} introduces a concept called ``traits'' which are collections of related messages that are shared among multiple protocols.
 Motors implement the ``\texttt{has-position}'' trait, which defines ``\texttt{set\_position}'', ``\texttt{get\_position}'', and ``\texttt{get\_units}''.
@@
 Because each \yaq{} daemon is running in its own process, the software environment can be tailored to its needs.
-A client running up to date Python 3.10 communicates seamlessly with a daemon running Python 3.7.
+A client running up-to-date Python 3.11 communicates seamlessly with a daemon running Python 3.7.
 
@@
 In these instances, the initial message simply starts the action and returns, with a separate message provided to retrieve results when they exist.
-In order to know how long to wait, the ``is-daemon`` trait provides a message called ``is\_busy'' which should return "true" while the long running action is not complete, and "false" once it is finished.
+In order to know how long to wait, the ``is-daemon`` trait provides a message called ``is\_busy'', which returns "true" while the long running action is not complete, and "false" once it is finished.
 Additionally, multiple clients can communicate with the same daemon simultaneously.
@@
 Existing experimental orchestration software is often highly inflexible.
-An experimentalist will spend many hours in lab repeating acquisitions because it is too challenging to add repetition functionality to their software.
+An experimentalist will spend many hours in lab manually repeating acquisitions because it is too challenging to add repetition functionality to their software.
 A laser lab needs to spend weeks on software development when introducing a single new step into their experimental procedure.
 Researchers are disappointed to realize that they are forced to start from scratch when developing software for a similar instrument built with trivially different hardware.
+
+Unique experiments will always need custom orchestration and user experience.
+We believe that novel instrumentation development naturally and necessarily includes the creation of targeted software.
+Developing experimental software is an iterative process tied to the scientific goals of the instrument.
+Often experimentalists must apply their specialty scientific knowledge to develop this software.\cite{SegalJudith2005a} % citation does not compile
 
@@
 
-Unique experiments will always need custom orchestration and user experience.
-We believe that novel instrumentation development naturally and necessarily includes the creation of targeted software.
-Developing experimental software is an iterative process tied to the scientific goals of the instrument.
-Often experimentalists must apply their specialty scientific knowledge to develop this software.\cite{SegalJudith2005a} % citation does not compile
 \yaq{} is architected to encourage better software development practices when creating such programs.

### 2023-01-24 SI.tex
supporting information tex file (#30)

% document
\documentclass[11pt, full]{article}
\usepackage[letterpaper, portrait, margin=0.75in]{geometry}
\usepackage{setspace}
\usepackage{color}

% text
\usepackage[utf8]{inputenc}
\setlength\parindent{0pt}
\setlength{\parskip}{1em}
\usepackage{enumitem}
\renewcommand{\familydefault}{\sfdefault}
\newcommand{\RomanNumeral}[1]{\textrm{\uppercase\expandafter{\romannumeral #1\relax}}}

% math
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage[cm]{sfmath}
\usepackage{commath}
\usepackage{multirow}
\DeclareMathAlphabet{\mathpzc}{OT1}{pzc}{m}{it}

% graphics
\usepackage{graphics}
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{epstopdf}
\usepackage{xpatch}
\graphicspath{{./figures/}}

% "S" prefix
\renewcommand{\theequation}{S\arabic{equation}}
\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thetable}{S\arabic{table}}

% each section begins new page
\let\stdsection\section
\renewcommand\section{\clearpage\stdsection}

% force all floats to center (see https://tex.stackexchange.com/a/53383)
\makeatletter
\g@addto@macro\@floatboxreset{\centering}
\makeatother

% bibliography
\usepackage[backend=bibtex, natbib=true, url=false, sorting=none, maxbibnames=99]{biblatex}
\bibliography{references}

% hyperref
\usepackage[colorlinks=true, linkcolor=black, urlcolor=blue, citecolor=black, anchorcolor=black]
  {hyperref}
\usepackage[all]{hypcap}  % helps hyperref work properly

\begin{document}
\pagenumbering{gobble}

\begin{center}
  \LARGE

  \textit{Supporting Information} \\
  The yaq Project: \\
  Standardized Software Enabling Flexible Instrumentation

  \normalsize

  \textit{Kyle F. Sunden, Daniel D. Kohler, Kent A. Meyer, Peter L. Cruz Parrilla, \\
    John C. Wright, and Blaise J. Thompson*}

  Department of Chemistry, University of Wisconsin--Madison\\
  1101 University Ave., Madison, Wisconsin 53706
\end{center}

\vspace{\fill}

*Corresponding Author \\
\hspace*{2ex} email: blaise.thompson@wisc.edu \\
\hspace*{2ex} phone: (608) 263-2573

\pagebreak
\setcounter{page}{1}
\pagenumbering{arabic}
\renewcommand{\thepage}{S\arabic{page}}

\pagebreak
\renewcommand{\baselinestretch}{0.75}\normalsize
\tableofcontents
\renewcommand{\baselinestretch}{1.0}\normalsize

\section{Package size analysis}

\clearpage

\printbibliography

\end{document}

### 2023-01-24 build.sh

@@ -3,3 +3,8 @@ pdflatex --interaction=nonstopmode paper.tex
 bibtex paper.aux
 pdflatex --interaction=nonstopmode paper.tex
 pdflatex --interaction=nonstopmode paper.tex
+#
+pdflatex --interaction=nonstopmode SI.tex
+bibtex SI.aux
+pdflatex --interaction=nonstopmode SI.tex
+pdflatex --interaction=nonstopmode SI.tex

### 2023-01-30 paper.tex
small changes (#31)

@@
 \usepackage{etoolbox}
+\usepackage{hyperref}
+\usepackage{url}
 
@@
 \yaq{} encourages a highly modular approach to experimental software development which is well suited to address the experimental flexibility needs of complex instruments.
-\yaq{} is designed to overcome hardware communication barriers that are insurmountable with typical experimental software.
+\yaq{} is designed to overcome hardware communication barriers that challenge typical experimental software.
 A large number of hardware peripherals are already supported, with tooling available to expand support.
@@
 Even the most complicated daemons are implemented in about 300 lines.
-Implementing \yaq{} daemons using Python is enabled through our own tooling mentioned above and the large and growing ecosystem of hardware interface tooling libraries that Python now provides\cite{pyserial, pyusb, pyvisa, pymodbus}.
+Implementing \yaq{} daemons using Python is enabled through our own tooling mentioned above and the large and growing ecosystem of hardware interface libraries that Python now provides\cite{pyserial, pyusb, pyvisa, pymodbus}.
 In our experience, the process of creating a new daemon involves about a day of work after mastering communication with the hardware.
@@
 
-Software documentation is famously difficult and thankless work.
+Open source software built by scientists often has incomplete and inadequate documentation.
 \yaq{} attempts to automate daemon documentation as much as possible.
@@
 \section*{References}
-% all yaq,yaqc,and yaqd citations have an unfilled "()" in them...
 \nocite{*}

### 2023-02-03 paper.tex
cite
address many motors

@@
 
-The third hardware interface barrier: Some hardware interfaces are blocking.
-Monolithic orchestration software often necessitates separate threads for each component hardware interface, a fragile pattern in which small mistakes become both critical and elusive errors.
-In \yaq{}, each daemon is only responsible for a single hardware interface.
-%Each \yaq{} daemon runs as a separate process (possibly not even on the same machine) and thus many of the hard concurrency problems with communicating with hardware are mitigated.
-It is expected that each message call over the \yaq{} interface will return rapidly, ensuring that client applications are not blocked for extended periods of time.
-This principle applies even when the message starts an action that might take several seconds to complete, such as homing a motor or initiating a measurement for a sensor.
-In these instances, the initial message simply starts the action and returns, with a separate message provided to retrieve results when they exist.
-In order to know how long to wait, the ``is-daemon`` trait provides a message called ``is\_busy'', which returns "true" while the long running action is not complete, and "false" once it is finished.
-Additionally, multiple clients can communicate with the same daemon simultaneously.
-A complex instrument may involve multiple operators watching sensor data in real time, while one program is orchestrating the hardware and recording the data.
+The third hardware interface barrier: Some hardware interfaces are blocking or slow.
+Experimental orchestration often means doing several things at once, \textit{e.g.} simultaneously moving several motors.
+To accomplish this task performantly, monolithic orchestration software often necessitates separate threads for each component hardware interface, a fragile pattern in which small mistakes become both critical and elusive errors.
+When writing orchestration software using \yaq{}, slow hardware interfaces are replaced by fast Avro-RPC.
+In our experience, the \yaq{} interface is responsive enough to give good performace when orchestrating tens of motors using a singly-threaded client.
+There is a longer discussion of timing and order-of-operations control details using \yaq{} in the Supplementary Material.
 
@@
 
-Open source software built by scientists often has incomplete and inadequate documentation.
+Software built by scientists often has incomplete and inadequate documentation. \cite{SegalJudith2007a}
 \yaq{} attempts to automate daemon documentation as much as possible.

### 2023-02-03 SI.tex
address many motors (#33)

@@
 
+Each message call over the yaq interface returns rapidly,
+ensuring that client applications are not blocked. Some mes-
+sages initiate long-running operations, e.g. motor motion and
+sensor measurement. Separate messages are provided to re-
+trieve results from e.g. sensor measurement. In order to know
+how long to wait, all yaq daemons provide a message called
+“is\_busy”, which returns "true" while the long running action
+is not complete, and "false" once it is finished. Additionally,
+multiple clients can communicate with the same daemon si-
+multaneously. A complex instrument may involve multiple
+operators watching sensor data in real time, while one pro-
+gram is orchestrating the hardware and recording the data.
+
 \section{Package size analysis}

### 2023-02-05 figures/scaling/plot-scaling.py
Add scaling data gen and plotting

@@ -1,0 +1,32 @@
+import pathlib
+
+import pandas as pd
+import matplotlib.pyplot as plt
+
+__here__ = pathlib.Path(__file__).parent
+
+data = pd.read_csv(__here__ / "scaling.csv")
+
+fig, ax = plt.subplots()
+
+ax.errorbar(data["npts"], data["mean"], yerr=data["stdev"], marker="o", lw=0, elinewidth=2, capsize=3, capthick=2, zorder=3)
+ax.grid()
+ax.set_xscale("log", base=2)
+ax.set_yscale("log")
+
+ax.set_xlabel("Number of Pixels")
+ax.set_ylabel("Time (seconds)")
+
+for pt in data.itertuples():
+    if pt.shapex != 1:
+        ax.errorbar([pt.npts], [pt.mean], yerr=[pt.stdev], marker="o", lw=0, elinewidth=2, capsize=3, capthick=2, zorder=3.1)
+        ax.annotate(
+            str((pt.shapex, pt.shapey)),
+            (pt.npts, pt.mean),
+            xytext=(-80, 10),
+            textcoords="offset points",
+            arrowprops=dict(arrowstyle="->"),
+        )
+
+plt.show()
+#fig.savefig("scaling.png", dpi=300)

### 2023-02-05 figures/scaling/plot-scaling.py
Include scaling png

@@ -28,5 +28,5 @@ for pt in data.itertuples():
             arrowprops=dict(arrowstyle="->"),
         )
 
-plt.show()
-#fig.savefig("scaling.png", dpi=300)
+#plt.show()
+fig.savefig("scaling.png", dpi=300)

### 2023-02-05 figures/scaling/scaling.py
Add scaling data gen and plotting (#34)
* Add scaling data gen and plotting

* Include scaling png

@@ -1,0 +1,13 @@
+import yaqd_core
+import numpy as np
+
+
+class NdarrayTestDaemon(yaqd_core.Base):
+    _kind = "scaling"
+
+    def rand_int32(self, dim1, dim2):
+        return np.random.randint(0, 4096, dim1 * dim2).reshape(dim1, dim2).astype(np.int32)
+
+
+if __name__ == "__main__":
+    NdarrayTestDaemon.main()

### 2023-02-05 paper.tex
provide hardware support
add sentence to part 2
add sentence to part 2 (#36)

@@
 We have built this standard to be self-describing, portable, and reusable wherever possible.
-Our primary goal has been to make an interface which simplifies orchestration software development as much as possible.
+Via the \yaq{} interface, we provide easy-to-use hardware support for scientists to use with their choice of experiment orchestration software.
 
@@
 
+We now describe how the \yaq{} architecture addresses the three hardware interface barriers described at the beginning of this section.
 The first hardware interface barrier: Multiple incompatible interfaces are used to communicate with each component of the system.

### 2023-02-05 paper.tex
rewrite description of figure 2 (#37)

@@
 
-At the bottom, we show several features associated with the integration between \yaq{} and the Bluesky project\cite{AllanDanielB2019a}.
+At the bottom, we graphically represent the integration between \yaq{} and the Bluesky project\cite{AllanDanielB2019a}.
 Bluesky provides a powerful orchestration layer for conducting and recording data for a wide variety of experimental procedures.
-Shown in Figure \ref{fig:foundation}, we see one such GUI for parameterizing a Bluesky procedure and a representation of the resultant data.
-\texttt{yaqc-bluesky}\footnote{\texttt{yaqc-bluesky} available on PyPI at https://pypi.org/project/yaqc-bluesky/} provides a bridge to the Bluesky ecosystem.
+Bluesky has no built-in hardware interface support, instead relying on packages to create hardware-interfaces that are compatible with their Hardware Protocol defintions.
+\texttt{yaqc-bluesky}\footnote{\texttt{yaqc-bluesky} available on PyPI at https://pypi.org/project/yaqc-bluesky/} is a specialized client which adapts any \yaq{} protocol into Bluesky's Hardware Protocol.
 Similar translation layers could be built for a variety of orchestration layers such as PyMoDAQ\cite{WeberSebastien2021a}, Instrumental\cite{Bogdanowicz2022}, or TRSpectrometer\cite{trspectrometer}.

### 2023-02-05 SI.tex
Write SI section for scaling analysis and histogram

@@
 \renewcommand{\thepage}{S\arabic{page}} 
+\renewcommand{\thefigure}{S\arabic{figure}}
 
@@
 
-Each message call over the yaq interface returns rapidly,
+\section{Timing and interface responsiveness}
+
+Each message call over the \yaq{} interface returns rapidly,
 ensuring that client applications are not blocked. Some mes-
@@
 sensor measurement. Separate messages are provided to re-
-trieve results from e.g. sensor measurement. In order to know
-how long to wait, all yaq daemons provide a message called
+trieve results from e.g. sensor measurement.
+
+\subsection{\texttt{is\_busy} for orchestrating order of operations}
+
+In order to know how long to wait, all \yaq{} daemons provide a message called
 “is\_busy”, which returns "true" while the long running action
@@
 
+\subsection{Scaling of large messages}
+
+While most messages are intended to be short and quick to respond, large single messages will take appreciable time to transfer data from daemon to client over TCP.
+One common usecase where a user may wish to transfer a large message over TCP would be camera data.
+Cameras can have large arrays which contain the scientifically interesting data.
+
+While \yaq{} is flexible enough to represent such arrays, it was not specifically designed with large cameras as the primary usecase.
+As such, performance does suffer above about 1 Megapixel.
+Figure \ref{si:fig:scaling} shows the relationship between number of pixels and time for a \yaq{} message to read the array.
+Below about $2^{18}$ (~250,000) pixels, there is virtually no dependence on size, with the standard overhead of a \yaq{} message dominating the time for transfer.
+\yaq{} remains usable for up to approximately $2^{20}$ (~1,000,000) pixels.
+``Usable'' is a relative term which will depend greatly on context.
+Here we generally mean ``seems responsive to a user trying to have feedback on human timescales''.
+Applications which require high speed, high throughput cameras are unlikely to be suitable for \yaq{} even with relatively small cameras.
+This test was conducted using 32-bit integer arrays.
+
+\begin{figure}
+  \includegraphics[width=.75\columnwidth]{figures/scaling/scaling}
+	\caption{\label{si:fig:scaling} Scaling of transport time as a function of number of pixels for 32-bit integer arrays.
+Both the x- and y-axes are logarithmically scaled.
+Common camera sizes (512, 512), (1024, 1024), and (1024, 4096) are highlighted to provide context for camera users.}
+\end{figure}
+
+Common camera sizes (512, 512), (1024, 1024), and (1024, 4096) are highlighted in \ref{si:fig:scaling} to provide context.
+
+There are some strategies that could be used to mitigate the performance problems of large cameras, but they have not been implemented because large arrays remain an edge case among current \yaq{} users.
+Such strategies could include enabling compression, using transport layers other than TCP, using regions-of-interest (ROIs) to limit array sizes, and writing arrays locally to disk and providing mechanisms for retrieval out of band.
+The Avro Specification \cite{AvroSpecification} specifically provides for compression codecs in the case of storing to disk, but there is nothing preventing the RPC pipe from similarly encoding the data, provided both ends of the communication support it.
+This would allow larger arrays to be transported using fewer bytes over TCP, which would likely improve performance for large arrays where TCP transport is the bottleneck.
+While TCP is the only currently supported transport layer for \yaq{}, this limitation could be lifted in the future.
+TCP was chosen as the preferred starting point because it is ubiquitous, being available on every desired platform and implementation language.
+One alternative that would be relatively easy to support would be Unix domain sockets (UDS).
+UDS has an extremely similar interface to TCP sockets, and are handled in Python by the same standard library module.
+UDS is potentially more performant, but as the name suggests are limited to Unix-like systems (i.e. it will not work on Windows).
+ROIs are implemented on the handful of cameras currently in the \yaq{} ecosystem \cite{yaqd_andor,yaqd_pi}, however we are still in the experimentation phase and have not standardized ROI configuration into a \yaq{} trait.
+When only a subset of the total camera area is interesting, this is one way to limit array size over the interface.
+Finally \yaq{} daemons could be implemented to write arrays locally to disk, rather than transferring over the \yaq{} interface directly.
+This has not yet been implemented, and would require clients to have knowledge of how to retrieve and display the images collected, but there is nothing preventing this if throughput is a limitation.
+If such behavior became a standard feature desired for cameras, it should be encapsulated in a \yaq{} trait to provide a consistent interface.
+
+To reproduce this figure, you will need the libraries specified in ``figures/requirements.txt''.
+The scripts for generating this figure, including the \yaq{} daemon, are located in ``figures/scaling''.
+To generate the data, run the \yaq{} daemon using \texttt{python scaling.py --config config.toml} from with in the ``scaling'' folder.
+Then run \texttt{python scaling-data-gen.py}, which will communicate with the running daemon and produce a CSV written to standard out.
+The result can be saved to \texttt{scaling.csv}.
+To visualize the results, run \texttt{python plot-scaling.py}, which will read the CSV and produce the image.
+
 \section{Package size analysis}
+
+The raw data for package size analysis was collected using Tokei\cite{tokei}, a command line tool for analysing line lengths of source code.
+It includes breaking down line length into ``code'', ``comments'', and ``blank'' lines.
+The package size data was curated into \texttt{figures/lines\_histogram.txt}, which includes annotations as to the ``type'' and ``class'' for each file.
+The ``type'' annotation is one of ``stub'' (meaning that the bulk of the implementation is in another file), ``normal'' (an individual daemon with the bulk of its implementation), ``protocol'' (an implementation the provides for multiple daemons, such as those referred to by ``stub''s, or ``serial'' (the implementation of a serial protocol rather than a full fledged daemon).
+The ``class'' refers to the primary function of the daemon and is one of the following values: ``is-sensor'' for detectors, ``has-position'' for setable hardware, ``serial'' (identical to ``type'' annotation), and ``other'' for daemons which do not fit in the above categories.
+
+The plotting script \texttt{figures/lines\_histogram.py} can be used to generate the figure from this text file.
+The figure uses the ``code'' lines to generate a histogram of file sizes in the \yaq{} python implementations.
+This is a measurable proxy for the amount of work that implementing an individual \yaq{} daemon entails, though of course cannot capture the work involved in learning the lower level interface.
 

### 2023-02-05 paper.tex

@@
 Protocols which implement a trait must support all of the messages from the trait.
+Traits are compositional: a given protocol may implement several traits at once.
+For example, a protocol representing a monochromator with several gratings might implement both ``\texttt{has-position}`` and ``\texttt{has-turret}``, where the latter defines special messages for choosing which grating is used.
 Importantly, specific protocols can also implement arbitrary additional messages that are not defined by any trait.
+
+We have carefully defined \yaq{} traits to maximize interoperability while not excluding hardware with unusual features.
+New traits can be discussed by the community through the yaq enhancement proposal system.
+We are somewhat conservative about adding new traits, and it's crucial that multiple hardware interface examples be considered when defining traits.
+Configuration that is unique to a single interface is best handled through protocol-specific messages.
+For example, our protocol for interfacing with data acquisition cards from Gage provides a message for setting the segment count.
+Unique messages are avaliable through script-based and graphical clients, but usage of such messages in scripts naturally limits portability.
 

### 2023-02-06 paper.tex
Add sentence about high bandwidth applications to main text
Direct comparison to EPICS/TANGO (#40)

@@
 In our experience, the \yaq{} interface is responsive enough to give good performace when orchestrating tens of motors using a singly-threaded client.
+There is a limit where using TCP becomes a bottleneck for large resonses, such as those from cameras or other high throughput sensors.
+In our experience, this limit is approximately 1 megapixel.
+For larger sensors, a pure \yaq{} approach may not be appropriate.
 There is a longer discussion of timing and order-of-operations control details using \yaq{} in the Supplementary Material.
@@
 
+We have designed \yaq{} to be easier to deploy and maintain when compared to EPICS and TANGO.
+\yaq{} daemons are Python packages that can be installed on any platform using pip or conda.
+There is no need for centralized management servers or databases when using \yaq{}.
+The Avro RPC standard unambiguously describes the message signatures for any protocol, so users are always aware of the capabilities their hardware supports.
+Traits enable interchangability where possible.
+\yaq{} can easily be used alongside other experiment control software.
+We provide tooling (\texttt{yaqd-control}) which makes it easy to configure, list, and manage daemons as background services on Windows, MacOS, or Linux (See SI for more details).
+
 \yaq{} is open source software.

### 2023-02-06 paper.tex
preemptive

@@
 Each individual communication between client and daemon involves one message being requested by the client and the response returned from the daemon.
+If multiple messages are sent simultaneously, the daemon processes one message at a time.
+The protocol is preemptive: in the case of a conflict, the daemon will obey the last message to arrive.
 Each daemon supports a collection of messages for its unique functionality, called the ``protocol''.
@@
 New traits can be discussed by the community through the yaq enhancement proposal system.
-We are somewhat conservative about adding new traits, and it's crucial that multiple hardware interface examples be considered when defining traits.
+It's crucial that multiple hardware interface examples be considered when defining traits.
 Configuration that is unique to a single interface is best handled through protocol-specific messages.

### 2023-02-06 paper-as-submitted.tex
diff

--- 2023-02-05 paper.tex
@@
 \yaq{} encourages a highly modular approach to experimental software development which is well suited to address the experimental flexibility needs of complex instruments.
-\yaq{} is designed to overcome hardware communication barriers that challenge typical experimental software.
+\yaq{} is designed to overcome hardware communication barriers that are insurmountable with typical experimental software.
 A large number of hardware peripherals are already supported, with tooling available to expand support.
@@
 We have built this standard to be self-describing, portable, and reusable wherever possible.
-Via the \yaq{} interface, we provide easy-to-use hardware support for scientists to use with their choice of experiment orchestration software.
+Our primary goal has been to make an interface which simplifies orchestration software development as much as possible.
 
@@
 Protocols which implement a trait must support all of the messages from the trait.
-Traits are compositional: a given protocol may implement several traits at once.
-For example, a protocol representing a monochromator with several gratings might implement both ``\texttt{has-position}`` and ``\texttt{has-turret}``, where the latter defines special messages for choosing which grating is used.
 Importantly, specific protocols can also implement arbitrary additional messages that are not defined by any trait.
 
-We have carefully defined \yaq{} traits to maximize interoperability while not excluding hardware with unusual features.
-New traits can be discussed by the community through the yaq enhancement proposal system.
-We are somewhat conservative about adding new traits, and it's crucial that multiple hardware interface examples be considered when defining traits.
-Configuration that is unique to a single interface is best handled through protocol-specific messages.
-For example, our protocol for interfacing with data acquisition cards from Gage provides a message for setting the segment count.
-Unique messages are avaliable through script-based and graphical clients, but usage of such messages in scripts naturally limits portability.
-
-We now describe how the \yaq{} architecture addresses the three hardware interface barriers described at the beginning of this section.
 The first hardware interface barrier: Multiple incompatible interfaces are used to communicate with each component of the system.
@@
 
-The third hardware interface barrier: Some hardware interfaces are blocking or slow.
-Experimental orchestration often means doing several things at once, \textit{e.g.} simultaneously moving several motors.
-To accomplish this task performantly, monolithic orchestration software often necessitates separate threads for each component hardware interface, a fragile pattern in which small mistakes become both critical and elusive errors.
-When writing orchestration software using \yaq{}, slow hardware interfaces are replaced by fast Avro-RPC.
-In our experience, the \yaq{} interface is responsive enough to give good performace when orchestrating tens of motors using a singly-threaded client.
-There is a longer discussion of timing and order-of-operations control details using \yaq{} in the Supplementary Material.
+The third hardware interface barrier: Some hardware interfaces are blocking.
+Monolithic orchestration software often necessitates separate threads for each component hardware interface, a fragile pattern in which small mistakes become both critical and elusive errors.
+In \yaq{}, each daemon is only responsible for a single hardware interface.
+%Each \yaq{} daemon runs as a separate process (possibly not even on the same machine) and thus many of the hard concurrency problems with communicating with hardware are mitigated.
+It is expected that each message call over the \yaq{} interface will return rapidly, ensuring that client applications are not blocked for extended periods of time.
+This principle applies even when the message starts an action that might take several seconds to complete, such as homing a motor or initiating a measurement for a sensor.
+In these instances, the initial message simply starts the action and returns, with a separate message provided to retrieve results when they exist.
+In order to know how long to wait, the ``is-daemon`` trait provides a message called ``is\_busy'', which returns "true" while the long running action is not complete, and "false" once it is finished.
+Additionally, multiple clients can communicate with the same daemon simultaneously.
+A complex instrument may involve multiple operators watching sensor data in real time, while one program is orchestrating the hardware and recording the data.
 
@@
 
-At the bottom, we graphically represent the integration between \yaq{} and the Bluesky project\cite{AllanDanielB2019a}.
+At the bottom, we show several features associated with the integration between \yaq{} and the Bluesky project\cite{AllanDanielB2019a}.
 Bluesky provides a powerful orchestration layer for conducting and recording data for a wide variety of experimental procedures.
-Bluesky has no built-in hardware interface support, instead relying on packages to create hardware-interfaces that are compatible with their Hardware Protocol defintions.
-\texttt{yaqc-bluesky}\footnote{\texttt{yaqc-bluesky} available on PyPI at https://pypi.org/project/yaqc-bluesky/} is a specialized client which adapts any \yaq{} protocol into Bluesky's Hardware Protocol.
+Shown in Figure \ref{fig:foundation}, we see one such GUI for parameterizing a Bluesky procedure and a representation of the resultant data.
+\texttt{yaqc-bluesky}\footnote{\texttt{yaqc-bluesky} available on PyPI at https://pypi.org/project/yaqc-bluesky/} provides a bridge to the Bluesky ecosystem.
 Similar translation layers could be built for a variety of orchestration layers such as PyMoDAQ\cite{WeberSebastien2021a}, Instrumental\cite{Bogdanowicz2022}, or TRSpectrometer\cite{trspectrometer}.
@@
 Even the most complicated daemons are implemented in about 300 lines.
-Implementing \yaq{} daemons using Python is enabled through our own tooling mentioned above and the large and growing ecosystem of hardware interface libraries that Python now provides\cite{pyserial, pyusb, pyvisa, pymodbus}.
+Implementing \yaq{} daemons using Python is enabled through our own tooling mentioned above and the large and growing ecosystem of hardware interface tooling libraries that Python now provides\cite{pyserial, pyusb, pyvisa, pymodbus}.
 In our experience, the process of creating a new daemon involves about a day of work after mastering communication with the hardware.
@@
 
-Software built by scientists often has incomplete and inadequate documentation. \cite{SegalJudith2007a}
+Software documentation is famously difficult and thankless work.
 \yaq{} attempts to automate daemon documentation as much as possible.

### 2023-02-06 SI.tex
Add yaqd-control user guide to SI

@@
 \usepackage{color}
+
+\RequirePackage{etoolbox}
+\RequirePackage{pmboxdraw}
+\RequirePackage[section]{minted}
+\usemintedstyle{default}
+\setminted[python]{linenos=false, frame=none}
+
+\definecolor{bg}{rgb}{0.95, 0.95, 0.95}
+
+\RequirePackage[many]{tcolorbox}
+\tcbuselibrary{minted}
 
@@
 \newcommand{\RomanNumeral}[1]{\textrm{\uppercase\expandafter{\romannumeral #1\relax}}}
+
+\newcommand\yaq{\texttt{yaq}}
 
@@
 \usepackage[all]{hypcap}  % helps hyperref work properly
+
+\NewTCBListing[number within=chapter, auto counter]{codefragment}{mo}{%
+  %colback=bg,
+  boxrule=1pt,
+  %colframe=bg,
+  arc=0pt,
+  shadow=false,
+  new/use counter=equation,
+  boxsep=1ex, top=0pt, left=0pt, right=0pt, bottom=0pt,
+  comment={\hfill(\arabic{chapter}.\arabic{equation})},
+  listing outside comment,
+  righthand width=3em,
+  sidebyside gap=0pt,
+  minted language=#1,
+  %before skip =-0.5\baselinestretch,
+  %after skip=2\baselinestretch,
+}
+
+\BeforeBeginEnvironment{codefragment}{\begin{singlespace}\stepcounter{equation}}
+\AfterEndEnvironment{codefragment}{\end{singlespace}}
+
+\newmintinline[bash]{bash}{bgcolor=bg}
+\newmintinline[python]{python}{bgcolor=bg}
 
@@
 
+\clearpage
+
+\section{\texttt{yaqd-control}: tooling for managing \yaq{} daemons}
+
+\texttt{yaqd-control} is a command line utility provided by the \yaq{} project to manage \yaq{} daemons.
+The primary purpose of the tool is to maintain a list of active daemons, provide access to their configuration, and their running state.
+This includes querying the active state of known daemons, as well as starting daemons running as backgound services.
+
+\subsection{installation}
+
+yaqd-control can be installed via
+PyPI\cite{yaqd-control} or
+conda-forge\cite{yaqd-control-conda}.
+
+\begin{codefragment}{bash}
+$ pip install yaqd-control
+\end{codefragment}
+
+\begin{codefragment}{bash}
+$ conda config --add channels conda-forge
+$ conda install yaqd-control
+\end{codefragment}
+
+\subsection{Usage}
+
+yaqd-control is a command line application.
+
+Help: learn more, right from your terminal.
+
+\begin{codefragment}{bash}
+$ yaqd --help
+Usage: yaqd [OPTIONS] COMMAND [ARGS]...
+
+Options:
+  --help  Show this message and exit.
+
+Commands:
+  clear-cache
+  disable
+  edit-config
+  enable
+  list
+  reload
+  restart
+  scan
+  start
+  status
+  stop
+\end{codefragment}
+
+Try \texttt{yaqd\ \ -\/-help} to learn more about a particular command.
+
+\paragraph{the cache}\label{the-cache}
+
+yaqd-control keeps track of known daemons, referred to as the cache
+
+Status: yaqd-control can quickly show you the status of all daemons in
+yaqd-control's cache. This is usually the most used subcommand, as it
+gives a quick overview of the system, which daemons are offline, and
+which are currently busy.
+
+\begin{codefragment}{bash}
+$ yaqd status
++-----------+-------+--------------------------+------+---------+-------+
+| host      | port  | kind                     | name | status  | busy  |
++-----------+-------+--------------------------+------+---------+-------+
+| 127.0.0.1 | 38202 | system-monitor           | foo  | online  | False |
+| 127.0.0.1 | 39054 | fake-continuous-hardware | bar  | online  | True  |
+| 127.0.0.1 | 39055 | fake-continuous-hardware | baz  | online  | False |
+| 127.0.0.1 | 39056 | fake-continuous-hardware | spam | offline | ?     |
+| 127.0.0.1 | 37067 | fake-discrete-hardware   | ham  | online  | False |
+| 127.0.0.1 | 37066 | fake-discrete-hardware   | eggs | online  | False |
++-----------+-------+--------------------------+------+---------+-------+
+\end{codefragment}
+
+List: this is essentially the same as \texttt{status} except that it
+does not attempt to contact the daemons, so it does not give you
+additional context. List supports a flag \texttt{--format} which accepts "json",
+"toml", "prettytable", or "happi".
+
+\begin{codefragment}{bash}
+$ yaqd list
++-----------+-------+--------------------------+------+
+| host      | port  | kind                     | name |
++-----------+-------+--------------------------+------+
+| 127.0.0.1 | 38202 | system-monitor           | foo  |
+| 127.0.0.1 | 39054 | fake-continuous-hardware | bar  |
+| 127.0.0.1 | 39055 | fake-continuous-hardware | baz  |
+| 127.0.0.1 | 39056 | fake-continuous-hardware | spam |
+| 127.0.0.1 | 37067 | fake-discrete-hardware   | ham  |
+| 127.0.0.1 | 37066 | fake-discrete-hardware   | eggs |
++-----------+-------+--------------------------+------+
+\end{codefragment}
+
+Scan: Scanning allows you to add currently running daemons to the cache.
+
+\begin{codefragment}{bash}
+$ yaqd scan
+scanning host 127.0.0.1 from 36000 to 39999...
+...saw unchanged daemon fake-discrete-hardware:eggs on port 37066
+...saw unchanged daemon fake-discrete-hardware:ham on port 37067
+...found new daemon system-monitor:foo on port 38202
+...found new daemon fake-continuous-hardware:bar on port 39054
+...saw unchanged daemon fake-continuous-hardware:baz on port 39055
+...known daemon fake-continuous-hardware:spam on port 39056 not responding
+...done!
+\end{codefragment}
+
+Scan has some additional options, passed as flags on the command line,
+which allow you to change the default scan range and host (for remotely
+accessed daemons):
+
+\begin{codefragment}{bash}
+$ yaqd scan --help
+Usage: yaqd scan [OPTIONS]
+
+Options:
+  --host TEXT      Host to scan.
+  --start INTEGER  Scan starting point.
+  --stop INTEGER   Scan stopping point.
+  --help           Show this message and exit.
+\end{codefragment}
+
+Edit Config: yaqd-control provides an easy way to edit the default
+config file location for a daemon kind. This uses your default editor
+(\texttt{EDITOR} environment variable), and defaults to \texttt{notepad.exe} on
+Windows, and \texttt{vi} on other platforms. Using yaqd-control to edit
+config files means that you do not need to know the default location.
+Additionally, it does some basic validity checks (that the toml parses
+and that each daemon section has the \texttt{port} keyword). If an error
+is found, you are prompted to re-edit the file. Daemons from the config
+file are added to the cache. You may pass multiple daemon kinds, which
+will be opened in succession.
+
+\begin{codefragment}{bash}
+$ yaqd edit-config fake-continuous-hardware system-monitor
+\end{codefragment}
+
+Clear Cache: Note that this is a destructive action.
+\texttt{clear-cache} deletes all daemons from the cache (thus
+\texttt{list} and \texttt{status} will give empty tables) There is no
+user feedback.
+
+\begin{codefragment}{bash}
+$ yaqd clear-cache
+$ yaqd status
++------+------+------+------+--------+------+
+| host | port | kind | name | status | busy |
++------+------+------+------+--------+------+
++------+------+------+------+--------+------+
+\end{codefragment}
+
+\paragraph{Running in the background}\label{running-in-the-background}
+Each of the commands in this section can take multiple daemon kinds.
+Enable: by enabling a daemon, you allow the operating system to manage
+that daemon in the background. An enabled daemon will always start again
+when you restart your computer. Enabling is required for the rest of the
+commands in this section to work as expected. After enabling, it's
+typical to start the daemon as well, this does not happen automatically.
+Enablement works in slightly different ways on different platforms, but
+the commands are the same (don't worry if the password prompts are
+different). Currently supported platforms are Linux (systemd), MacOS
+(launchd) and Windows (via NSSM, bundled with the distribution).
+
+\begin{codefragment}{bash}
+$ yaqd enable system-monitor
+[sudo] password for scipy2020:
+==== AUTHENTICATING FOR org.freedesktop.systemd1.manage-unit-files ===
+Authentication is required to manage system service or unit files.
+Password:
+==== AUTHENTICATION COMPLETE ===
+\end{codefragment}
+
+Disable: this is the inverse operation to enable, which makes it so that
+the daemon does not start on reboot. This does not affect the running
+daemon.
+
+\begin{codefragment}{bash}
+$ yaqd disable system-monitor
+==== AUTHENTICATING FOR org.freedesktop.systemd1.manage-unit-files ===
+Authentication is required to manage system service or unit files.
+Password:
+==== AUTHENTICATION COMPLETE ===
+Removed /etc/systemd/system/multi-user.target.wants/yaqd-system-monitor.service.
+\end{codefragment}
+
+Start: This starts the daemon running in the background immediately. It
+must have been enabled to run in the background using this command.
+
+\begin{codefragment}{bash}
+$ yaqd start system-monitor
+==== AUTHENTICATING FOR org.freedesktop.systemd1.manage-units ===
+Authentication is required to start 'yaqd-system-monitor.service'.
+Password:
+==== AUTHENTICATION COMPLETE ===
+\end{codefragment}
+
+Stop: This stops the daemon running in the background immediately. It
+must have been running in the background using yaqd-control (either on
+startup via enable or via the start command above).
+
+\begin{codefragment}{bash}
+$ yaqd stop system-monitor
+==== AUTHENTICATING FOR org.freedesktop.systemd1.manage-units ===
+Authentication is required to stop 'yaqd-system-monitor.service'.
+Password:
+==== AUTHENTICATION COMPLETE ===
+\end{codefragment}
+
+Restart/Reload: This stops (if running) and restarts the daemon running
+in the background immediately. Reload is slightly different in that it
+signals to the daemon to reload its configuration rather than completely
+restart, but effectively it is the same as restart (and is a pure alias
+where such a signal is not supported). It must have been enabled to run
+in the background using this command.
+
+\begin{codefragment}{bash}
+$ yaqd restart system-monitor
+==== AUTHENTICATING FOR org.freedesktop.systemd1.manage-units ===
+Authentication is required to restart 'yaqd-system-monitor.service'.
+Password:
+==== AUTHENTICATION COMPLETE ===
+\end{codefragment}
+
+\clearpage
+
 \section{Package size analysis}

### 2023-02-06 build.sh

@@ -4,7 +4,7 @@ bibtex paper.aux
 pdflatex --interaction=nonstopmode paper.tex
 pdflatex --interaction=nonstopmode paper.tex
 #
-pdflatex --interaction=nonstopmode SI.tex
+pdflatex --interaction=nonstopmode -shell-escape SI.tex
 bibtex SI.aux
-pdflatex --interaction=nonstopmode SI.tex
-pdflatex --interaction=nonstopmode SI.tex
+pdflatex --interaction=nonstopmode -shell-escape SI.tex
+pdflatex --interaction=nonstopmode -shell-escape SI.tex

### 2023-02-07 SI.tex
orchestration section (#49)

@@
 
+\section{Orchestration Examples}
+
+Here we elaborate with detailed information about what orchestration looks like for \yaq{}.
+As we discussed in Section 3 of the manuscript, \yaq{} is designed to accommodate a wide variety of orchestration tools and approaches.
+
+\subsection{Raster Script}
+
+For those unfamiliar with \yaq{}, the best orchestration example is a simple self-contained script.
+
+\begin{codefragment}{Python}
+import time
+import yaqc
+motor1 = yaqc.Client(port=38000)
+motor2 = yaqc.Client(port=38001)
+sensor = yaqc.Client(port=38002)
+data = []
+for m1 in range(-10, 10, 1):
+    for m2 in range(0, 300, 5):
+        motor1.set_position(m1)
+        motor2.set_position(m2)
+        for c in [motor1, motor2]:
+            while c.busy():
+                time.sleep(0.001)
+        sensor.measure()
+        while sensor.busy():
+            time.sleep(0.001)
+        reading = dict()
+        reading["timestamp"] = time.time()
+        reading["motor1"] = motor1.get_position()
+        reading["motor2"] = motor2.get_position()
+        reading.update(sensor.get_measured())
+        data.append(reading)
+\end{codefragment}
+
+Here we are doing a two dimensional motor scan while collecting data from a sensor.
+Motor one is stepping from -10 to 10, and motor two is scanning from 0 to 300.
+Two examples of polling while not busy are used: the first to ensure that motors have stopped moving before sensors acquire, and the second to ensure that sensors stop collecting before the next motor motion happens.
+This ``tick tock'' moving multiple motors, collecting sensor data, and repeating is a common pattern.
+
+We believe that Python and the Scientific Python ecosystem is a powerfully simple tool for defining experimental procedures.
+Lightweight scripts such as these are encouraged by the \yaq{} project.
+
+\subsection{Landis}
+
+We discussed the Landis Group and their WiQK reactor in Section 3 of the manuscript.
+The Landis Group has defined several procedures for their flow reactor.
+These procedures are highly idiosyncratic and are based on the particular tubing lengths and the flow topology of the reactor.
+Each procedure involves articulating solenoid valves while injecting and withdrawing syringe pumps with a particular timing.
+The Landis Group has written Python functions that drive this hardware using \yaq{}.
+The functions parameterize the procedures in terms of chemically interesting variables like flow rate
+and reaction time.
+
+WiQK procedures and nascent graphical user interface can be found on GitHub at \\ \url{https://github.com/uw-madison-chem-shops/yaqc-wiqk}.
+
+\subsection{Wright \& Bluesky}
+
 \section{Timing and interface responsiveness}
 
-Each message call over the \yaq{} interface returns rapidly,
-ensuring that client applications are not blocked. Some mes-
-sages initiate long-running operations, e.g. motor motion and
-sensor measurement. Separate messages are provided to re-
-trieve results from e.g. sensor measurement.
+Each message call over the \yaq{} interface returns rapidly, ensuring that client applications are not blocked.
+Some messages initiate long-running operations, \emph{e.g.} motor motion and sensor measurement.
+Separate messages are provided to retrieve results from \emph{e.g.} sensor measurement.
 

### 2023-02-09 paper.tex
remove reference to gage (#51)
pymeasure citation, use word software

@@
 Configuration that is unique to a single interface is best handled through protocol-specific messages.
-For example, our protocol for interfacing with data acquisition cards from Gage provides a message for setting the segment count.
+For example, our protocol for interfacing with a specific data acquisition card provides a message for setting the segment count.
 Unique messages are avaliable through script-based and graphical clients, but usage of such messages in scripts naturally limits portability.
@@
 \texttt{yaqc-bluesky}\footnote{\texttt{yaqc-bluesky} available on PyPI at https://pypi.org/project/yaqc-bluesky/} is a specialized client which adapts any \yaq{} protocol into Bluesky's Hardware Protocol.
-Similar translation layers could be built for a variety of orchestration layers such as PyMoDAQ\cite{WeberSebastien2021a}, Instrumental\cite{Bogdanowicz2022}, or TRSpectrometer\cite{trspectrometer}.
+Similar translation layers could be built for a variety of orchestration software such as PyMoDAQ\cite{WeberSebastien2021a}, PyMeasure\cite{pymeasure}, or TRSpectrometer\cite{trspectrometer}.
 

### 2023-02-09 SI.tex

@@
 \setminted[python]{linenos=false, frame=none}
+\usepackage{lscape}
 
@@
 \newcommand\yaq{\texttt{yaq}}
+\newcommand\yaqccmds{\texttt{yaqc-cmds}}
 
@@
 \subsection{Wright \& Bluesky}
+
+In the main body of the window, a table represents the current state of the queue and history.
+More recently enqueued items appear towards the top of the table, older items towards the bottom.
+The items waiting in queue, thus, appear at the top, and the items in the history appear at the bottom.
+The currently running item (if it exists) appears at the boundary between the queue and the history.
+The enqueued and running items have descriptions shown in bright white, while the completed items appear with gray descriptions.
+The right most column indicates the position in the queue, given as a zero-based index.
+Items can be reordered by editing the value in that column.
+On left, there are two columns with buttons: remove and load.
+Items in the queue (but not the history) can be removed individually with the red ``REMOVE'' buttons.
+All items can be loaded into the sidebar where the parameters can be modified and a similar (or identical) item enqueued.
+The other information provided include the plan name, the status (enqueued, RUNNING, completed, failed, aborted, stopped or unknown) and the description field, which lists the arguments passed to the plan.
+Hovering over the description shows the full JSON of the item (see Figure \ref{acq:fig:hover_text}, including an error message if applicable.
+The description has a contextual menu accessed via right clicking which provides additional options as shown in Figure \ref{acq:fig:right_click}.
+Most of the items in the context menu simply copy information to the clipboard: The full JSON as shown by the hover text, the item \Gls{UUID} which identifies the item in the queue, and the run \Gls{UUID} which identifies the data that is collected by the scan.
+
+The plot tab (Figure \ref{acq:fig:plot_tab} presents a live representation of recent data collected for the current plan.
+The primary window is a PyQtGraph\cite{pyqtgraph} plot showing the most recent five slices of collected data.
+The current slice is the brightest cyan slice, with the previous slices shown as duller colors progressively until the last one which is gray.
+At the top, the currently plotted channel name and most recent collected value are shown in large font size.
+The plotting does handle array detectors, for which the top number shown is the maximum value of the most recent array collected.
+At present, cameras and other higher dimensional sensors are not able to be plotted using the live plot tab.
+On the right hand side, there are controls to adjust the plot.
+At the top, a selector allows for choosing which channel to plot.
+The second selector determines which values appear along the X-axis of the graph.
+In particular, to switch between a scalar channel and an array detected channel, both selectors will need to be adjusted, as the appropriate axes differ.
+Additionally, there is a selector which allows for changing the units of the X-axis.
+Finally, there is a Scan Index indicator, which indicates which indexed pixel was the most recent collected.
+
+\clearpage
+
+\begin{landscape}
+\begin{figure}
+\includegraphics[width=9in]{"figures/yaqc-cmds.png"}
+\caption{The main queue window of \yaqccmds{}.}
+\label{acq:fig:yaqccmds}
+\end{figure}
+\end{landscape}
+
+\clearpage
+
+\begin{figure}
+\includegraphics[width=3in]{"figures/grid_scan.png"}
+\caption{The user interface for enqueueing a grid scan plan.}
+\label{acq:fig:grid_scan}
+\end{figure}
+
+\clearpage
+
+\begin{landscape}
+\begin{figure}
+\includegraphics[width=9in]{"figures/plot_tab.png"}
+\caption{\blueskycmds with the Plot tab open.}
+\label{acq:fig:grid_scan}
+\end{figure}
+\end{landscape}
+
+\clearpage
 

### 2023-02-09 SI.tex
History of yaq section

@@
 \NewTCBListing[number within=chapter, auto counter]{codefragment}{mo}{%
-  %colback=bg,
   boxrule=1pt,
-  %colframe=bg,
   arc=0pt,
   shadow=false,
-  new/use counter=equation,
   boxsep=1ex, top=0pt, left=0pt, right=0pt, bottom=0pt,
-  comment={\hfill(\arabic{chapter}.\arabic{equation})},
   listing outside comment,
-  righthand width=3em,
+  righthand width=0pt,
   sidebyside gap=0pt,
   minted language=#1,
-  %before skip =-0.5\baselinestretch,
-  %after skip=2\baselinestretch,
 }
@@
 As we discussed in Section 3 of the manuscript, \yaq{} is designed to accommodate a wide variety of orchestration tools and approaches.
+
+\subsection{\texttt{is\_busy} for orchestrating order of operations}
+
+Each message call over the \yaq{} interface returns rapidly, ensuring that client applications are not blocked.
+Some messages initiate long-running operations, \emph{e.g.} motor motion and sensor measurement.
+Separate messages are provided to retrieve results from \emph{e.g.} sensor measurement.
+
+In order to know how long to wait, all \yaq{} daemons provide a message called
+“is\_busy”, which returns "true" while the long running action
+is not complete, and "false" once it is finished. Additionally,
+multiple clients can communicate with the same daemon si-
+multaneously. A complex instrument may involve multiple
+operators watching sensor data in real time, while one pro-
+gram is orchestrating the hardware and recording the data.
 
@@
 
-\section{Timing and interface responsiveness}
+\section{\yaq{} transport layer}
 
-Each message call over the \yaq{} interface returns rapidly, ensuring that client applications are not blocked.
-Some messages initiate long-running operations, \emph{e.g.} motor motion and sensor measurement.
-Separate messages are provided to retrieve results from \emph{e.g.} sensor measurement.
+\subsection{History of \yaq{} protocol format}
 
-\subsection{\texttt{is\_busy} for orchestrating order of operations}
+In the first days of \yaq{}, the message format was entirely made up of JSON strings, forming a generic Remote Procedure Call (RPC) framework.
+There was no pre-formed protocol, though there were standardized messages to query for all allowed message names, signatures, and documentation.
 
-In order to know how long to wait, all \yaq{} daemons provide a message called
-“is\_busy”, which returns "true" while the long running action
-is not complete, and "false" once it is finished. Additionally,
-multiple clients can communicate with the same daemon si-
-multaneously. A complex instrument may involve multiple
-operators watching sensor data in real time, while one pro-
-gram is orchestrating the hardware and recording the data.
+Only after having a largely functional implementation, we descovered a community standard, JSON-RPC \cite{jsonrpc}.
+While the exact format was not identical, almost all of the ideas we had built into our home baked JSON-based RPC were easy to translate into this standard format.
+Instead of relying on our own esoteric format, we converted all \yaq{} daemons and clients to use JSON-RPC.
+
+As \yaq{} gained hardware support and usage, it became evident that JSON was not an appropriate serialization format, especially when it comes to large binary data like images from cameras.
+As such, we transitioned to using an RPC based on JSON-RPC, but using Msgpack \cite{msgpack} for serialization.
+Msgpack is a format that can encode a strict superset of JSON messages, so swapping existing daemons required only changes to the core implementation and the client program (i.e.
+no changes to individual daemon implementations).
+Along with using Msgpack, we defined an extension type to represent NDArrays.
+
+The last major change to \yaq{}'s protocol format was the adoption of Apache Avro-RPC\cite{AvroSpecification}.
+Avro is similarly terse like Msgpack for serializing messages, while having a well defined specification for defining an RPC schema: a static definition of available messages, including their signatures and documentation.
+This statically defined schema pairs well with \yaq{}'s trait system.
+Daemon authors provide a short, minimal TOML file which describes the implemented traits and unique messages for a particular daemon.
+This then gets transformed by an automated tool to a full Avro-RPC JSON file describing the protocol.
+Avro is then used to encode individual messages with minimal overhead.
 
@@
 
-Common camera sizes (512, 512), (1024, 1024), and (1024, 4096) are highlighted in \ref{si:fig:scaling} to provide context.
-
 There are some strategies that could be used to mitigate the performance problems of large cameras, but they have not been implemented because large arrays remain an edge case among current \yaq{} users.
 Such strategies could include enabling compression, using transport layers other than TCP, using regions-of-interest (ROIs) to limit array sizes, and writing arrays locally to disk and providing mechanisms for retrieval out of band.
+
 The Avro Specification \cite{AvroSpecification} specifically provides for compression codecs in the case of storing to disk, but there is nothing preventing the RPC pipe from similarly encoding the data, provided both ends of the communication support it.
-This would allow larger arrays to be transported using fewer bytes over TCP, which would likely improve performance for large arrays where TCP transport is the bottleneck.
+EPICS provides a similar NDArray data type, which has individual image level compression configureable \cite{normative_types}.
+Compression would allow larger arrays to be transported using fewer bytes over TCP, which would likely improve performance for large arrays where TCP transport is the bottleneck.
 While TCP is the only currently supported transport layer for \yaq{}, this limitation could be lifted in the future.
+
 TCP was chosen as the preferred starting point because it is ubiquitous, being available on every desired platform and implementation language.
@@
 UDS is potentially more performant, but as the name suggests are limited to Unix-like systems (i.e. it will not work on Windows).
+
 ROIs are implemented on the handful of cameras currently in the \yaq{} ecosystem \cite{yaqd_andor,yaqd_pi}, however we are still in the experimentation phase and have not standardized ROI configuration into a \yaq{} trait.
 When only a subset of the total camera area is interesting, this is one way to limit array size over the interface.
+
 Finally \yaq{} daemons could be implemented to write arrays locally to disk, rather than transferring over the \yaq{} interface directly.
-This has not yet been implemented, and would require clients to have knowledge of how to retrieve and display the images collected, but there is nothing preventing this if throughput is a limitation.
+Anecdotally, many users of Bluesky, EPICS, and related technologies use a similar strategy for large image data.
+This has not yet been implemented for \ya{}, and would require clients to have knowledge of how to retrieve and display the images collected, but there is nothing preventing this if throughput is a limitation.
 If such behavior became a standard feature desired for cameras, it should be encapsulated in a \yaq{} trait to provide a consistent interface.

### 2023-02-09 paper.tex
prior art portability
message supersed paragraph

@@
 Messages make up the lowest level functionality of the \yaq{} interface.
+Each daemon supports a collection of messages for its unique functionality, called the ``protocol''.
+When a client first connects to a daemon over TCP/IP, the daemon provides a complete description of its own protocol.
+We therefore refer to the daemons as ``self-describing''.
+
 Each individual communication between client and daemon involves one message being requested by the client and the response returned from the daemon.
@@
 The protocol is preemptive: in the case of a conflict, the daemon will obey the last message to arrive.
-Each daemon supports a collection of messages for its unique functionality, called the ``protocol''.
-When a client first connects to a daemon over TCP/IP, the daemon provides a complete description of its own protocol.
-We therefore refer to the daemons as ``self-describing''.
+By being completely open to messages from multiple clients, \yaq{} makes it possible to accidentally supersede a message.
+This has not become a problem for our usage where instruments are controlled by single scientists who can clearly see all of the clients interacting with their system.
+Typically there is just one client sending `set\_position` and many clients polling `get\_position` to record data or display real-time information.
+Users may choose to implement access control systems, but we do not foresee building access control into the \yaq{} protocol.
+This is similar to how access control works in practice at some large facilities.
 
@@
 Often, this separation allows initial hardware enablement work to be done on a researchers personal machine before the new hardware peripheral is installed in the instrument.
+Separability, portability, and distributed development are advantages common to many open-source hardware interface projects.
 Third, when developing trait-compliant protocols, it becomes easy to design and fully test your hardware interface.

### 2023-02-09 SI.tex
Update section about wright group and bluesky

@@
 
-In the main body of the window, a table represents the current state of the queue and history.
+The Wright Group uses a purpose-build graphical program to perform data acquisition.
+All of the hardware used is controlled via \yaq{}.
+Data is collected using Bluesky, which runs in a background process.
+
+The graphical program includes a queue of acquisition procedures (plans, in Bluesky terminology).
+In the main body of the queue window (Figure \ref{acq:fig:queue_tab}), a table represents the current state of the queue and history.
 More recently enqueued items appear towards the top of the table, older items towards the bottom.
@@
 The other information provided include the plan name, the status (enqueued, RUNNING, completed, failed, aborted, stopped or unknown) and the description field, which lists the arguments passed to the plan.
-Hovering over the description shows the full JSON of the item (see Figure \ref{acq:fig:hover_text}, including an error message if applicable.
-The description has a contextual menu accessed via right clicking which provides additional options as shown in Figure \ref{acq:fig:right_click}.
-Most of the items in the context menu simply copy information to the clipboard: The full JSON as shown by the hover text, the item \Gls{UUID} which identifies the item in the queue, and the run \Gls{UUID} which identifies the data that is collected by the scan.
+Hovering over the description shows the full JSON of the item, including an error message if applicable.
 
-The plot tab (Figure \ref{acq:fig:plot_tab} presents a live representation of recent data collected for the current plan.
+Along the left side of the window, there is a panel that allows enqueuing new items.
+Figure \ref{acq:fig:grid_scan} shows the configuration panel for the most popular acquisition procedure, \texttt{grid\_scan}.
+This plan allows an arbitrary number of dimensions to be scanned against one another, selecting the hardware to be scanned and the detectors to read.
+
+The plot tab (Figure \ref{acq:fig:plot_tab}) presents a live representation of recent data collected for the current plan.
 The primary window is a PyQtGraph\cite{pyqtgraph} plot showing the most recent five slices of collected data.
@@
 The second selector determines which values appear along the X-axis of the graph.
-In particular, to switch between a scalar channel and an array detected channel, both selectors will need to be adjusted, as the appropriate axes differ.
 Additionally, there is a selector which allows for changing the units of the X-axis.
@@
 \begin{figure}
-\includegraphics[width=9in]{"figures/yaqc-cmds.png"}
-\caption{The main queue window of \yaqccmds{}.}
-\label{acq:fig:yaqccmds}
+\includegraphics[width=9in]{"figures/queue_tab.png"}
+	\caption{The main queue window of \texttt{bluesky-cmds}.}
+\label{acq:fig:queue_tab}
 \end{figure}
@@
 \includegraphics[width=9in]{"figures/plot_tab.png"}
-\caption{\blueskycmds with the Plot tab open.}
-\label{acq:fig:grid_scan}
+	\caption{\texttt{bluesky-cmds} with the Plot tab open.}
+\label{acq:fig:plot_tab}
 \end{figure}

### 2023-02-10 SI.tex
SI cleanup

@@
 
-\subsection{\texttt{is\_busy} for orchestrating order of operations}
+\subsection{\texttt{is\_busy} for ensuring order of operations}
 
@@
 
-In order to know how long to wait, all \yaq{} daemons provide a message called
-“is\_busy”, which returns "true" while the long running action
-is not complete, and "false" once it is finished. Additionally,
-multiple clients can communicate with the same daemon si-
-multaneously. A complex instrument may involve multiple
-operators watching sensor data in real time, while one pro-
-gram is orchestrating the hardware and recording the data.
+In order to know how long to wait, all \yaq{} daemons provide a message called “is\_busy”, which returns "true" while the long running action is not complete, and "false" once it is finished.
+Additionally, multiple clients can communicate with the same daemon simultaneously.
+A complex instrument may involve multiple operators watching sensor data in real time, while one program is orchestrating the hardware and recording the data.
 
+Script writers may simply wait for ``is\_busy'' to return ``false''.
+
+\begin{codefragment}{Python}
+import yaqc
+import time
+motor = yaqc.Client(port=38000)
+motor.set_position(500)  # initiate a long motor motion
+while motor.is_busy():
+    time.sleep(0.001)
+print("done")  # will only print when motion is complete
+\end{codefragment}
+
+This same pattern exists for sensors and more complex peripherals.
+
+\clearpage
 \subsection{Raster Script}
@@
 
-The Wright Group uses a purpose-build graphical program to perform data acquisition.
+The Wright Group uses a purpose-built graphical program to perform data acquisition.
 All of the hardware used is controlled via \yaq{}.

### 2023-03-16 SI.tex
non-blocking

@@
 
-Script writers may simply wait for ``is\_busy'' to return ``false''.
+Script writers may simply wait for ``busy'' to return ``false''.
 
@@
 motor.set_position(500)  # initiate a long motor motion
-while motor.is_busy():
+while motor.busy():
     time.sleep(0.001)
@@
 for m1 in range(-10, 10, 1):
+    motor1.set_position(m1)
     for m2 in range(0, 300, 5):
-        motor1.set_position(m1)
         motor2.set_position(m2)
@@
 Lightweight scripts such as these are encouraged by the \yaq{} project.
+
+\clearpage
+
+\subsection{Non-Blocking Script}
+
+\begin{codefragment}{Python}
+import time
+import yaqc
+temp = yaqc.Client(port=38003)
+sensor = yaqc.Client(port=38004)
+data = []
+# begin heating
+temp.set_position(100)
+# take data until temp greater than 90 deg
+while temp.get_position <= 90:
+    sensor.measure()
+    while sensor.busy():
+        time.sleep(0.001)
+    reading = dict()
+    reading["timestamp"] = time.time()
+    reading["temperature"] = temp.get_position()
+    reading.update(sensor.get_measured())
+    data.append(reading)
+\end{codefragment}
+
+\clearpage
 
@@
 The plot tab (Figure \ref{acq:fig:plot_tab}) presents a live representation of recent data collected for the current plan.
-The primary window is a PyQtGraph\cite{pyqtgraph} plot showing the most recent five slices of collected data.
+The primary window is a PyQtGraph \cite{pyqtgraph} plot showing the most recent five slices of collected data.
 The current slice is the brightest cyan slice, with the previous slices shown as duller colors progressively until the last one which is gray.
@@
 
-The last major change to \yaq{}'s protocol format was the adoption of Apache Avro-RPC\cite{AvroSpecification}.
+The last major change to \yaq{}'s protocol format was the adoption of Apache Avro-RPC \cite{AvroSpecification}.
 Avro is similarly terse like Msgpack for serializing messages, while having a well defined specification for defining an RPC schema: a static definition of available messages, including their signatures and documentation.
@@
 yaqd-control can be installed via
-PyPI\cite{yaqd-control} or
-conda-forge\cite{yaqd-control-conda}.
+PyPI \cite{yaqd-control} or
+conda-forge \cite{yaqd-control-conda}.
 
@@
 
-The raw data for package size analysis was collected using Tokei\cite{tokei}, a command line tool for analysing line lengths of source code.
+The raw data for package size analysis was collected using Tokei \cite{tokei}, a command line tool for analysing line lengths of source code.
 It includes breaking down line length into ``code'', ``comments'', and ``blank'' lines.
@@
 
+\section{The name \yaq{}}
+
+The name \yaq{} started out as an ``inside joke'' with several meanings. \\
+For posterity, we describe those meanings here.
+\begin{enumerate}
+  \item \yaq{} is similar to the popular acronym ``daq'' (Data AcQuisition)
+  \item \yaq{} could stand for Yet Another acQuisition, a reference to the popular tongue-in-cheek naming convention ``yet another X'' in software development \cite{Raymond_1996}.
+This is a particularly poignant reference for the authors, who are well aware that \yaq{} is ``yet another project'' in a long line of increasingly sophisticated hardware interface software packages in the Wright Group.
+  \item \yaq{} can reference the animal ``yak'', \textit{Bos grunniens}, who serves as a mascot for the project.
+  \item \yaq{} is an obscure reference to the term ``yak shaving'', coined by Carlin Vieri at MIT in the 1990s \cite{CarlinVieri}.
+Quoting from Wiktionary \cite{yak_shaving}: yak shaving is \textit{any apparently useless activity which, by allowing you to overcome intermediate difficulties, allows you to solve a larger problem}.
+The entire \yaq{} project is one giant ``yak shaving'' task that ultimately enables progress towards our primary mission---creative, reproducible, insightful experimental science.
+\end{enumerate}
+
 \clearpage

### 2023-03-16 paper.tex
citation cleanup (#57)
non-blocking

@@
 \author{John C. Wright}
-\author{Blaise J. Thompson}
+\author{Blaise J. Thompson*}
 \affiliation{
-  University of Wisconsin--Madison
+  University of Wisconsin--Madison;
+  Madison, Wisconsin, 53706;
+  USA
 }
@@
 While this growth is encouraging, many of these are limited by their focus on particular types of hardware or particular experimental domains.
+More work needs to be done to catalog, compare, and contrast the large number of open-source projects that now exist.
 Most small custom research instrumentation continues to rely on monolithic software which has hard-coded interface support for each particular connected device.
@@
 
-We have created a new network-based communication standard for scientific instrumentation, \yaq{}.
+We have created a new network-based communication standard for scientific instrumentation, \yaq{} (Yet Another acQuisition)\footnote{see Supplementary Material for a full description of the name ``\yaq{}''}.
 This standard borrows the most important ideas from established projects used by large user facilities while retaining the simplicity appropriate for small research labs.
@@
 Bluesky has no built-in hardware interface support, instead relying on packages to create hardware-interfaces that are compatible with their Hardware Protocol defintions.
-\texttt{yaqc-bluesky}\footnote{\texttt{yaqc-bluesky} available on PyPI at https://pypi.org/project/yaqc-bluesky/} is a specialized client which adapts any \yaq{} protocol into Bluesky's Hardware Protocol.
+\texttt{yaqc-bluesky}\footnote{\texttt{yaqc-bluesky} available on PyPI at https://pypi.org/project/yaqc-bluesky/} is a specialized client which adapts any \yaq{} protocol into Bluesky.
 Similar translation layers could be built for a variety of orchestration software such as PyMoDAQ\cite{WeberSebastien2021a}, PyMeasure\cite{pymeasure}, or TRSpectrometer\cite{trspectrometer}.
@@
 Full set of line-count data and script used to produce Figure \ref{fig:histogram}.
+More detailed examples of orchestration using \yaq{}, including scripts and graphical user interfaces.
+Detailed description of \texttt{yaqd-control}.
+Discussion and quantitative analysis of the \yaq{} interface's performance with large arrays.
 
@@
 The authors would like to thank all \yaq{} users and contributors.
-We would also like to acknowledge the developers of the broader open source software community upon-which this project rests.
+We would also like to acknowledge the developers of the broader open-source software community upon-which this project rests.
 
@@
 \section*{References}
-\nocite{*}
 \bibliography{references}

### 2023-03-16 SI.tex
wait for next second

@@
 # take data until temp greater than 90 deg
-while temp.get_position <= 90:
+while temp.get_position() <= 90:
     sensor.measure()
@@
     data.append(reading)
+    # wait for next second
+    while time.time() - reading["timestamp"] < 1:
+        time.sleep(0.001)
+# stop heating
+temp.set_position(25)
 \end{codefragment}

### 2023-03-16 SI.tex
measure

--- 2023-03-16 SI.tex
@@
 # take data until temp greater than 90 deg
-while temp.get_position <= 90:
+while temp.get_position() <= 90:
+    reading = dict()
+    reading["timestamp"] = time.time()
+    reading["temperature"] = temp.get_position()
+    # measure
     sensor.measure()
@@
         time.sleep(0.001)
-    reading = dict()
-    reading["timestamp"] = time.time()
-    reading["temperature"] = temp.get_position()
     reading.update(sensor.get_measured())
     data.append(reading)
+    # wait for next second
+    while time.time() - reading["timestamp"] < 1:
+        time.sleep(0.001)
+# stop heating
+temp.set_position(25)
 \end{codefragment}

### 2023-03-16 SI.tex
si-typos

@@
 
+We believe that Python and the Scientific Python ecosystem is a powerfully simple tool for defining experimental procedures.
+Lightweight scripts such as these are encouraged by the \yaq{} project.
+The following two sections contain example orchestration scripts for common tasks.
+
 \clearpage
@@
 
-We believe that Python and the Scientific Python ecosystem is a powerfully simple tool for defining experimental procedures.
-Lightweight scripts such as these are encouraged by the \yaq{} project.
-
 \clearpage
@@
 \subsection{Non-Blocking Script}
+
+Sometimes you want to acquire data while a peripheral continuously moves, asynchronously with data acquisition.
+The following is an example where sample temperature is continuously ramped and data is collected every second.
 
@@
 \end{codefragment}
+
+Imagine that the temperature starts out at room temperature, 25.
+We first set the temp client to 100, beginning to heat the sample.
+The outer while loop will continue until the temperature reaches our desired end-point of 90.
+Each second, the while loop collects the current temperature and the current sensor reading.
+This type of experiment is common in the chemical sciences.
+
+Note that we pay no attention to the busy state of the temp client.
+When compared with the previous raster scanning example, we see that \yaq{} can be flexible about which hardware operations are kept synchronous and which are allowed to run ``freely'' without blocking other operations.
+The ubiquity and simplicity of the busy message allows for this flexibility.
 
@@
 Figure \ref{si:fig:scaling} shows the relationship between number of pixels and time for a \yaq{} message to read the array.
-Below about $2^{18}$ (~250,000) pixels, there is virtually no dependence on size, with the standard overhead of a \yaq{} message dominating the time for transfer.
-\yaq{} remains usable for up to approximately $2^{20}$ (~1,000,000) pixels.
+Below about $2^{18}$ ($\approx250,000$) pixels, there is virtually no dependence on size, with the standard overhead of a \yaq{} message dominating the time for transfer.
+\yaq{} remains usable for up to approximately $2^{20}$ ($\approx1,000,000$) pixels.
 ``Usable'' is a relative term which will depend greatly on context.
@@
 Anecdotally, many users of Bluesky, EPICS, and related technologies use a similar strategy for large image data.
-This has not yet been implemented for \ya{}, and would require clients to have knowledge of how to retrieve and display the images collected, but there is nothing preventing this if throughput is a limitation.
+This has not yet been implemented for \yaq{}, and would require clients to have knowledge of how to retrieve and display the images collected, but there is nothing preventing this if throughput is a limitation.
 If such behavior became a standard feature desired for cameras, it should be encapsulated in a \yaq{} trait to provide a consistent interface.

### 2023-03-16 build.sh
diff

@@ -1,9 +1,18 @@
 #!/bin/bash
+
+# manuscript
 pdflatex --interaction=nonstopmode paper.tex
 bibtex paper.aux
 pdflatex --interaction=nonstopmode paper.tex
 pdflatex --interaction=nonstopmode paper.tex
-#
+
+# diff
+pdflatex --interaction=nonstopmode paper-diff.tex
+bibtex paper-diff.aux
+pdflatex --interaction=nonstopmode paper-diff.tex
+pdflatex --interaction=nonstopmode paper-diff.tex
+
+# SI
 pdflatex --interaction=nonstopmode -shell-escape SI.tex
 bibtex SI.aux
 pdflatex --interaction=nonstopmode -shell-escape SI.tex

### 2023-03-16 review/as-submitted-2/manuscript.tex
SI cite bluesky

--- 2023-03-16 paper.tex
@@
 While this growth is encouraging, many of these are limited by their focus on particular types of hardware or particular experimental domains.
-More work needs to be done to catalog, compare, and contrast the large number of open-source projects that now exist.
 Most small custom research instrumentation continues to rely on monolithic software which has hard-coded interface support for each particular connected device.
@@
 
-We have created a new network-based communication standard for scientific instrumentation, \yaq{} (Yet Another acQuisition)\footnote{see Supplementary Material for a full description of the name ``\yaq{}''}.
+We have created a new network-based communication standard for scientific instrumentation, \yaq{}.
 This standard borrows the most important ideas from established projects used by large user facilities while retaining the simplicity appropriate for small research labs.

### 2023-03-16 SI.tex
SI cite bluesky
more SI typos

@@
 
-\subsection{\texttt{is\_busy} for ensuring order of operations}
+\subsection{\texttt{busy} for ensuring order of operations}
 
@@
 All of the hardware used is controlled via \yaq{}.
-Data is collected using Bluesky, which runs in a background process.
+Data is collected using Bluesky \cite{AllanDanielB2019a}, which runs in a background process.
 
@@
 The Avro Specification \cite{AvroSpecification} specifically provides for compression codecs in the case of storing to disk, but there is nothing preventing the RPC pipe from similarly encoding the data, provided both ends of the communication support it.
-EPICS provides a similar NDArray data type, which has individual image level compression configureable \cite{normative_types}.
+EPICS provides a similar NDArray data type, which has individual image level compression configurable \cite{normative_types}.
 Compression would allow larger arrays to be transported using fewer bytes over TCP, which would likely improve performance for large arrays where TCP transport is the bottleneck.

### 2023-03-16 review/as-submitted-3/manuscript.tex
as submitted

--- 2023-03-16 paper.tex
@@
 \section*{References}
-\bibliography{references}
+%aipnum4-2.bst 2019-01-14 (MD) hand-edited version of apsrev4-1.bst
+%Control: key (0)
+%Control: author (8) initials jnrlst
+%Control: editor formatted (1) identically to author
+%Control: production of article title (0) allowed
+%Control: page (1) range
+%Control: year (1) truncated
+%Control: production of eprint (0) enabled
+\begin{thebibliography}{45}%
+\makeatletter
+\providecommand \@ifxundefined [1]{%
+ \@ifx{#1\undefined}
+}%
+\providecommand \@ifnum [1]{%
+ \ifnum #1\expandafter \@firstoftwo
+ \else \expandafter \@secondoftwo
+ \fi
+}%
+\providecommand \@ifx [1]{%
+ \ifx #1\expandafter \@firstoftwo
+ \else \expandafter \@secondoftwo
+ \fi
+}%
+\providecommand \natexlab [1]{#1}%
+\providecommand \enquote  [1]{``#1''}%
+\providecommand \bibnamefont  [1]{#1}%
+\providecommand \bibfnamefont [1]{#1}%
+\providecommand \citenamefont [1]{#1}%
+\providecommand \href@noop [0]{\@secondoftwo}%
+\providecommand \href [0]{\begingroup \@sanitize@url \@href}%
+\providecommand \@href[1]{\@@startlink{#1}\@@href}%
+\providecommand \@@href[1]{\endgroup#1\@@endlink}%
+\providecommand \@sanitize@url [0]{\catcode `\\12\catcode `\$12\catcode
+  `\&12\catcode `\#12\catcode `\^12\catcode `\_12\catcode `\%12\relax}%
+\providecommand \@@startlink[1]{}%
+\providecommand \@@endlink[0]{}%
+\providecommand \url  [0]{\begingroup\@sanitize@url \@url }%
+\providecommand \@url [1]{\endgroup\@href {#1}{\urlprefix }}%
+\providecommand \urlprefix  [0]{URL }%
+\providecommand \Eprint [0]{\href }%
+\providecommand \doibase [0]{https://doi.org/}%
+\providecommand \selectlanguage [0]{\@gobble}%
+\providecommand \bibinfo  [0]{\@secondoftwo}%
+\providecommand \bibfield  [0]{\@secondoftwo}%
+\providecommand \translation [1]{[#1]}%
+\providecommand \BibitemOpen [0]{}%
+\providecommand \bibitemStop [0]{}%
+\providecommand \bibitemNoStop [0]{.\EOS\space}%
+\providecommand \EOS [0]{\spacefactor3000\relax}%
+\providecommand \BibitemShut  [1]{\csname bibitem#1\endcsname}%
+\let\auto@bib@innerbib\@empty
+%</preamble>
+\bibitem [{nid(2022)}]{nidaqmx}%
+  \BibitemOpen
+  \href@noop {} {\enquote {\bibinfo {title} {Getting started with
+  {NI-DAQmx}},}\ }\bibinfo {howpublished}
+  {\url{https://www.ni.com/en-us/support/documentation/supplemental/06/getting-started-with-ni-daqmx--main-page.html}}
+  (\bibinfo {year} {2022}),\ \bibinfo {note} {accessed: 2022-10-02}\BibitemShut
+  {NoStop}%
+\bibitem [{scp(1999)}]{scpi}%
+  \BibitemOpen
+  \href@noop {} {\emph {\bibinfo {title} {Standard Commands for Programmable
+  Instruments (SCPI)}}},\ \bibinfo {organization} {SCPI Consortium} (\bibinfo
+  {year} {1999})\BibitemShut {NoStop}%
+\bibitem [{mod(2012)}]{modbus}%
+  \BibitemOpen
+  \href@noop {} {\emph {\bibinfo {title} {MODBUS Application Protocol
+  Specification}}},\ \bibinfo {organization} {Modbus} (\bibinfo {year}
+  {2012})\BibitemShut {NoStop}%
+\bibitem [{pic(2021)}]{picam}%
+  \BibitemOpen
+  \href@noop {} {\emph {\bibinfo {title} {PICam 5.x Programmer’s Manual}}},\
+  \bibinfo {organization} {Teledyne Princeton Instruments} (\bibinfo {year}
+  {2021})\BibitemShut {NoStop}%
+\bibitem [{tho(2022)}]{thorlabs_apt}%
+  \BibitemOpen
+  \href@noop {} {\emph {\bibinfo {title} {Thorlabs Motion Controllers
+  Host-Controller Communications Protocol}}},\ \bibinfo {organization}
+  {Thorlabs} (\bibinfo {year} {2022})\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Dalesio}, \citenamefont {Kraimer},\ and\
+  \citenamefont {Kozubal}(1991)}]{DalesioLR1991a}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {L.~R.}\ \bibnamefont
+  {Dalesio}}, \bibinfo {author} {\bibfnamefont {M.~R.}\ \bibnamefont
+  {Kraimer}},\ and\ \bibinfo {author} {\bibfnamefont {A.~J.}\ \bibnamefont
+  {Kozubal}},\ }\bibfield  {title} {\enquote {\bibinfo {title} {{EPICS}
+  architecture},}\ }\href@noop {} {\bibfield  {journal} {\bibinfo  {journal}
+  {ICALEPCS}\ }\textbf {\bibinfo {volume} {91}} (\bibinfo {year}
+  {1991})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Chaize}\ \emph {et~al.}(1999)\citenamefont {Chaize},
+  \citenamefont {Götz}, \citenamefont {Meyer}, \citenamefont {Pérez},\ and\
+  \citenamefont {Taurel}}]{AGotz1999TANGOA}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {J.-M.}\ \bibnamefont
+  {Chaize}}, \bibinfo {author} {\bibfnamefont {A.}~\bibnamefont {Götz}},
+  \bibinfo {author} {\bibfnamefont {J.}~\bibnamefont {Meyer}}, \bibinfo
+  {author} {\bibfnamefont {M.}~\bibnamefont {Pérez}},\ and\ \bibinfo {author}
+  {\bibfnamefont {E.}~\bibnamefont {Taurel}},\ }\bibfield  {title} {\enquote
+  {\bibinfo {title} {{TANGO} - an object oriented control system based on
+  {CORBA}},}\ }in\ \href@noop {} {\emph {\bibinfo {booktitle} {ICALEPCS}}}\
+  (\bibinfo {year} {1999})\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Weber}(2021)}]{WeberSebastien2021a}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {S.~J.}\ \bibnamefont
+  {Weber}},\ }\bibfield  {title} {\enquote {\bibinfo {title} {{PyMoDAQ}: An
+  open-source python-based software for modular data acquisition},}\ }\href
+  {https://doi.org/10.1063/5.0032116} {\bibfield  {journal} {\bibinfo
+  {journal} {Review of Scientific Instruments}\ }\textbf {\bibinfo {volume}
+  {92}},\ \bibinfo {pages} {045104} (\bibinfo {year} {2021})}\BibitemShut
+  {NoStop}%
+\bibitem [{\citenamefont {Bogdanowicz}\ \emph {et~al.}(2022)\citenamefont
+  {Bogdanowicz}, \citenamefont {Rogers}, \citenamefont {Weber}, \citenamefont
+  {{Zakv}}, \citenamefont {Pelissier}, \citenamefont {Wheeler}, \citenamefont
+  {{Cxz}}, \citenamefont {Marazzi}, \citenamefont {{Eedm}},\ and\ \citenamefont
+  {Galinskiy}}]{Bogdanowicz2022}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {N.}~\bibnamefont
+  {Bogdanowicz}}, \bibinfo {author} {\bibfnamefont {C.}~\bibnamefont {Rogers}},
+  \bibinfo {author} {\bibfnamefont {S.}~\bibnamefont {Weber}}, \bibinfo
+  {author} {\bibnamefont {{Zakv}}}, \bibinfo {author} {\bibfnamefont
+  {S.}~\bibnamefont {Pelissier}}, \bibinfo {author} {\bibfnamefont
+  {J.}~\bibnamefont {Wheeler}}, \bibinfo {author} {\bibnamefont {{Cxz}}},
+  \bibinfo {author} {\bibfnamefont {F.}~\bibnamefont {Marazzi}}, \bibinfo
+  {author} {\bibnamefont {{Eedm}}},\ and\ \bibinfo {author} {\bibfnamefont
+  {I.}~\bibnamefont {Galinskiy}},\ }\href
+  {https://doi.org/10.5281/ZENODO.2556398} {\enquote {\bibinfo {title}
+  {Instrumental: A python-based library for controlling lab hardware},}\ }
+  (\bibinfo {year} {2022})\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Tapping}(2021)}]{trspectrometer}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {P.}~\bibnamefont
+  {Tapping}},\ }\href@noop {} {\enquote {\bibinfo {title} {{TRSpectrometer}
+  documentation},}\ }\bibinfo {howpublished}
+  {\url{https://trspectrometer.readthedocs.io/}} (\bibinfo {year} {2021}),\
+  \bibinfo {note} {accessed: 2022-10-02}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Koerner}(2019)}]{Koerner_2019}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {L.}~\bibnamefont
+  {Koerner}},\ }\bibfield  {title} {\enquote {\bibinfo {title} {Instrbuilder: A
+  {Python} package for electrical instrument control},}\ }\href
+  {https://doi.org/10.21105/joss.01172} {\bibfield  {journal} {\bibinfo
+  {journal} {Journal of Open Source Software}\ }\textbf {\bibinfo {volume}
+  {4}},\ \bibinfo {pages} {1172} (\bibinfo {year} {2019})}\BibitemShut
+  {NoStop}%
+\bibitem [{\citenamefont {Campagnola}, \citenamefont {Kratz},\ and\
+  \citenamefont {Manis}(2014)}]{Campagnola_2014}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {L.}~\bibnamefont
+  {Campagnola}}, \bibinfo {author} {\bibfnamefont {M.~B.}\ \bibnamefont
+  {Kratz}},\ and\ \bibinfo {author} {\bibfnamefont {P.~B.}\ \bibnamefont
+  {Manis}},\ }\bibfield  {title} {\enquote {\bibinfo {title} {Acq4: an
+  open-source software platform for data acquisition and analysis in
+  neurophysiology research},}\ }\href
+  {https://doi.org/10.3389/fninf.2014.00003} {\bibfield  {journal} {\bibinfo
+  {journal} {Frontiers in Neuroinformatics}\ }\textbf {\bibinfo {volume} {8}}
+  (\bibinfo {year} {2014}),\ 10.3389/fninf.2014.00003}\BibitemShut {NoStop}%
+\bibitem [{pym(2022)}]{pymeasure}%
+  \BibitemOpen
+  \href@noop {} {\enquote {\bibinfo {title} {{PyMeasure}},}\ }\bibinfo
+  {howpublished} {\url{https://pymeasure.readthedocs.io/}} (\bibinfo {year}
+  {2022}),\ \bibinfo {note} {accessed: 2022-10-02}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Giesbrecht}\ \emph {et~al.}(2022)\citenamefont
+  {Giesbrecht}, \citenamefont {Amsellem}, \citenamefont {Bauer}, \citenamefont
+  {Mak}, \citenamefont {Wynne}, \citenamefont {Qin},\ and\ \citenamefont
+  {Persaud}}]{Giesbrecht_2022}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {G.}~\bibnamefont
+  {Giesbrecht}}, \bibinfo {author} {\bibfnamefont {A.}~\bibnamefont
+  {Amsellem}}, \bibinfo {author} {\bibfnamefont {T.}~\bibnamefont {Bauer}},
+  \bibinfo {author} {\bibfnamefont {B.}~\bibnamefont {Mak}}, \bibinfo {author}
+  {\bibfnamefont {B.}~\bibnamefont {Wynne}}, \bibinfo {author} {\bibfnamefont
+  {Z.}~\bibnamefont {Qin}},\ and\ \bibinfo {author} {\bibfnamefont
+  {A.}~\bibnamefont {Persaud}},\ }\bibfield  {title} {\enquote {\bibinfo
+  {title} {Hardware-control: Instrument control and automation package},}\
+  }\href {https://doi.org/10.21105/joss.02688} {\bibfield  {journal} {\bibinfo
+  {journal} {Journal of Open Source Software}\ }\textbf {\bibinfo {volume}
+  {7}},\ \bibinfo {pages} {2688} (\bibinfo {year} {2022})}\BibitemShut
+  {NoStop}%
+\bibitem [{\citenamefont {Shkarin}(2022)}]{pylablib}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {A.}~\bibnamefont
+  {Shkarin}},\ }\href {https://doi.org/10.5281/zenodo.7324876} {\enquote
+  {\bibinfo {title} {{pyLabLib}},}\ }\bibinfo {howpublished}
+  {\url{https://doi.org/10.5281/zenodo.7324876}} (\bibinfo {year}
+  {2022})\BibitemShut {NoStop}%
+\bibitem [{Note1()}]{Note1}%
+  \BibitemOpen
+  \bibinfo {note} {See Supplementary Material for a full description of the
+  name ``\protect \texttt {yaq}{}''}\BibitemShut {NoStop}%
+\bibitem [{Note2()}]{Note2}%
+  \BibitemOpen
+  \bibinfo {note} {The \protect \texttt {yaq}{} protocol is fully specified by
+  the set of accepted \protect \texttt {yaq}{} Enhancement Proposals at
+  https://yeps.yaq.fyi/}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Raymond}(1996)}]{Raymond_1996}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {E.~S.}\ \bibnamefont
+  {Raymond}},\ }\href@noop {} {\emph {\bibinfo {title} {The new hacker’s
+  dictionary}}},\ \bibinfo {edition} {3rd}\ ed.\ (\bibinfo  {publisher} {MIT
+  Press},\ \bibinfo {address} {Cambridge, Mass},\ \bibinfo {year}
+  {1996})\BibitemShut {NoStop}%
+\bibitem [{Avr(2022)}]{AvroSpecification}%
+  \BibitemOpen
+  \href@noop {} {\enquote {\bibinfo {title} {Apache avro specification},}\
+  }\bibinfo {howpublished}
+  {\url{https://avro.apache.org/docs/1.11.1/specification/}} (\bibinfo {year}
+  {2022}),\ \bibinfo {note} {accessed: 2022-10-02}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Segal}(2005)}]{SegalJudith2005a}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {J.}~\bibnamefont
+  {Segal}},\ }\bibfield  {title} {\enquote {\bibinfo {title} {When software
+  engineers met research scientists: A case study},}\ }\href
+  {https://doi.org/10.1007/s10664-005-3865-y} {\bibfield  {journal} {\bibinfo
+  {journal} {Empirical Software Engineering}\ }\textbf {\bibinfo {volume}
+  {10}},\ \bibinfo {pages} {517--536} (\bibinfo {year} {2005})}\BibitemShut
+  {NoStop}%
+\bibitem [{Note3()}]{Note3}%
+  \BibitemOpen
+  \bibinfo {note} {\protect \texttt {yaqc} available on PyPI at
+  https://pypi.org/project/yaqc/}\BibitemShut {NoStop}%
+\bibitem [{Note4()}]{Note4}%
+  \BibitemOpen
+  \bibinfo {note} {\protect \texttt {yaqc-qtpy} available on PyPI at
+  https://pypi.org/project/yaqc-qtpy/}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Allan}\ \emph {et~al.}(2019)\citenamefont {Allan},
+  \citenamefont {Caswell}, \citenamefont {Campbell},\ and\ \citenamefont
+  {Rakitin}}]{AllanDanielB2019a}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {D.}~\bibnamefont
+  {Allan}}, \bibinfo {author} {\bibfnamefont {T.}~\bibnamefont {Caswell}},
+  \bibinfo {author} {\bibfnamefont {S.}~\bibnamefont {Campbell}},\ and\
+  \bibinfo {author} {\bibfnamefont {M.}~\bibnamefont {Rakitin}},\ }\bibfield
+  {title} {\enquote {\bibinfo {title} {Bluesky's ahead: A multi-facility
+  collaboration for an a la carte software project for data acquisition and
+  management},}\ }\href {https://doi.org/10.1080/08940886.2019.1608121}
+  {\bibfield  {journal} {\bibinfo  {journal} {Synchrotron Radiation News}\
+  }\textbf {\bibinfo {volume} {32}},\ \bibinfo {pages} {19--22} (\bibinfo
+  {year} {2019})}\BibitemShut {NoStop}%
+\bibitem [{Note5()}]{Note5}%
+  \BibitemOpen
+  \bibinfo {note} {\protect \texttt {yaqc-bluesky} available on PyPI at
+  https://pypi.org/project/yaqc-bluesky/}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Mukamel}(2000)}]{MukamelShaul2000a}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {S.}~\bibnamefont
+  {Mukamel}},\ }\bibfield  {title} {\enquote {\bibinfo {title}
+  {Multidimensional femtosecond correlation spectroscopies of electronic and
+  vibrational excitations},}\ }\href
+  {https://doi.org/10.1146/annurev.physchem.51.1.691} {\bibfield  {journal}
+  {\bibinfo  {journal} {Annual Review of Physical Chemistry}\ }\textbf
+  {\bibinfo {volume} {51}},\ \bibinfo {pages} {691--729} (\bibinfo {year}
+  {2000})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Wright}(2011)}]{WrightJohnCurtis2011a}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {J.~C.}\ \bibnamefont
+  {Wright}},\ }\bibfield  {title} {\enquote {\bibinfo {title} {Multiresonant
+  coherent multidimensional spectroscopy},}\ }\href
+  {https://doi.org/10.1146/annurev-physchem-032210-103551} {\bibfield
+  {journal} {\bibinfo  {journal} {Annual Review of Physical Chemistry}\
+  }\textbf {\bibinfo {volume} {62}},\ \bibinfo {pages} {209--230} (\bibinfo
+  {year} {2011})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Carlson}(1988)}]{CarlsonRogerJohn1988a}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {R.~J.}\ \bibnamefont
+  {Carlson}},\ }\emph {\bibinfo {title} {Quantitative Aspects of High
+  Resolution, Fully Resonant, Four-Wave Mixing Spectroscopy For the Analysis of
+  Vibronic Mode Coupling in Molecules}},\ \href@noop {} {Ph.D. thesis},\
+  \bibinfo  {school} {University of Wisconsin-Madison} (\bibinfo {year}
+  {1988})\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Meyer}(2004)}]{MeyerKentAlbert2004b}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {K.~A.}\ \bibnamefont
+  {Meyer}},\ }\emph {\bibinfo {title} {Frequency-Scanned Ultrafast
+  Spectroscopic Techniques Applied to Infrared Four-Wave Mixing
+  Spectroscopy}},\ \href@noop {} {Ph.D. thesis},\ \bibinfo  {school}
+  {University of Wisconsin--Madison} (\bibinfo {year} {2004})\BibitemShut
+  {NoStop}%
+\bibitem [{\citenamefont {Kain}(2017)}]{KainSchuyler2017a}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {S.}~\bibnamefont
+  {Kain}},\ }\emph {\bibinfo {title} {Transition of Frequency-Domain Coherent
+  Multidimensional Spectroscopic Methods to the Femtosecond Time Regime with
+  Applications to Nanoscale Semiconductors}},\ \href@noop {} {Ph.D. thesis},\
+  \bibinfo  {school} {University of Wisconsin-Madison} (\bibinfo {year}
+  {2017})\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Thompson}(2018)}]{ThompsonBlaiseJonathan2018a}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {B.~J.}\ \bibnamefont
+  {Thompson}},\ }\emph {\bibinfo {title} {Development of Frequency Domain
+  Multidimensional Spectroscopy with Applications in Semiconductor
+  Photophysics}},\ \href {https://doi.org/10.5281/zenodo.7627321} {Ph.D.
+  thesis},\ \bibinfo  {school} {University of Wisconsin--Madison} (\bibinfo
+  {year} {2018})\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Sunden}(2022)}]{SundenKyle2022}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {K.~F.}\ \bibnamefont
+  {Sunden}},\ }\emph {\bibinfo {title} {yaq: Yet Another Acquisition A modular
+  approach to spectroscopy software and instrumentation}},\ \href
+  {https://doi.org/10.5281/zenodo.7627331} {Ph.D. thesis},\ \bibinfo  {school}
+  {University of Wisconsin--Madison} (\bibinfo {year} {2022})\BibitemShut
+  {NoStop}%
+\bibitem [{Note6()}]{Note6}%
+  \BibitemOpen
+  \bibinfo {note} {\protect \texttt {yaqd-core-python} available on PyPI at
+  https://pypi.org/project/yaqd-core}\BibitemShut {NoStop}%
+\bibitem [{Note7()}]{Note7}%
+  \BibitemOpen
+  \bibinfo {note} {\protect \texttt {yaq-traits} available on PyPI at
+  https://pypi.org/project/yaq-traits}\BibitemShut {NoStop}%
+\bibitem [{Note8()}]{Note8}%
+  \BibitemOpen
+  \bibinfo {note} {\protect \texttt {yaqd-brooks-mfc-gf} source code available
+  on GitHub at \protect \texttt
+  {https://github.com/yaq-project/yaqd-brooks/}}\BibitemShut {NoStop}%
+\bibitem [{pys()}]{pyserial}%
+  \BibitemOpen
+  \href@noop {} {\enquote {\bibinfo {title} {{PySerial}},}\ }\bibinfo
+  {howpublished} {\url{https://pypi.org/project/pyserial}},\ \bibinfo {note}
+  {accessed: 2022-10-02}\BibitemShut {NoStop}%
+\bibitem [{pyu()}]{pyusb}%
+  \BibitemOpen
+  \href@noop {} {\enquote {\bibinfo {title} {{PyUSB}},}\ }\bibinfo
+  {howpublished} {\url{https://pypi.org/project/pyusb}},\ \bibinfo {note}
+  {accessed: 2022-11-06}\BibitemShut {NoStop}%
+\bibitem [{pyv()}]{pyvisa}%
+  \BibitemOpen
+  \href@noop {} {\enquote {\bibinfo {title} {{PyVISA}},}\ }\bibinfo
+  {howpublished} {\url{https://pypi.org/project/pyvisa}},\ \bibinfo {note}
+  {accessed: 2022-11-06}\BibitemShut {NoStop}%
+\bibitem [{pym()}]{pymodbus}%
+  \BibitemOpen
+  \href@noop {} {\enquote {\bibinfo {title} {{PyModbus}},}\ }\bibinfo
+  {howpublished} {\url{https://pypi.org/project/pymodbus}},\ \bibinfo {note}
+  {accessed: 2022-11-06}\BibitemShut {NoStop}%
+\bibitem [{Note9()}]{Note9}%
+  \BibitemOpen
+  \bibinfo {note} {See https://yaq.fyi/daemons/ and https://yaq.fyi/hardware/
+  for living lists of all \protect \texttt {yaq}{} daemons and supported
+  hardware, respectively.}\BibitemShut {Stop}%
+\bibitem [{\citenamefont {Salazar}\ \emph {et~al.}(2021)\citenamefont
+  {Salazar}, \citenamefont {Thompson}, \citenamefont {Knapp}, \citenamefont
+  {Myers},\ and\ \citenamefont {Stahl}}]{SalazarChaseA2021a}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {C.~A.}\ \bibnamefont
+  {Salazar}}, \bibinfo {author} {\bibfnamefont {B.~J.}\ \bibnamefont
+  {Thompson}}, \bibinfo {author} {\bibfnamefont {S.~M.~M.}\ \bibnamefont
+  {Knapp}}, \bibinfo {author} {\bibfnamefont {S.~R.}\ \bibnamefont {Myers}},\
+  and\ \bibinfo {author} {\bibfnamefont {S.~S.}\ \bibnamefont {Stahl}},\
+  }\bibfield  {title} {\enquote {\bibinfo {title} {Multichannel
+  gas-uptake/evolution reactor for monitoring liquid-phase chemical
+  reactions},}\ }\href {https://doi.org/10.1063/5.0043007} {\bibfield
+  {journal} {\bibinfo  {journal} {Review of Scientific Instruments}\ }\textbf
+  {\bibinfo {volume} {92}},\ \bibinfo {pages} {044103} (\bibinfo {year}
+  {2021})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Allman}(2012)}]{Allman_2012}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {E.}~\bibnamefont
+  {Allman}},\ }\bibfield  {title} {\enquote {\bibinfo {title} {Managing
+  technical debt},}\ }\href {https://doi.org/10.1145/2160718.2160733}
+  {\bibfield  {journal} {\bibinfo  {journal} {Communications of the ACM}\
+  }\textbf {\bibinfo {volume} {55}},\ \bibinfo {pages} {50--55} (\bibinfo
+  {year} {2012})}\BibitemShut {NoStop}%
+\bibitem [{Note10()}]{Note10}%
+  \BibitemOpen
+  \bibinfo {note} {\protect \texttt {yaqd-fakes} available on PyPI at
+  https://pypi.org/project/yaqd-fakes/}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Segal}(2007)}]{SegalJudith2007a}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {J.}~\bibnamefont
+  {Segal}},\ }\bibfield  {title} {\enquote {\bibinfo {title} {Some problems of
+  professional end user developers},}\ }in\ \href
+  {https://doi.org/10.1109/vlhcc.2007.17} {\emph {\bibinfo {booktitle} {{IEEE}
+  Symposium on Visual Languages and Human-Centric Computing ({VL}/{HCC}
+  2007)}}}\ (\bibinfo  {publisher} {{IEEE}},\ \bibinfo {year}
+  {2007})\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Cohen}\ \emph {et~al.}(2021)\citenamefont {Cohen},
+  \citenamefont {Katz}, \citenamefont {Barker}, \citenamefont {Hong},
+  \citenamefont {Haines},\ and\ \citenamefont {Jay}}]{CohenJeremy2021a}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {J.}~\bibnamefont
+  {Cohen}}, \bibinfo {author} {\bibfnamefont {D.~S.}\ \bibnamefont {Katz}},
+  \bibinfo {author} {\bibfnamefont {M.}~\bibnamefont {Barker}}, \bibinfo
+  {author} {\bibfnamefont {N.~C.}\ \bibnamefont {Hong}}, \bibinfo {author}
+  {\bibfnamefont {R.}~\bibnamefont {Haines}},\ and\ \bibinfo {author}
+  {\bibfnamefont {C.}~\bibnamefont {Jay}},\ }\bibfield  {title} {\enquote
+  {\bibinfo {title} {The four pillars of research software engineering},}\
+  }\href {https://doi.org/10.1109/ms.2020.2973362} {\bibfield  {journal}
+  {\bibinfo  {journal} {IEEE Software}\ }\textbf {\bibinfo {volume} {38}},\
+  \bibinfo {pages} {97--105} (\bibinfo {year} {2021})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Nowogrodzki}(2019)}]{NowogrodzkiAnna2019a}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {A.}~\bibnamefont
+  {Nowogrodzki}},\ }\bibfield  {title} {\enquote {\bibinfo {title} {How to
+  support open-source software and stay sane},}\ }\href
+  {https://doi.org/10.1038/d41586-019-02046-0} {\bibfield  {journal} {\bibinfo
+  {journal} {Nature}\ }\textbf {\bibinfo {volume} {571}},\ \bibinfo {pages}
+  {133--134} (\bibinfo {year} {2019})}\BibitemShut {NoStop}%
+\end{thebibliography}%
 

### 2023-03-16 review/as-submitted-3/SI/figures/lines_histogram.py

@@ -1,0 +1,23 @@
+import numpy as np
+import pandas as pd
+import pathlib
+import matplotlib.pyplot as plt
+
+__here__ = pathlib.Path(__file__).parent
+
+data = pd.read_table(__here__ / "lines_histogram.txt")
+
+bins = np.arange(0, 351, 10)
+
+data["code"].hist(bins=bins)
+
+fig = plt.gcf()
+fig.set_size_inches(3.5, 2.5)
+ax = plt.gca()
+ax.set_xlabel("Lines of Code")
+ax.set_ylabel("Number of Daemons")
+ax.set_yticks(np.arange(9))
+ax.set_xticks(np.arange(0, 352, 50))
+ax.set_xlim(0, 350)
+
+plt.savefig("lines_histogram.png", dpi=300, bbox_inches="tight")

### 2023-03-16 review/as-submitted-3/SI/figures/scaling/plot-scaling.py

@@ -1,0 +1,32 @@
+import pathlib
+
+import pandas as pd
+import matplotlib.pyplot as plt
+
+__here__ = pathlib.Path(__file__).parent
+
+data = pd.read_csv(__here__ / "scaling.csv")
+
+fig, ax = plt.subplots()
+
+ax.errorbar(data["npts"], data["mean"], yerr=data["stdev"], marker="o", lw=0, elinewidth=2, capsize=3, capthick=2, zorder=3)
+ax.grid()
+ax.set_xscale("log", base=2)
+ax.set_yscale("log")
+
+ax.set_xlabel("Number of Pixels")
+ax.set_ylabel("Time (seconds)")
+
+for pt in data.itertuples():
+    if pt.shapex != 1:
+        ax.errorbar([pt.npts], [pt.mean], yerr=[pt.stdev], marker="o", lw=0, elinewidth=2, capsize=3, capthick=2, zorder=3.1)
+        ax.annotate(
+            str((pt.shapex, pt.shapey)),
+            (pt.npts, pt.mean),
+            xytext=(-80, 10),
+            textcoords="offset points",
+            arrowprops=dict(arrowstyle="->"),
+        )
+
+#plt.show()
+fig.savefig("scaling.png", dpi=300)

### 2023-03-16 review/as-submitted-3/SI/figures/scaling/scaling-data-gen.py

@@ -1,0 +1,49 @@
+import yaqc
+import numpy as np
+import time
+
+shapes = [
+    (1, 2**0),
+    (1, 2**1),
+    (1, 2**2),
+    (1, 2**3),
+    (1, 2**4),
+    (1, 2**5),
+    (1, 2**6),
+    (1, 2**7),
+    (1, 2**8),
+    (1, 2**9),
+    (1, 2**10),
+    (1, 2**11),
+    (1, 2**12),
+    (1, 2**13),
+    (1, 2**14),
+    (1, 2**15),
+    (1, 2**16),
+    (1, 2**17),
+    (1, 2**18),
+    (1, 2**19),
+    (1, 2**20),
+    (1, 2**21),
+    (1, 2**22),
+    (1, 2**23),
+    (1, 6000000),
+    (512, 512),
+    (1024, 1024),
+    (1024, 4096),
+]
+
+c = yaqc.Client(36656)
+
+print(f"shapex,shapey,npts,n,mean,stdev")
+for shape in shapes:
+    start_time = time.time()
+    times = []
+    for _ in range(10):
+        run_time = time.time()
+        a = c.rand_int32(*shape)
+        end_time = time.time()
+        times.append(end_time - run_time)
+        if end_time - start_time > 60:
+            break
+    print(f"{','.join(str(i) for i in shape)}, {np.prod(shape)}, {len(times)}, {np.mean(times)}, {np.std(times)}")
